\documentclass{article}%
\usepackage{Rd}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{geometry}
\usepackage{caption}
\usepackage[debugshow,pdftex]{graphicx}%
\usepackage{amsmath}%
\setcounter{MaxMatrixCols}{30}%
\usepackage{amsfonts}
%TCIDATA{OutputFilter=latex2.dll}
%TCIDATA{Version=5.50.0.2953}
%TCIDATA{CSTFile=40 LaTeX article.cst}
%TCIDATA{Created=Wednesday, August 01, 2007 14:12:27}
%TCIDATA{LastRevised=Sunday, December 07, 2008 07:20:01}
%TCIDATA{<META NAME="GraphicsSave" CONTENT="32">}
%TCIDATA{<META NAME="SaveForMode" CONTENT="1">}
%TCIDATA{BibliographyScheme=BibTeX}
%TCIDATA{<META NAME="DocumentShell" CONTENT="Articles\SW\Standard LaTeX Article (hyperref)">}
%TCIDATA{Language=American English}
%BeginMSIPreambleData
\providecommand{\U}[1]{\protect\rule{.1in}{.1in}}
%EndMSIPreambleData
\newtheorem{theorem}{Theorem}
\newtheorem{acknowledgement}[theorem]{Acknowledgement}
\newtheorem{algorithm}[theorem]{Algorithm}
\newtheorem{axiom}[theorem]{Axiom}
\newtheorem{case}[theorem]{Case}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{conclusion}[theorem]{Conclusion}
\newtheorem{condition}[theorem]{Condition}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{criterion}[theorem]{Criterion}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{exercise}[theorem]{Exercise}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{notation}[theorem]{Notation}
\newtheorem{problem}[theorem]{Problem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{solution}[theorem]{Solution}
\newtheorem{summary}[theorem]{Summary}
\newenvironment{proof}[1][Proof]{\noindent\textbf{#1.} }{\ \rule{0.5em}{0.5em}}
\geometry{left=108bp,right=72bp,top=90bp,bottom=90bp}
\begin{document}

\title{gsDesign: An R Package for Designing \\Group Sequential Clinical Trials\\Version 2.0 Manual}
\author{Keaven M. Anderson, Dan (Jennifer) Sun, Zhongxin (John) Zhang\\Merck Research Laboratories}
\maketitle

\begin{abstract}
The gsDesign package supports group sequential clinical trial design. While
there is a strong focus on designs using $\alpha-$ and $\beta-$spending
functions, Wang-Tsiatis designs, including O'Brien-Fleming and Pocock designs,
are also available. The ability to design with non-binding futility rules is
an important feature to control Type I error in a manner acceptable to
regulatory authorities when futility bounds are employed.

The routines are designed to provide simple access to commonly used designs
using default arguments. Standard, published spending functions are supported
as well as the ability to write custom spending functions. A \texttt{gsDesign}
class is defined and returned by the \texttt{gsDesign()} function. A plot
function for this class provides a wide variety of plots: boundaries, power,
estimated treatment effect at boundaries, conditional power at boundaries,
spending function plots, expected sample size plot, and B-values at
boundaries. Using function calls to access the package routines provides a
powerful capability to derive designs or output formatting that could not be
anticipated through a gui interface. This enables the user to easily create
designs with features they desire, such as designs with minimum expected
sample size.

Thus, the intent of the gsDesign package is to easily create, fully
characterize, and even optimize routine group sequential trial designs as well
as provide a tool to evaluate innovative designs.

\end{abstract}
\tableofcontents
\input{overview.tex}
\input{basicfeatures.tex}
\input{install.tex}
\input{detailedex.tex}
\input{syntax.tex}
\input{objtypes.tex}
\input{output.tex}
\input{spendfn.tex}
\input{statmethods.tex}

\input{gsDesign_package_doc.tex}
\input{gsDesign_main_doc.tex}
\input{bin_trial_doc.tex}
\input{spending_functions_doc.tex}
\input{other_doc.tex}

\bibliographystyle{plain}
\bibliography{acompat,design}


\section{Contacts}

\bigskip

Keaven M. Anderson: keaven\_anderson@merck.com

\bigskip
\end{document}
