---
title: "Calibrating futility spending to conditional power"
output:
  rmarkdown::html_vignette:
    css: vignette.css
vignette: >
  %\VignetteIndexEntry{Calibrating futility spending to conditional power}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                     fig.height = 4.8, dpi = 150)
options(width = 80)
```

```{r libraries, message=FALSE}
library(gsDesign)
library(lt)
library(ggplot2)
```

`gsCPFutilitySpending()` selects lower spending parameters so that conditional
power (CP) at selected interim futility bounds matches requested targets.
It provides a way to specify futility in terms of the probability of future
efficacy rejection. Each candidate is rebuilt with `gsDesign()`, including
the information needed to retain the design's target unconditional power.
Timing and efficacy spending remain fixed; maximum information need not
remain fixed.

The boundary summaries show three perspectives on future efficacy rejection:
**CP** assumes the observed effect at the bound continues, **CP H1** assumes
the planned alternative effect, and **PP** averages conditional power over
the posterior effect distribution using `gsPP()`. By default,
`gsBoundSummary()` uses the broad, mildly informative normal prior
`normalGrid(mu = x$delta / 2, sigma = 10 / sqrt(x$n.fix))` on the standardized
effect scale. Its center is halfway between the null (zero) and the planned
alternative (`x$delta`); it is updated using the interim result at the bound.
Thus PP incorporates uncertainty about the effect rather than assuming one
fixed value. CP H1 and PP provide context in the tables; the examples here
primarily calibrate **CP**. A parallel survival example uses
`gsPPFutilitySpending()` to target **PP** instead.

## A design warning: futility, sample size, and overall power

Targeting high conditional power at a futility boundary can require a much
larger maximum sample size than a fixed design with no interims if overall
power must remain unchanged. A stringent futility rule stops some trials
even when the planned treatment effect is real. Recovering that lost power
among trials that continue can be expensive in sample size. An alternative
is to accept lower overall power by increasing the design's beta and
recalibrating, rather than insisting on 90% power at any sample-size cost.

An aggressive futility analysis can be viewed as a Phase II proof-of-concept
gate embedded in a Phase III confirmatory trial. For perspective, two
independent studies with 80% and 90% power have a probability of success in
both of `0.80 * 0.90 = 0.72`. This is an illustration, not the power formula
for an integrated group sequential trial: its interim and final statistics
are dependent, and its overall power must be calculated from the joint
design. The same multiplication applies if the second probability is
explicitly conditional on passing the first gate.

For some development programs, 85% overall power may therefore be a more
realistic compromise than 90% with stringent futility. This is a design
choice to justify in context, not a universal recommendation. Compare the
maximum sample size, expected sample size, and probability of stopping a
truly effective treatment. The examples below retain 90% power to make the
sample-size cost visible; the binomial and normal examples compare a fixed
design, efficacy-only monitoring, and monitoring with calibrated futility.

A useful planning constraint is to limit the maximum sample size to, say,
120% of the fixed-design size at the same alpha, beta, and effect assumptions.
Rather than selecting interim CP targets in isolation, seek targets compatible
with that inflation budget. This may require lower CP thresholds (less
aggressive futility), different interim timing, or reconsideration of overall
power. With several interims, assess the targets jointly: the constraint
applies to the complete design, not separately to each look. The 20% limit is
an illustrative planning choice, not a default constraint enforced by
`gsCPFutilitySpending()`. The sample-size comparisons below show whether the
chosen targets meet that illustrative budget; lowering an early CP target
does not by itself guarantee inflation below 20%.

As cautious starting points for these examples, avoid CP targets above 0.2
before 50% information and consider up to 0.3 at or after 50%. These are
planning heuristics, not universal cutoffs or assurances of acceptable
futility. Inspect the effect estimates at the bounds, beta spending, and
sample-size inflation before deciding whether even these targets are too
stringent.

## What is being calibrated?

With `theta = NULL`, the default used below, CP assumes the observed effect
implied by each candidate bound, `lower$bound[i] / sqrt(n.I[i])`. Thus we target
the **CP** row in `gsBoundSummary()`, not **CP H1** or **PP**. A specified future
effect can instead be supplied with `theta`; a scalar is used at all targeted
looks. This is a standardized effect on the scale used by `gsCP()`.

The function supports beta-spending futility for `test.type` 3, 4, 7, and 8.
The examples use type 4, fixed information fractions, and Lan-DeMets
O'Brien-Fleming efficacy spending (`sfu = sfLDOF`), without `sfupar`.

| Number of CP targets | Supported spending families |
| --- | --- |
| One | `sfHSD`, `sfPower`, `sfExponential`, `sfLDOF` |
| Two | `sfLogistic`, `sfBetaDist`, `sfCauchy`, `sfNormal`, `sfExtremeValue`, `sfExtremeValue2` |
| One or more | `sfLinear`, with one free cumulative spending proportion per target |

The number of targets must match the number of free spending parameters.
Targeted looks must be active interim futility analyses; the final analysis
is not a calibration target. Multiple-target fitting uses latest-to-earliest
initialization followed by joint refinement. A fit is returned only when
every residual meets `control$cp_tol` (default `1e-4`).

We independently recompute CP from each returned or reconstructed design,
rather than relying only on the calibration metadata:

```{r verification-helper}
check_cp <- function(design, target, i = seq_along(target), tol = 1e-4) {
  achieved <- vapply(i, function(j) {
    theta <- design$lower$bound[j] / sqrt(design$n.I[j])
    sum(gsCP(design, i = j, zi = design$lower$bound[j], theta = theta)$upper$prob)
  }, numeric(1))
  stopifnot(max(abs(achieved - target)) <= tol)
  stopifnot(abs(sum(design$upper$prob[, 2]) - (1 - design$beta)) < 2e-5)
  out <- data.frame(
    Analysis = i, Target = target, CP = achieved,
    Residual = achieved - target
  )
  size_label <- if (inherits(design, "gsSurv")) "Events" else if (design$n.fix > 1) {
    "Participants"
  } else "Information"
  out[[size_label]] <- round(design$n.I[i], 1)
  out
}
```

The comparison tables label cumulative event counts for survival designs
and total participants across both groups for binomial and normal designs.
These are planning quantities shown to one decimal, without integer
conversion or recalibration. Boundary summaries also display the size at
each analysis (rounded up for presentation).

## One parameter: CP 0.3 at the first interim

Here futility is tested only at IA 1. The second interim retains efficacy
testing but has no futility test.

```{r one-parameter}
x1 <- gsDesign(
  k = 3, test.type = 4, timing = c(.5, .75),
  sfu = sfLDOF, sfl = sfHSD, sflpar = 1,
  testLower = c(TRUE, FALSE, FALSE)
)
fit1 <- gsCPFutilitySpending(x1, target_cp = .3, i = 1)
fit1$cpFutilitySpending$sflpar
check_cp(fit1, .3) |> lt()
```

### Event-driven survival design with `gsSurv()`

The same information fractions and spending specification can be used in an
event-driven survival design. Reuse the fitted parameter from `fit1`; other
survival assumptions use `gsSurv()` defaults, with equal randomization made
explicit. Check CP on the survival design itself, since numerical tolerances
can introduce small differences.

```{r survival-design}
surv_design <- gsSurv(
  k = 3, test.type = 4, timing = c(.5, .75),
  sfu = sfLDOF, sfl = sfHSD,
  sflpar = fit1$cpFutilitySpending$sflpar,
  testLower = c(TRUE, FALSE, FALSE), ratio = 1
)
check_cp(surv_design, .3) |> lt()
gsBoundSummary(surv_design, digits = 4, exclude = "B-value") |>
  lt() |> lt_format(columns = c("Efficacy", "Futility"), decimals = 4)
```

`gsCPFutilitySpending()` does not currently accept a `gsSurv` object directly.
Here calibration is performed on `gsDesign()` and its fitted spending is
then passed to `gsSurv()`. The survival summary reports approximate hazard
ratios at the bounds, rather than a generic standardized effect.

This first survival example is deliberately a cautionary one. CP 0.3 after
50% of the planned events gives a futility HR of about
`r round(gsHR(surv_design$lower$bound[1], 1, surv_design), 2)` and spends
`r round(100 * surv_design$lower$spend[1], 1)` percentage points of the total
10% Type II error budget at that first analysis. Thus it may stop a trial
showing approximately a 20% hazard reduction, and uses roughly three quarters
of the beta budget there. For a trial planned with 90% overall power, this
looks overly aggressive and motivates considering a lower CP target even
at 50% information.

### Targeting predictive power instead

`gsPPFutilitySpending()` can target PP using the default summary prior
described in the introduction. For comparison with the CP-targeted survival
design, keep the same timing, testing indicators, spending families, and
survival assumptions. First construct a matching statistical reference on
the **event-count scale**. The prior's numeric support must be on that same
standardized-effect scale; the normalized `n.fix = 1` reference from the
opening example is not interchangeable with it.

```{r predictive-power-calibration}
pp_reference <- gsDesign(
  k = 3, test.type = 4, timing = c(.5, .75),
  n.fix = surv_design$n.fix,
  delta0 = surv_design$delta0, delta1 = surv_design$delta1,
  sfu = sfLDOF, sfl = sfHSD, sflpar = 1,
  testLower = c(TRUE, FALSE, FALSE)
)
prior <- normalGrid(
  mu = pp_reference$delta / 2, sigma = 10 / sqrt(pp_reference$n.fix)
)
fit_pp <- gsPPFutilitySpending(pp_reference, target_pp = .3, i = 1, prior = prior)
surv_pp <- gsSurv(
  k = 3, test.type = 4, timing = c(.5, .75),
  sfu = sfLDOF, sfl = sfHSD,
  sflpar = fit_pp$ppFutilitySpending$sflpar,
  testLower = c(TRUE, FALSE, FALSE), ratio = 1
)
achieved_pp <- gsPP(
  surv_pp, i = 1, zi = surv_pp$lower$bound[1],
  theta = prior$z, wgts = prior$wgts
)
stopifnot(abs(achieved_pp - .3) <= 1e-4)
stopifnot(abs(sum(surv_pp$upper$prob[, 2]) - .9) < 2e-5)
# Verify that the explicit prior is also this survival design's default prior.
stopifnot(isTRUE(all.equal(
  prior, normalGrid(mu = surv_pp$delta / 2, sigma = 10 / sqrt(surv_pp$n.fix))
)))
gsBoundSummary(surv_pp, prior = prior, digits = 4, exclude = "B-value") |>
  lt() |> lt_format(columns = c("Efficacy", "Futility"), decimals = 4)
```

The **PP** row, not the CP or CP H1 row, now attains 0.3 at the first futility
bound. The prior stays fixed during fitting, but each candidate's posterior
uses that candidate's bound and information. As with the CP example, this is
statistical calibration followed by survival reconstruction, not direct
calibration of a survival object.

```{r compare-cp-pp-survival}
survival_comparison <- function(design, label) {
  data.frame(
    Target = label,
    FutilityHR = gsHR(design$lower$bound[1], 1, design),
    BetaSpending = design$lower$spend[1],
    IA1Events = design$n.I[1], FinalEvents = max(design$n.I),
    MaximumN = max(design$N)
  )
}
rbind(
  survival_comparison(surv_design, "CP = 0.3"),
  survival_comparison(surv_pp, "PP = 0.3")
) |> lt() |>
  lt_format(columns = c("FutilityHR", "BetaSpending"), decimals = 4) |>
  lt_format(columns = c("IA1Events", "FinalEvents", "MaximumN"), decimals = 1)
```

Equal numerical targets for CP and PP need not imply equal futility bounds
or required event counts and sample sizes. The earlier cautions about beta
spending and sample-size inflation apply to PP targets too; averaging over
a prior does not remove these design trade-offs.

### Fixed calendar analyses with `gsSurvCalendar()`

For analyses at months 18, 27, and 36, the information fractions are generally
not the calendar fractions, nor the `.5, .75` fractions above. Derive them
from a reference calendar design before calibrating. This example has a
12-month control median, hazard ratio 0.7, equal randomization, and a fixed
36-month trial with 12 months of minimum follow-up. Enrollment is scaled to
power the design; the last enrollment period is extended as needed to fill
the 24-month enrollment window.

The first calendar analysis below occurs before 50% of expected information.
A CP target of 0.3 also looks too stringent here. We therefore lower it to
0.2 and inspect the resulting bounds; this is a more cautious starting point,
not an automatic endorsement of the design.

```{r calendar-reference}
calendar_args <- list(
  calendarTime = c(18, 27, 36), spending = "information",
  test.type = 4, sfu = sfLDOF, sfl = sfHSD, sflpar = 1,
  testLower = c(TRUE, FALSE, FALSE),
  lambdaC = log(2) / 12, hr = .7, ratio = 1,
  R = 12, minfup = 12
)
calendar_reference <- do.call(gsSurvCalendar, calendar_args)
```

Direct calibration of this object is also unsupported (it has class
`gsSurv`). Instead, construct a statistical `gsDesign()` reference with its
information fractions, spending times, fixed-design event count, and effect
scale. This is an explicit reconstruction, not removal of the survival class
to bypass validation.

```{r calendar-calibration}
calendar_statistical <- gsDesign(
  k = calendar_reference$k, test.type = 4,
  alpha = calendar_reference$alpha, beta = calendar_reference$beta,
  n.fix = calendar_reference$n.fix, timing = calendar_reference$timing,
  usTime = calendar_reference$upper$sTime,
  lsTime = calendar_reference$lower$sTime,
  delta0 = calendar_reference$delta0, delta1 = calendar_reference$delta1,
  sfu = sfLDOF, sfl = sfHSD, sflpar = 1,
  testLower = c(TRUE, FALSE, FALSE)
)
calendar_fit <- gsCPFutilitySpending(calendar_statistical, target_cp = .2, i = 1)
calendar_args$sflpar <- calendar_fit$cpFutilitySpending$sflpar
calendar_design <- do.call(gsSurvCalendar, calendar_args)
stopifnot(max(abs(calendar_design$timing - calendar_reference$timing)) < 1e-6)
check_cp(calendar_design, .2) |> lt()
data.frame(
  Analysis = seq_len(calendar_design$k),
  Month = calendar_design$T,
  InformationFraction = calendar_design$timing,
  Events = round(calendar_design$n.I, 1)
) |> lt()
gsBoundSummary(calendar_design, digits = 4, exclude = "B-value") |>
  lt() |> lt_format(columns = c("Efficacy", "Futility"), decimals = 4)
```

This works because proportional enrollment scaling at fixed calendar times
preserves the expected information fractions in this example. We explicitly
verify that property and the resulting CP. It is not a general calibration
method for designs that change enrollment shape, follow-up, or information
fractions during optimization. We use information-based spending here;
calendar-based spending would require retaining its distinct spending times
as well. Do not simply reuse `fit1`'s parameter with different information
fractions.

## Two parameters: a risk-difference design

We now enable futility at both interims. A two-parameter spending function
allows us to specify two CP targets. Consider adverse-event probabilities
0.15 on control and 0.10 on experimental treatment, with equal randomization.
The benefit is the risk difference `p[C] - p[E] = 0.05`. `nBinomial()` gives
the fixed-design total sample size; `gsDesign()` inflates it for sequential
monitoring. Alpha is 0.025 one-sided and power is 90% in both calculations.
Use `sfCauchy` for the worked example and compare all six families below.
At 40% information, CP 0.3 would also be a stringent first futility target.
We instead use CP 0.2 at 40% and CP 0.3 at 75% information.

```{r two-parameter}
n_binomial <- nBinomial(p1 = .15, p2 = .10, alpha = .025, beta = .1, ratio = 1)
n_binomial
x2 <- gsDesign(
  k = 3, test.type = 4, timing = c(.4, .75), sfu = sfLDOF,
  n.fix = n_binomial, delta0 = 0, delta1 = .05, endpoint = "Binomial",
  testLower = c(TRUE, TRUE, FALSE)
)
target2 <- c(.2, .3)
fit2 <- gsCPFutilitySpending(x2, target_cp = target2, i = 1:2, sfl = sfCauchy)
check_cp(fit2, target2) |> lt()
```

Use the fitted parameter vector directly in a fresh `gsDesign()` call; do not
copy a rounded parameter vector from printed output.

```{r reconstruct-two}
final2 <- gsDesign(
  k = 3, test.type = 4, timing = c(.4, .75),
  n.fix = n_binomial, delta0 = 0, delta1 = .05, endpoint = "Binomial",
  sfu = sfLDOF, sfl = sfCauchy,
  sflpar = fit2$cpFutilitySpending$sflpar,
  testLower = c(TRUE, TRUE, FALSE)
)
check_cp(final2, target2) |> lt()
gsBoundSummary(final2, deltaname = "p[C] - p[E]", Nname = "Participants",
               digits = 4, exclude = "B-value") |>
  lt() |> lt_format(columns = c("Efficacy", "Futility"), decimals = 4)
```

### Fixed sample size and inflation from futility

The fixed-design total is `r round(n_binomial, 1)` participants across both
groups. For comparison, remove futility but keep the same efficacy spending,
timing, alpha, and power. The difference between this efficacy-only design
and the calibrated design shows the additional maximum sample size associated
with the futility rule.

```{r binomial-sample-size}
binomial_efficacy_only <- gsDesign(
  k = 3, test.type = 1, timing = c(.4, .75), sfu = sfLDOF,
  n.fix = n_binomial, delta0 = 0, delta1 = .05, endpoint = "Binomial"
)
binomial_sizes <- c(n_binomial, max(binomial_efficacy_only$n.I), max(final2$n.I))
data.frame(
  Design = c("Fixed, no interims", "Efficacy only", "Efficacy and CP-based futility"),
  MaximumTotalN = binomial_sizes,
  InflationPercent = 100 * (binomial_sizes / n_binomial - 1)
) |> lt() |> lt_format(columns = c("MaximumTotalN", "InflationPercent"), decimals = 1)
```

Here the maximum sample size with calibrated futility is
`r round(max(final2$n.I) / n_binomial, 2)` times the fixed-design size.
The comparison attributes most of the inflation to retaining 90% power in
the presence of the futility bounds, rather than to the efficacy monitoring
alone. These are unrounded planning sample sizes, not expected enrollment:
early stopping can reduce the average number of participants used.

### What does delta at the bound mean?

It is the effect estimate corresponding to the boundary Z-statistic, not the
alternative effect used to power the trial. `delta1 = .05` and `delta0 = 0`
tell the summary how to translate the canonical effect into a risk difference.
The transformation used by `gsDelta()` is

\[
\widehat\delta_i = \delta_0 +
 (\delta_1-\delta_0)\frac{Z_i / \sqrt{n.I_i}}{\texttt{x\$delta}}.
\]

Here `x$delta` is the standardized alternative, whereas `delta1` is the
clinical effect. Thus setting `n.fix` alone supplies a sample-size scale but
does not supply the clinical effect scale. For binomial data these boundary
risk differences are normal-approximation translations using design-based
information, not exact binomial boundaries.

```{r risk-difference-at-bound}
data.frame(
  Analysis = 1:2,
  Participants = round(final2$n.I[1:2], 1),
  FutilityRiskDifference = gsDelta(final2$lower$bound[1:2], 1:2, final2),
  EfficacyRiskDifference = gsDelta(final2$upper$bound[1:2], 1:2, final2)
) |> lt()
```

At a futility boundary, the default CP calculation assumes that this observed
effect continues for future observations. It need not equal the planned
0.05 risk difference, and the same CP at different looks need not imply the
same boundary effect. `CP H1` instead uses the planned alternative. All
sample sizes here are unrounded; integer conversion can change achieved CP.

### Compare all six two-parameter families

The same two targets are feasible for each supported two-parameter family in
this example. Function names and function objects are both accepted.

```{r compare-families}
families <- c(
  "sfLogistic", "sfBetaDist", "sfCauchy", "sfNormal",
  "sfExtremeValue", "sfExtremeValue2"
)
fits2 <- setNames(lapply(families, function(family) {
  if (family == "sfCauchy") return(fit2)
  gsCPFutilitySpending(x2, target_cp = target2, i = 1:2, sfl = family)
}), families)
invisible(lapply(fits2, check_cp, target = target2))
spending_curves <- do.call(rbind, lapply(names(fits2), function(family) {
  fit <- fits2[[family]]
  times <- sort(unique(c(seq(0, 1, length.out = 501), .4, .75)))
  data.frame(
    Family = family, Time = times,
    Spending = fit$lower$sf(fit$beta, times, fit$lower$param)$spend
  )
}))
```

The fitted parameter vectors differ and have different meanings across
families. The spending curves are more informative: they meet, up to numerical
tolerance, at the two specified interim times, but have different shapes
around them. The black points mark the fitted `sfCauchy` spending at those looks.

```{r spending-plot, fig.cap="Cumulative Type II error spending for six calibrated families. Dashed lines mark the interim information fractions; black points mark their common spending, up to calibration tolerance."}
spending_plot <- ggplot(spending_curves, aes(Time, Spending, colour = Family)) +
  geom_vline(xintercept = c(.4, .75), linetype = "dashed", colour = "grey65") +
  geom_line(linewidth = .8) +
  geom_point(
    data = subset(spending_curves, Family == "sfCauchy" & Time %in% c(.4, .75)),
    colour = "black", size = 2.2
  ) +
  scale_colour_manual(values = c(
    sfLogistic = "#0072B2", sfBetaDist = "#D55E00", sfCauchy = "#7B3294",
    sfNormal = "#009E73", sfExtremeValue = "#E69F00", sfExtremeValue2 = "#CC79A7"
  )) +
  scale_x_continuous(breaks = c(0, .25, .4, .75, 1)) +
  labs(x = "Information fraction", y = "Cumulative Type II error spending",
       colour = "Spending family") +
  theme_classic(base_size = 12) +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(nrow = 2, byrow = TRUE))
spending_plot
```

All six curves are similar over the range in which these interim analyses
are most likely to occur, around and between the planned information
fractions. The distinctive feature of `sfCauchy` is its spending when the
first interim occurs substantially earlier than planned: in this example it
spends less Type II error there, while smoothly approaching the same spending
at the planned looks. That is an attractive shape when considering an early
first analysis. All six curves are smooth inside (0, 1); Cauchy is not uniquely
smooth. Adding looks or changing their timing requires recalibration if the
CP targets are to be retained. The plot describes the spending rule away
from the planned looks, not a guarantee of the targeted CP at an unplanned early look.

## Piecewise-linear spending: comparing two normal means

Consider equal-sized groups with a planned mean difference of 0.5 and common
standard deviation 1.1. `nNormal()` supplies the fixed-design total sample
size for a one-sided alpha of 0.025 and 90% power. This uses a known-variance
normal approximation, not a sequential t-test with estimated variance.

With three targets, use `sfLinear`. Its knot times are fixed at the targeted
lower spending times. The free parameters are cumulative proportions of the
total lower spending, constrained to be strictly increasing in (0, 1).

The starting proportions below help the numerical search. They are **not**
CP targets and are **not** fixed final spending values. They may be omitted
to use automatically derived starts; suitable explicit starts can speed up
convergence.

```{r piecewise-linear}
n_normal <- nNormal(delta1 = .5, sd = 1.1, alpha = .025, beta = .1, ratio = 1)
n_normal
x3 <- gsDesign(
  k = 4, test.type = 4, timing = c(.3, .5, .7), sfu = sfLDOF,
  n.fix = n_normal, delta0 = 0, delta1 = .5, endpoint = "Normal",
  testLower = c(TRUE, TRUE, TRUE, FALSE)
)
target3 <- c(.2, .3, .3)
fit3 <- gsCPFutilitySpending(
  x3, target_cp = target3, i = 1:3, sfl = sfLinear,
  control = list(start = c(.85, .95, .98))
)
check_cp(fit3, target3) |> lt()
fit3$cpFutilitySpending$sflpar
```

For `sfLinear`, the `sflpar` vector contains **all knot times first, followed
by all fitted cumulative spending proportions**. There are six entries here
but only three free parameters. The endpoint (1, 1) is implicit. These
proportions are not the incremental spending amounts and are not CP values.

```{r linear-knots}
knots <- data.frame(
  Analysis = 1:3,
  Participants = round(fit3$n.I[1:3], 1),
  Time = x3$lower$sTime[1:3],
  CumulativeProportion = fit3$cpFutilitySpending$free_parameters
)
knots |> lt()
stopifnot(
  all(knots$CumulativeProportion > 0 & knots$CumulativeProportion < 1),
  all(diff(knots$CumulativeProportion) > 0)
)
```

As with the two-parameter example, verify a newly constructed design with the
entire fitted vector, including the knot times:

```{r reconstruct-three}
final3 <- gsDesign(
  k = 4, test.type = 4, timing = c(.3, .5, .7),
  n.fix = n_normal, delta0 = 0, delta1 = .5, endpoint = "Normal",
  sfu = sfLDOF, sfl = sfLinear,
  sflpar = fit3$cpFutilitySpending$sflpar,
  testLower = c(TRUE, TRUE, TRUE, FALSE)
)
check_cp(final3, target3) |> lt()
gsBoundSummary(final3, deltaname = "Mean difference", Nname = "Participants",
               digits = 4, exclude = "B-value") |>
  lt() |> lt_format(columns = c("Efficacy", "Futility"), decimals = 4)
```

Because `delta1 = .5`, the effect-at-bound rows are in mean-difference units.
They are not standardized mean differences or CP values. Sample sizes are
totals across both groups, with their scale supplied by `nNormal()`.

### Fixed sample size and inflation from futility

The fixed-design total is `r round(n_normal, 1)` participants. Again, compare
efficacy-only monitoring with the design that must also achieve the specified
CP at three futility bounds, while keeping 90% overall power.

```{r normal-sample-size}
normal_efficacy_only <- gsDesign(
  k = 4, test.type = 1, timing = c(.3, .5, .7), sfu = sfLDOF,
  n.fix = n_normal, delta0 = 0, delta1 = .5, endpoint = "Normal"
)
normal_sizes <- c(n_normal, max(normal_efficacy_only$n.I), max(final3$n.I))
data.frame(
  Design = c("Fixed, no interims", "Efficacy only", "Efficacy and CP-based futility"),
  MaximumTotalN = normal_sizes,
  InflationPercent = 100 * (normal_sizes / n_normal - 1)
) |> lt() |> lt_format(columns = c("MaximumTotalN", "InflationPercent"), decimals = 1)
```

The calibrated design requires a maximum sample size
`r round(max(final3$n.I) / n_normal, 2)` times the fixed-design size.
As in the binomial example, most of the increase comes from compensating for
futility stopping under the alternative, not from efficacy monitoring alone.
This illustrates why a lower overall power target can be worth considering
with aggressive futility. A different power target requires recalculating
`nNormal()` and the sequential design with the same new beta, then refitting
the spending parameters. These numbers are maximum, not expected, sample sizes.

### A more conservative sequence of CP targets

The main example uses CP 0.2 at 30% information and CP 0.3 at 50% and 70%.
We can also consider lower targets: here CP 0.1, 0.2, and 0.3 at the three
futility bounds, retaining the same timing and efficacy spending.

```{r different-targets}
target_varying <- c(.1, .2, .3)
fit_varying <- gsCPFutilitySpending(
  x3, target_cp = target_varying, i = 1:3, sfl = "sfLinear",
  control = list(start = c(.85, .95, .98))
)
check_cp(fit_varying, target_varying) |> lt()
gsBoundSummary(fit_varying, deltaname = "Mean difference", Nname = "Participants",
               digits = 4, exclude = "B-value") |>
  lt() |> lt_format(columns = c("Efficacy", "Futility"), decimals = 4)
```

## Practical limits and diagnostics

- A single early futility analysis may often be sufficient. More futility
  looks add opportunities to stop a truly effective treatment and need not
  improve the development strategy.
- Even when several futility looks are planned, fitting a one-parameter
  spending function to the first interim CP target may be adequate. Keep
  later desired futility looks enabled in `testLower`, fit only `i = 1`,
  and inspect every resulting bound, its CP, effect estimate, beta spending,
  and sample-size implications. Later CP values are implied by the fitted
  spending function, not independently targeted. Use extra parameters only
  when the resulting later bounds do not meet the design objectives.
- More targets require more free parameters, not just a larger `i` vector.
  `sfLinear` supports additional targeted interim looks, but arbitrary target
  vectors are not guaranteed feasible under monotone cumulative spending.
- `control$cp_tol` sets the maximum absolute CP residual; `reltol` controls
  internal convergence and does not replace this acceptance check. See
  `?gsCPFutilitySpending` for the complete controls and family-specific limits.
  For `sfLinear`, supply only cumulative proportions in `control$start`, not
  the complete `sflpar`; user-supplied `lower` and `upper` limits are unsupported.
- Inspect `fit$cpFutilitySpending$residual`, `$information_ratio`, and `$solver`
  for accuracy, information changes, and convergence diagnostics. Invalid
  inputs, infeasible targets, and convergence failures have distinct error
  classes; an unsuccessful solve does not return a calibrated design.
- All examples verify unconditional power as well as CP. In a general
  reconstruction, retain all reference settings, including alpha, beta,
  effect assumptions, testing indicators, and any harm-spending inputs.
  Changing those settings or efficacy spending requires recalibration.
- `toInteger()` retains the fitted spending specification, but rounding can
  change achieved CP and does not automatically recalibrate the design.
- The calibration input must currently be a fixed-timing `gsDesign()` object.
  Direct survival or calendar-time calibration is not implemented by either
  calibration function. Use `gsPPFutilitySpending()` for prior-averaged
  predictive power rather than `gsCPFutilitySpending()`.
