---
title: "Futility spending calibrated to assurance and probability of success"
output:
  rmarkdown::html_vignette:
    css: vignette.css
vignette: >
  %\VignetteIndexEntry{Futility spending calibrated to assurance and probability of success}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                     fig.height = 4.8, dpi = 150)
options(width = 90)
```

```{r libraries, message=FALSE}
library(gsDesign)
library(ggplot2)
```

The functions in this vignette translate probability-of-success requirements
into **futility spending functions**. They build on established computational
tools in gsDesign: boundary construction, joint crossing probabilities, and
prior-averaged probabilities. The new work is calibration of the spending
parameters, not replacement of the underlying group sequential calculations.

Three related questions lead to three different designs:

| Function | Probability targeted | What stays fixed? | What may change? |
|:--|:--|:--|:--|
| `gsCPOSFutilitySpending(mode = "preserve_power")` | Conditional POS at selected interims | Planned frequentist power and relative timing | Maximum information |
| `gsCPOSFutilitySpending(mode = "fixed_information")` | Conditional POS at selected interims | Information and efficacy boundaries | Overall power/beta |
| `gsPOSFutilitySpending()` | Unconditional POS for the whole design | Planned frequentist power and relative timing | Maximum information |

All three return a spending family and fitted parameter set, a complete design,
and diagnostics. They differ from `gsCPFutilitySpending()` and
`gsPPFutilitySpending()`, which target probabilities at an exact interim
futility bound. Those methods are described in the
[conditional-power vignette](CPFutilitySpending.html).

## What is conditioned on?

Let \(C_i\) mean that the trial has crossed neither stopping boundary through
interim \(i\), and let \(E_{>i}\) mean a later efficacy rejection. Conditional
assurance is

\[
  \Pr(E_{>i}\mid C_i)
  = \frac{\Pr(E_{>i}\cap C_i)}{\Pr(C_i)}.
\]

Both probabilities average over the prespecified effect prior. Conditioning on
continuation implicitly changes the relative plausibility of the possible
effects. This is the quantity calculated by `gsCPOS()`, including the full
earlier stopping history.

In contrast, `gsPOS()` is the probability of any efficacy rejection before
seeing data; it includes early efficacy. `gsPP()` conditions on a particular
interim statistic and averages future success over its posterior effect
distribution. Thus conditional assurance is **not PP at the futility bound**.
A high assurance among trials allowed to continue does not say that a trial
just at the boundary has high predictive power.

Without early efficacy, conditional assurance is POS divided by the prior
probability of continuation. With early efficacy, its numerator must exclude
those earlier successes. A point prior changes the averaging, but not this
distinction between a continuation event and an exact observed statistic.

## A Dragalin-related fixed-information example

Dragalin's *Informative Futility Rules Based on Conditional Assurance* motivates
comparing assurance among continuing trials with the assurance of a larger,
feasible fixed design. Here we use the paper's normal-outcome setting:
interim sample size 72, maximum sample size 144, equal allocation, common
standard deviation 70, and an effect prior \(N(33,16^2)\). These are total
sample sizes across both treatment groups.

For a difference of normal means with equal allocation, the standardized effect
is the mean difference divided by \(2\,SD=140\). We use that scale throughout.
This is the paper's informative prior, **not** the broader default prior in
`gsBoundSummary()`. We supply it explicitly to every prior-based calculation.

```{r dragalin-reference}
effect_sd <- 70
effect_h1 <- 33
prior <- normalGrid(mu = effect_h1 / (2 * effect_sd),
                    sigma = 16 / (2 * effect_sd))
prior$wgts <- prior$wgts / sum(prior$wgts)

reference <- gsDesign(
  k = 2, test.type = 4, delta = effect_h1 / (2 * effect_sd),
  n.I = c(72, 144), maxn.IPlan = 144,
  testUpper = c(FALSE, TRUE), sfl = sfHSD, sflpar = 0, r = 32
)
```

The benchmark has a single final analysis at \(N=230\) and one-sided alpha
0.025. Since `gsDesign()` constructs sequential designs, we specify the
one-look boundaries directly for probability evaluation. Both terminal bounds
are the same: every trial has a final decision.

```{r fixed-benchmark}
fixed_benchmark <- structure(
  list(k = 1L, test.type = 4L, n.I = 230,
       upper = list(bound = qnorm(.975)),
       lower = list(bound = qnorm(.975)), r = 32L, overrun = 0),
  class = "gsDesign"
)
benchmark <- gsPOS(fixed_benchmark, theta = prior$z, wgts = prior$wgts)
benchmark
```

Compute this benchmark **once**, not again from each candidate design.

```{r dragalin-fit}
fit_dragalin <- gsCPOSFutilitySpending(
  reference, target_cpos = benchmark, i = 1, prior = prior,
  mode = "fixed_information"
)
dragalin_results <- c(
  Target_assurance = benchmark,
  Achieved_assurance = gsCPOS(1, fit_dragalin, prior$z, prior$wgts),
  Interim_N = fit_dragalin$n.I[1],
  Final_N = fit_dragalin$n.I[2],
  Effect_at_futility = fit_dragalin$lower$bound[1] *
    2 * effect_sd / sqrt(fit_dragalin$n.I[1]),
  Overall_power_H1 = sum(fit_dragalin$upper$prob[, 2]),
  Fitted_beta = fit_dragalin$beta,
  Fitted_sflpar = fit_dragalin$cposFutilitySpending$sflpar
)
data.frame(Quantity = names(dragalin_results), Value = unname(dragalin_results)) |>
  knitr::kable(digits = 4)
```

The target is approximately 0.7902. Solving equality gives an effect threshold
of approximately 7.274, rather than the illustrative threshold of 10 in the
paper (which exceeds the benchmark). The sample sizes remain 72 and 144;
the achieved overall power at effect 33 is about 79.3%.

This is a spending-based implementation of the fixed-information assurance
criterion. It is **not** a reproduction of an entire sample-size re-estimation
procedure or an inverse-normal combination test.

## Compare all three methods on the same reference

For a broader comparison, retain the same effect prior, SD and planned effect,
but use a reference with 85% power, information fractions 0.4 and 0.75, and
one active futility analysis at the first interim. There are efficacy analyses
at all three looks. A single early futility analysis is often sufficient;
a one-parameter spending function can also be adequate with additional
futility looks after inspection of every bound.

```{r comparison-reference}
n_fixed <- nNormal(delta1 = effect_h1, sd = effect_sd, beta = .15)
x <- gsDesign(
  k = 3, test.type = 4, n.fix = n_fixed, beta = .15,
  timing = c(.4, .75), delta1 = effect_h1,
  sfu = sfLDOF, sfl = sfHSD, sflpar = -2,
  testLower = c(TRUE, FALSE, FALSE)
)
data.frame(Fixed_design_N = n_fixed, Sequential_reference_N = tail(x$n.I, 1)) |>
  knitr::kable(digits = 2)
```

### Same conditional assurance, different sample-size and power constraints

Both fits below target conditional assurance of 0.80 at the first interim.
This is a continuation-conditioned target, **not CP = 0.80 at a boundary**.
The package calls this conditional probability of success (CPOS). Both fits
use `gsCPOSFutilitySpending()`, with `mode` selecting the design constraint.
`"preserve_power"` is the default. Both modes use `control$cpos_tol` and return
diagnostics in `cposFutilitySpending`. The old `gsCAFutilitySpending()` interface
remains a compatibility wrapper for `mode = "fixed_information"`.

```{r assurance-comparison}
fit_cpos <- gsCPOSFutilitySpending(x, target_cpos = .80, i = 1, prior = prior)
fit_ca <- gsCPOSFutilitySpending(x, target_cpos = .80, i = 1, prior = prior,
                               mode = "fixed_information")

stopifnot(identical(fit_ca$n.I, x$n.I),
          identical(fit_ca$upper$bound, x$upper$bound))
```

The first fit increases information to retain 85% power. The second freezes
the reference information and efficacy boundaries, allowing power to fall.
The fixed-information solver also determines total beta: using the reference
beta unchanged would generally not produce a consistent spending rule at the
fixed final efficacy boundary.

### Unconditional POS is a different requirement

POS is a single scalar for the whole design, so this interface fits one free
parameter and has no interim-index argument. It rejects an unconstrained
two-parameter family; one target cannot determine two parameters uniquely.

```{r pos-comparison}
fit_pos <- gsPOSFutilitySpending(
  x, target_pos = .7206, prior = prior,
  control = list(start = 0, lower = -2, upper = 2, pos_tol = 1e-7)
)
```

The precision here illustrates the solver, not a recommendation that such
small POS differences are practically important. For this prior, preserving
power while changing information largely offsets the effect of changing the
futility rule, leaving POS nearly unchanged. This can make POS calibration
weakly informative about spending shape. The objective may also be nonmonotone:
matching the target does not prove a unique solution. A point prior at the
planned alternative is an extreme example, because POS then equals the power
already being preserved.

### Verify with established probability calculations

These are fresh calculations from the returned bounds and information, rather
than just the optimizer's reported residuals.

```{r independent-verification}
verification <- data.frame(
  Method = c("Power-preserving CA", "Fixed-information CA", "Unconditional POS"),
  Target = c(.80, .80, .7206),
  Achieved = c(gsCPOS(1, fit_cpos, prior$z, prior$wgts),
               gsCPOS(1, fit_ca, prior$z, prior$wgts),
               gsPOS(fit_pos, prior$z, prior$wgts))
)
knitr::kable(verification, digits = 6)
stopifnot(max(abs(verification$Achieved - verification$Target)) < 1e-4)

# Verify that total beta and the fitted lower spending agree.
desired_beta <- fit_ca$lower$sf(
  fit_ca$beta, fit_ca$lower$sTime, fit_ca$lower$param
)$spend
# At the inactive second futility look, spending is deferred.
desired_beta[2] <- desired_beta[1]
stopifnot(max(abs(cumsum(fit_ca$lower$prob[, 2]) - desired_beta)) < 1e-5)
```

For the power-preserving fit, replay the fitted spending parameter in an
ordinary `gsDesign()` call with the complete reference specification:

```{r replay}
replay <- gsDesign(
  k = 3, test.type = 4, n.fix = n_fixed, beta = .15,
  timing = c(.4, .75), delta1 = effect_h1,
  sfu = sfLDOF, sfl = sfHSD,
  sflpar = fit_cpos$cposFutilitySpending$sflpar,
  testLower = c(TRUE, FALSE, FALSE)
)
gsCPOS(1, replay, prior$z, prior$wgts)
```

For the fixed-information fit, replay with `gsCPOSFutilitySpending(mode = "fixed_information")` and the
original reference, using the fitted free parameters as the start. An ordinary
power-preserving reconstruction is not equivalent.

## Operating characteristics, not just a target

```{r operating-characteristics}
designs <- list(Reference = x, "Power-preserving CA" = fit_cpos,
                "Fixed-information CA" = fit_ca, "Unconditional POS" = fit_pos)
operating <- do.call(rbind, lapply(names(designs), function(label) {
  d <- designs[[label]]
  p <- gsProbability(d = d, theta = c(0, x$delta))
  data.frame(
    Method = label, Max_N = tail(d$n.I, 1),
    Inflation_percent = 100 * (tail(d$n.I, 1) / n_fixed - 1),
    Actual_type_I = sum(p$upper$prob[, 1]),
    Power_H1 = sum(p$upper$prob[, 2]),
    POS = gsPOS(d, prior$z, prior$wgts),
    CA_at_IA1 = gsCPOS(1, d, prior$z, prior$wgts),
    Futility_H0 = sum(p$lower$prob[-d$k, 1]),
    Futility_H1 = sum(p$lower$prob[-d$k, 2]),
    EN_H0 = p$en[1], EN_H1 = p$en[2]
  )
}))
knitr::kable(operating[c("Method", "Max_N", "Inflation_percent",
                        "Actual_type_I", "Power_H1")], digits = 4)
knitr::kable(operating[c("Method", "POS", "CA_at_IA1",
                        "Futility_H0", "Futility_H1")], digits = 4)
knitr::kable(operating[c("Method", "EN_H0", "EN_H1")], digits = 2)
```

The fixed design requires about `r round(n_fixed, 1)` participants.
Sample sizes are continuous planning values; rounding requires another check
of the resulting operating characteristics and target probabilities.
The inflation column uses that fixed design, not the sequential reference, as
its denominator. Much of the inflation in these examples is the cost of
preserving power despite futility stopping. A practical maximum, such as 20%
inflation, should be considered alongside the probability target; none of these
functions automatically enforces such a cap.

For nonbinding futility, the nominal efficacy-only alpha remains 0.025, while
actual type I error with futility enforced may be smaller. With a binding
reference (`test.type = 3`), freezing efficacy while changing futility can
increase actual type I error above the reference alpha. The fixed-information
method reports that probability; it does not guarantee alpha preservation.

```{r power-curves}
effect <- seq(-10, 66, length.out = 101)
curves <- do.call(rbind, lapply(names(designs), function(label) {
  p <- gsProbability(d = designs[[label]], theta = effect / (2 * effect_sd))
  data.frame(Effect = effect, Power = colSums(p$upper$prob), Method = label)
}))
ggplot(curves, aes(Effect, Power, colour = Method, linetype = Method)) +
  geom_line(linewidth = .8) +
  geom_vline(xintercept = effect_h1, colour = "grey60", linetype = 3) +
  labs(x = "True difference in means", y = "Probability of efficacy rejection") +
  theme_minimal() + theme(legend.position = "bottom") +
  guides(colour = guide_legend(nrow = 2), linetype = guide_legend(nrow = 2))
```

### Sample sizes and approximate effects at all bounds

A probability target is not a substitute for judging whether the effect at a
futility boundary is clinically reasonable. Approximate mean differences below
are \(Z \times 2\,SD/\sqrt{N}\); they are normal-theory approximations, not
bias-adjusted sequential estimates. Inactive futility analyses have no reported
futility effect.

```{r bound-effects}
bounds <- do.call(rbind, lapply(names(designs), function(label) {
  d <- designs[[label]]
  lower_effect <- d$lower$bound * 2 * effect_sd / sqrt(d$n.I)
  lower_effect[!d$testLower] <- NA_real_
  data.frame(Method = label, Analysis = seq_len(d$k), N = d$n.I,
             Efficacy_effect = d$upper$bound * 2 * effect_sd / sqrt(d$n.I),
             Futility_effect = lower_effect)
}))
knitr::kable(bounds, digits = 3)
```

The standard summary adds CP, CP H1 and PP at the bounds. The prior here remains
the explicit Dragalin prior. Those bound-based quantities should not be mistaken
for the continuation-conditioned targets above.

```{r bound-summary}
gsBoundSummary(fit_ca, prior = prior, exclude = "B-value", digits = 4)
```

Here CP at the futility bound is about 0.081 and PP is about 0.345, even though
conditional assurance among continuing trials is 0.80. This concrete contrast
is why the conditioning event needs to be stated whenever a target is quoted.

## The final specification is still a spending function

```{r spending-curves}
times <- seq(0, 1, length.out = 301)
spending <- do.call(rbind, lapply(names(designs), function(label) {
  d <- designs[[label]]
  data.frame(Time = times, Spend = d$lower$sf(d$beta, times, d$lower$param)$spend,
             Method = label)
}))
ggplot(spending, aes(Time, Spend, colour = Method, linetype = Method)) +
  geom_line(linewidth = .8) +
  geom_vline(xintercept = c(.4, .75), colour = "grey70", linetype = 3) +
  labs(x = "Information fraction", y = "Cumulative beta spending") +
  theme_minimal() + theme(legend.position = "bottom") +
  guides(colour = guide_legend(nrow = 2), linetype = guide_legend(nrow = 2))
```

These curves need not cross at the calibration time: the target is conditional
assurance or POS, not cumulative beta spending. The fixed-information design
can also have a different total beta. The curves describe the spending family;
actual spending is deferred at inactive futility analyses.

Spending functions retain the original motivation of the Lan--DeMets approach:
the specification can accommodate different information times and, under an
appropriate monitoring plan, a different number of analyses. They provide a
continuous rule between planned looks, a compact reproducible specification,
and a framework for recalculating boundaries while respecting spending already
used. This flexibility is particularly useful when actual information accrues
differently from the original plan.

It does **not** mean that one should refit parameters freely after seeing
treatment effects, or that changing the monitoring schedule preserves the
original assurance target or power exactly. Re-evaluate the resulting design
and its operating characteristics. Timing changes informed by unblinded
outcomes need additional justification.

## Building on established computational tools with AI

The calibration framework builds on existing gsDesign tools rather than new
probability-integration machinery. Most of the initial CP/PP development was
done interactively with AI using **GPT-6 Astra High in Codex over a long
weekend**. The later assurance extensions follow the same principle:
rapidly build new design interfaces on established computational tools.

Validation need not rest solely on line-by-line inspection of generated code.
Reconstruct fitted designs and check their properties with `gsDesign()`,
`gsProbability()`, `gsCPOS()`, `gsPOS()` and `gsBoundSummary()`.
Unit tests also check independent bivariate-normal integration, the numerical
Dragalin benchmark, parameter replay, input errors and numerical failure
handling. These checks support the statistical results; they do not establish
that every possible software defect has been excluded.

## Scope and practical limits

- Current assurance/POS calibration supports fixed-timing statistical
  `gsDesign` objects with test types 3 and 4, not direct survival objects.
- Harm-bound designs are excluded: current `gsCPOS()` does not include harm
  stopping in its continuation denominator.
- Conditional-assurance fits support the existing one-/two-parameter families
  and piecewise linear spending, with one target per free parameter.
- POS has only one design-wide target and fits one free parameter.
- Prior weights are normalized, and the prior is fixed throughout calibration.
  Continuation probabilities at or below `sqrt(.Machine$double.eps)` are
  rejected for conditional assurance.
- A failed numerical search is not a proof of mathematical infeasibility.
  A matched target is not proof that the design is clinically attractive.
- Review overall power, actual type I error, expected and maximum sample size,
  stopping probabilities, effect sizes at bounds, and sensitivity to the prior
  and monitoring schedule before choosing a design.

## References

Dragalin V. *Informative Futility Rules Based on Conditional Assurance*.
Statistics in Medicine (2026).
[doi:10.1002/sim.70330](https://doi.org/10.1002/sim.70330).

Lan KKG, DeMets DL. *Discrete sequential boundaries for clinical trials*.
Biometrika (1983), 70:659--663.
[doi:10.1093/biomet/70.3.659](https://doi.org/10.1093/biomet/70.3.659).
