---
title: "Calibrating spending to effects at interim bounds"
output:
  rmarkdown::html_vignette:
    css: vignette.css
vignette: >
  %\VignetteIndexEntry{Calibrating spending to effects at interim bounds}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                     fig.height = 4.8, dpi = 150)
options(width = 85)
```

```{r libraries, message=FALSE}
library(gsDesign)
library(ggplot2)
```

## Specify an effect; retain a spending function

**gsEffectSpending()** selects spending parameters so that approximate observed
treatment effects at specified interim efficacy, futility or harm bounds match
natural-scale targets. A single function supports one boundary or a joint fit.
It builds on the established gsDesign boundary and probability calculations.

This answers a different question from CP, PP or conditional assurance:
"What observed treatment effect should correspond to this stopping boundary?"
The result still specifies spending functions, rather than only isolated
boundary values. Planned power and nominal error budgets are retained by
recalculating information. Maximum sample size may therefore change.

Effects at bounds are normal-theory approximations, not bias-adjusted
sequential estimates. The candidate's information must be used in each
transformation. A target cannot generally be converted to a fixed Z boundary
once, using the original sample size.

## Joint efficacy, futility and harm targets for normal means

Consider equal allocation, a common standard deviation of 1, and an
alternative mean difference of 0.2. Positive differences favor the experimental
treatment. Information fractions are 0.4 and 0.75, with 90% power and
nonbinding futility/harm (test type 8).

```{r reference}
n_fixed <- nNormal(delta1 = .2, sd = 1)
x <- gsDesign(
  k = 3, test.type = 8, n.fix = n_fixed, delta1 = .2,
  timing = c(.4, .75), sfu = sfHSD, sfupar = 0,
  sfl = sfHSD, sflpar = 0, sfharm = sfHSD, sfharmparam = 0,
  astar = .05
)
n_fixed
```

The following targets were chosen near a known feasible configuration to
illustrate joint calibration, not as clinical recommendations. Specify them
directly as differences in means, not as Z values or standardized effects.

```{r joint-fit}
fit <- gsEffectSpending(
  x,
  target_effect = c(.185346, .064696, -.161755),
  i = c(1, 1, 1),
  bound = c("efficacy", "futility", "harm"),
  spending = list(efficacy = sfHSD, futility = sfHSD, harm = sfHSD),
  effect = list(endpoint = "mean_difference"),
  control = list(effect_tol = 1e-5)
)
knitr::kable(fit$effectSpending$targets, digits = 6)
fit$effectSpending$free_parameters
```

Efficacy requires evidence of benefit; futility stops for an insufficiently
promising result; harm is further in the unfavorable direction. These are
three targets on the **same endpoint scale**. Their numerical values and
clinical interpretation must be assessed together.

Fitting one boundary and then another need not preserve the first match:
information and boundaries are coupled. The joint fit checks all requested
effects simultaneously.

### Independent reconstruction and verification

```{r replay}
replay <- do.call(gsDesign, fit$effectSpending$replay)
achieved <- gsDelta(
  c(replay$upper$bound[1], replay$lower$bound[1], replay$harm$bound[1]),
  i = c(1, 1, 1), x = replay
)
stopifnot(max(abs(achieved - fit$effectSpending$targets$target_effect)) < 1e-5)

p <- gsProbability(d = replay, theta = c(0, replay$delta))
knitr::kable(data.frame(
  Quantity = c("Fixed-design N", "Reference maximum N", "Fitted maximum N",
               "Inflation over fixed design (%)", "Actual type I error",
               "Power at planned alternative"),
  Value = c(n_fixed, tail(x$n.I, 1), tail(replay$n.I, 1),
            100 * (tail(replay$n.I, 1) / n_fixed - 1),
            sum(p$upper$prob[, 1]), sum(p$upper$prob[, 2]))
), digits = 4)
```

Actual type I error with lower stopping enforced can differ from the nominal
efficacy budget. Check the design's binding/nonbinding conventions, not merely
its stored alpha. A successful effect match also does not establish that its
sample-size inflation is acceptable; a practical limit such as 20% over a
fixed design may be part of design selection.

Sample sizes here are continuous planning values. After rounding, recheck
effects and operating characteristics.

### All bounds, not only the targeted interim

```{r all-bounds}
bounds <- data.frame(
  Analysis = seq_len(replay$k), N = replay$n.I,
  Efficacy_effect = gsDelta(replay$upper$bound, seq_len(replay$k), replay),
  Futility_effect = gsDelta(replay$lower$bound, seq_len(replay$k), replay),
  Harm_effect = gsDelta(replay$harm$bound, seq_len(replay$k), replay)
)
knitr::kable(bounds, digits = 4)
gsBoundSummary(replay, exclude = "B-value", digits = 4)
```

CP, CP H1 and PP in the summary give additional perspectives on these
boundaries; none was directly targeted. The default PP prior is the broad
normal prior used by gsBoundSummary(), centered halfway between the null and
planned alternative on the standardized-effect scale.

## Compare the fitted spending shapes

The three spending budgets have different interpretations and totals. The
graph uses fractions of each budget to compare **shape**, not to imply that
alpha, beta and harm probabilities are interchangeable.

```{r spending-curves}
times <- seq(0, 1, length.out = 201)
slots <- c(efficacy = "upper", futility = "lower", harm = "harm")
budget <- c(efficacy = replay$alpha, futility = replay$beta, harm = replay$astar)
curves <- do.call(rbind, lapply(names(slots), function(b) {
  s <- replay[[slots[[b]]]]
  data.frame(Time = times,
             Fraction = s$sf(budget[b], times, s$param)$spend / budget[b],
             Boundary = b)
}))
ggplot(curves, aes(Time, Fraction, colour = Boundary, linetype = Boundary)) +
  geom_line(linewidth = .9) +
  geom_vline(xintercept = c(.4, .75), linetype = 3, colour = "grey65") +
  labs(x = "Information fraction", y = "Fraction of corresponding spending budget") +
  theme_minimal() + theme(legend.position = "bottom")
```

These illustrative targets are close to the same HSD shape for all three
boundaries, so the normalized curves nearly coincide. Other clinical targets
can produce different shapes.

Spending functions retain their usefulness when information accrues at
different times or the monitoring plan changes. Preserve spending already
used and reevaluate the revised design: the original effect targets and power
are not guaranteed to remain exact. This flexibility does not authorize
unrestricted refitting after observing unblinded results.

## Two-parameter futility spending for a risk difference

For a response endpoint, let the control response probability be 0.1 and
experimental response probability 0.2. The alternative risk difference is 0.1.
nBinomial() supplies the fixed sample size; delta0 and delta1 define the
natural-scale interpretation.

```{r risk-difference}
n_binomial <- nBinomial(p1 = .1, p2 = .2)
rd <- gsDesign(
  k = 3, test.type = 4, timing = c(.4, .75),
  n.fix = n_binomial, delta0 = 0, delta1 = .1,
  sfl = sfLogistic, sflpar = c(0, 1)
)
# Use attainable reference effects to demonstrate parameter recovery.
rd_targets <- gsDelta(rd$lower$bound[1:2], 1:2, rd)
rd_fit <- gsEffectSpending(
  rd, rd_targets, i = 1:2, spending = sfLogistic,
  effect = list(endpoint = "risk_difference"),
  control = list(start = c(-.1, 1.1))
)
knitr::kable(rd_fit$effectSpending$targets, digits = 5)
knitr::kable(data.frame(
  Fixed_N = n_binomial, Maximum_N = tail(rd_fit$n.I, 1),
  Inflation_percent = 100 * (tail(rd_fit$n.I, 1) / n_binomial - 1)
), digits = 2)
```

One target per free parameter is required for each selected family. Piecewise
linear spending is also supported, with knots fixed at the selected spending
times. Equal numbers of parameters and targets do not prove uniqueness.

The risk-difference mapping is approximate: it does not identify an exact
observed 2-by-2 table. Invalid risk-difference targets are rejected, and
out-of-range achieved approximations are not silently clipped.

## Hazard ratios require explicit event-scale metadata

An HR transformation must distinguish event information from patient sample
size. Here ratio = 2 is experimental/control allocation, the null HR is 1.1
and the alternative HR is 0.7. The canonical drift per square root of total
events is determined by the HR contrast and allocation.

```{r hazard-ratio}
hr_effect <- list(information = "events", ratio = 2, hr0 = 1.1, hr1 = .7)
delta_events <- abs(log(.7 / 1.1)) * sqrt(2) / 3
hr_reference <- gsDesign(k = 3, test.type = 4, delta = delta_events, sflpar = 1)
hr_for_summary <- hr_reference
hr_for_summary$hr <- .7
hr_for_summary$hr0 <- 1.1
hr_target <- gsHR(hr_reference$lower$bound[1], 1, hr_for_summary, ratio = 2)

hr_fit <- gsEffectSpending(
  hr_reference, hr_target, scale = "hr", effect = hr_effect,
  control = list(start = 0)
)
knitr::kable(hr_fit$effectSpending$targets, digits = 5)
data.frame(
  Analysis = seq_len(hr_fit$k), Events = hr_fit$n.I,
  Futility_HR = gsHR(hr_fit$lower$bound, seq_len(hr_fit$k), hr_fit, ratio = 2)
) |> knitr::kable(digits = 4)
```

The function checks the event-scale drift against the supplied HR metadata.
It also supports an alternative HR above the null; the effect direction is
explicit, rather than assuming HR below 1 is always the favorable direction.

This is a statistical event-driven design, not an accrual/calendar-time
survival design. Direct gsSurv()/gsSurvCalendar() objects remain unsupported.
An accrual model and its reconstruction require separate validation.

For risk ratios, use scale = "rr" with correctly specified **log-ratio**
delta0/delta1 reference metadata, but supply target_effect as a ratio.

## Diagnostics and interpretation

```{r diagnostics}
fit$effectSpending[c("local_rank", "locally_identified", "condition_number",
                     "harm_futility_coincidence")]
```

The numerical Jacobian describes local sensitivity only. A capped harm bound
may be unchanged when its spending parameter changes, and different parameter
sets may satisfy the same targets. Rank/conditioning diagnostics and
harm/futility coincidence help identify such concerns, but do not establish
global uniqueness.

The effect tolerance is in natural units, not probability units. Failed
searches return diagnostic errors, not a certificate of mathematical
infeasibility. Review clinical relevance, prior-independent operating
characteristics, sample size, and every boundary before choosing a design.

## Built on established computational tools

The new interface reuses gsDesign(), spending families, effect-at-bound
transformations and joint probability calculations. Its innovation is the
clinical-scale calibration request and coordinated parameter search.
Tests reconstruct designs and verify their properties using these existing
tools, including joint targets, ratio direction and allocation, and error
handling.

This is the same development principle as the CP/PP and assurance methods:
rapid AI-assisted development **on top of established computational tools**,
with independently checkable statistical properties. For related methods,
see [CP/PP spending](CPFutilitySpending.html) and
[assurance/POS spending](DragalinFutilitySpending.html).
