---
title: "Using gsDesign AI skills"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using gsDesign AI skills}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

The `gsDesign` repository includes lightweight AI workflow instructions for
coding agents. These files are not statistical methods documentation. They are
promptable development guides that help coding agents use consistent package
maintenance habits: checking git status, updating `NEWS.md`, bumping development
versions, running focused and broad tests, rebuilding pkgdown, and pushing
issue branches.

The repository includes native instruction files for common coding agents:

- `AGENTS.md` for Codex-style agents.
- `CLAUDE.md` for Claude Code.
- `GEMINI.md` for Gemini CLI.
- `.github/copilot-instructions.md` for GitHub Copilot.
- `.gemini/styleguide.md` for Gemini Code Assist.

The installed R package also ships copies under `inst/ai/`, available with:

```{r, eval=FALSE}
system.file("ai", package = "gsDesign")
```

This makes the instructions available to users who install the package without
cloning the GitHub repository.

## Installing instructions in a project

After installing `gsDesign`, you can copy an instruction file into a local
project. For example, to add Codex-style instructions:

```{r, eval=FALSE}
file.copy(
  from = system.file("ai", "AGENTS.md", package = "gsDesign"),
  to = "AGENTS.md",
  overwrite = FALSE
)
```

For GitHub Copilot:

```{r, eval=FALSE}
dir.create(".github", showWarnings = FALSE)
file.copy(
  from = system.file("ai", "copilot-instructions.md", package = "gsDesign"),
  to = file.path(".github", "copilot-instructions.md"),
  overwrite = FALSE
)
```

For Gemini Code Assist:

```{r, eval=FALSE}
dir.create(".gemini", showWarnings = FALSE)
file.copy(
  from = system.file("ai", "gemini-styleguide.md", package = "gsDesign"),
  to = file.path(".gemini", "styleguide.md"),
  overwrite = FALSE
)
```

## Example prompt and response

An AI skill works best when the user gives a specific development task and lets
the instruction file supply the package workflow details.

Example prompt:

> Use the gsDesign workflow instructions. Fix issue #264 on the current branch:
> update `simBinomialSeasonalExact()` so simulated trials stop at efficacy or
> futility boundary crossings, update `toInteger()` for rare-event equal
> allocation, add regression tests, update `NEWS.md`, rebuild pkgdown, run the
> package checks, commit, and push if everything is OK.

Example response:

> I checked the branch and worktree, implemented the simulation stopping and
> event-rounding updates, added focused regression tests, bumped the package to
> the next development version, updated `NEWS.md`, rebuilt pkgdown, and ran
> `R CMD check --no-manual`. The only remaining check warning is a local
> macOS/R toolchain warning from `R_ext/Boolean.h`, not package code. I committed
> the changes and pushed the issue branch to `origin`.

The response is useful because it reports the workflow steps the agent took and
separates package-code results from local toolchain noise.

## What these instructions do not do

AI workflow instructions do not change the statistical behavior of `gsDesign`.
They also do not replace human review. They help make agent-assisted package
maintenance more reproducible by placing the expected repository workflow where
coding agents can find it.
