---
title: "Using gsDesign AI skills"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using gsDesign AI skills}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

The `gsDesign` repository includes lightweight AI workflow instructions for
coding agents. These files are not full statistical methods documentation. They
are promptable development guides that help coding agents use consistent package
maintenance habits: checking git status, updating `NEWS.md`, bumping development
versions, running focused and broad tests, rebuilding pkgdown, and pushing
issue branches. They also include brief function-selection cues for common
design requests, such as using `gsSurvCalendar()` when the prompt fixes analyses
at calendar months after enrollment opens, while preserving the original design
specifications unless the user asks to change them. Very low planned event
counts can cue a discussion of exact-binomial rare-event methods, but should not
by itself override a calendar-timed survival design request.

The repository includes native instruction files for common coding agents:

- `AGENTS.md` for Codex-style agents.
- `CLAUDE.md` for Claude Code.
- `GEMINI.md` for Gemini CLI.
- `.github/copilot-instructions.md` for GitHub Copilot.
- `.gemini/styleguide.md` for Gemini Code Assist.

The installed R package also ships copies under `inst/ai/`, available with:

```{r, eval=FALSE}
system.file("ai", package = "gsDesign")
```

This makes the instructions available to users who install the package without
cloning the GitHub repository.

## Installing instructions in a project

After installing `gsDesign`, you can copy an instruction file into a local
project. For example, to add Codex-style instructions:

```{r, eval=FALSE}
file.copy(
  from = system.file("ai", "AGENTS.md", package = "gsDesign"),
  to = "AGENTS.md",
  overwrite = FALSE
)
```

For GitHub Copilot:

```{r, eval=FALSE}
dir.create(".github", showWarnings = FALSE)
file.copy(
  from = system.file("ai", "copilot-instructions.md", package = "gsDesign"),
  to = file.path(".github", "copilot-instructions.md"),
  overwrite = FALSE
)
```

For Gemini Code Assist:

```{r, eval=FALSE}
dir.create(".gemini", showWarnings = FALSE)
file.copy(
  from = system.file("ai", "gemini-styleguide.md", package = "gsDesign"),
  to = file.path(".gemini", "styleguide.md"),
  overwrite = FALSE
)
```

## Example prompt and response

An AI skill works best when the user gives a specific package-development or
design-coding task and lets the instruction file supply the package workflow
and routing details.

Example prompt:

> Use the gsDesign AI instructions. I am designing a standard time-to-event
> group sequential trial, not a seasonal rare-event exact-binomial example.
> Start with the original design specs: non-binding futility (`test.type = 4`),
> information-based spending, Hwang-Shih-DeCani efficacy spending with
> `sfupar = -4`, Hwang-Shih-DeCani futility spending with `sflpar = -2`,
> one-sided alpha 0.025, beta 0.1, a control median of 12 months, null HR 1,
> alternative HR 0.7, no dropout, uniform enrollment over 18 months, and
> minimum follow-up of 18 months. Keep those specs. Use equal randomization and
> add another interim analysis at 24 months after enrollment opens, so analyses
> occur at 12, 24, and 36 months. Show the R code and explain which gsDesign
> function matches this text.

Example response:

> The phrase "analyses at 12, 24, and 36 months after enrollment opens" fixes
> the analysis schedule on the calendar scale, so I would use
> `gsSurvCalendar(calendarTime = c(12, 24, 36))`. The timing request changes
> the analysis schedule only; the original test type, spending functions,
> alpha, beta, hazard assumptions, dropout, accrual, and follow-up inputs are
> preserved. I would not use `simBinomialSeasonalExact()` because the prompt
> does not describe a seasonal rare-event exact-binomial monitoring problem.
> If the resulting planned event counts are very low, such as fewer than 100
> events, I might mention exact-binomial rare-event methods as a sensitivity or
> alternative workflow to consider, but I would not switch to them unless the
> endpoint and monitoring plan are actually binomial rare-event based. Equal
> randomization is `ratio = 1`.

```r
library(gsDesign)

design <- gsSurvCalendar(
  calendarTime = c(12, 24, 36),
  spending = "information",
  test.type = 4,
  alpha = 0.025,
  beta = 0.10,
  sfu = sfHSD,
  sfupar = -4,
  sfl = sfHSD,
  sflpar = -2,
  lambdaC = log(2) / 12,
  hr = 0.70,
  hr0 = 1,
  eta = 0,
  gamma = 1,
  R = 18,
  minfup = 18,
  ratio = 1
)

design_integer <- toInteger(design)
gsBoundSummary(design_integer)
```

The response is useful because it maps the user's timing language to the
appropriate `gsDesign` function before writing code. It also distinguishes
calendar timing from calendar spending: `spending = "information"` keeps
spending tied to information, while `spending = "calendar"` would tie spending
to the calendar fractions. The original design specifications are kept in the
function call rather than being replaced by generic defaults. The low-count
heuristic is treated as a prompt to discuss alternatives, not as permission to
change the design family.

## What these instructions do not do

AI workflow instructions do not change the statistical behavior of `gsDesign`.
They also do not replace human review. They help make agent-assisted package
maintenance more reproducible by placing the expected repository workflow where
coding agents can find it.
