---
title: "Basic time-to-event group sequential design using gsSurv"
author: "Keaven M. Anderson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
references:
  - id: KimTsiatis
    title: Study duration for clinical trials with survival response and early stopping rule.
    author: 
      - family: Kim
        given: Kyungmann
      - family: Tsiatis
        given: Anastasios
    container-title: Kim and Tsiatis (1990)
    volume: 46
    URL: 'https://www.ncbi.nlm.nih.gov/pubmed/2350572'
    publisher: Biometrics
    page: 81-92
    type: article-journal
    issued:
      year: 1990
      month: 3
  - id: LachinFoulkes
    title: Evaluation of sample size and power for analyses of survival with allowance for nonuniform patient entry, losses to follow-up noncompliance, and stratification.
    author: 
      - family: Lachin
        given: John
      - family: Foulkes
        given: Mary
    container-title: Lachin and Foulkes(1986)
    volume: 42
    URL: 'https://www.jstor.org/stable/2531201'
    DOI: 10.2307/2531201
    publisher: Biometrics
    page: 507-519
    type: article-journal
    issued:
      year: 1986
      month: 9
vignette: >
  %\VignetteIndexEntry{Basic time-to-event group sequential design using gsSurv}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

options( width = 58 )

```

\section{Introduction}

This article/vignette provides a basic time-to-event endpoint designs for fixed designs using \texttt{nSurv} and group sequential designs using \texttt{gsSurv}.
Some detail in specification comes with the flexibility allowed by the @LachinFoulkes method for sample size with piecewise constant enrollment, piecewise exponential failure and dropout rates under a proportional hazards assumption.
Users may also be interested in the [Shiny interface](https://gsdesign.shinyapps.io/prod/) as a learning tool. 
We only use the simplest options here with a single stratum and exponential failure and dropout rates.
We demonstrate tabular, textual and graphical output.
While \texttt{nSurv} and \texttt{gsSurv} allow stratified populations and piecewise exponential failure rates, we restrict ourselves to a single stratum with an exponential failure distribution here; see the help file for \texttt{gsSurv} for examples with a stratified population or piecewise exponential failure.
We demonstrate this in a separate article/vignette for sample size using a cure model.

We apply the @LachinFoulkes sample size method and extend it to group sequential design.
This method fixes the duration of a study and varies enrollment rates to power a trial.
We also use the @LachinFoulkes basic power calculation to compute sample size along the lines of @KimTsiatis where enrollment rates are fixed and enrollment duration is allowed to vary to enroll a sufficient sample size to power a study.

\section{Parameter setup and design derivation}

\subsection{Enrollment and trial duration}

We begin by setting up trial duration and the enrollment pattern.
There are two basic methods for doing this.
The @LachinFoulkes method demonstrated here fixes the enrollment pattern and duration as well as the trial duration and changes the absolute enrollment rates to obtain desired power.
The alternate recommended method is along the lines of @KimTsiatis, fixing the enrollment rates and follow-up duration, varying the total trial duration to power the design; this will also be demonstrated below.

```{r,message=FALSE}
library(gsDesign)
# study duration
T <- 36 
# follow-up duration of last patient enrolled
minfup <- 12
# enrollment period durations
R <- c(1,2,3,4)
# relative enrollment rates during above periods
gamma<-c(1,1.5,2.5,4)
# randomization ratio, experimental/control
ratio <- 1
```


\subsection{Outcome and dropout distributions}

Next we provide information about the median time-to-event in the control group, dropout rate, hazard ratios under the null and alternate hypotheses for experimental therapy compared to control, and the desired Type I and II error rates.

```{r}
# median control time-to-event
median <- 12
# exponential dropout rate per unit of time
eta <- .001
# hypothesized experimental/control hazard ratio 
# (alternate hypothesis)
hr <- .75
# null hazard ratio (1 for superiority, >1 for non-inferiority)
hr0 <- 1
# Type I error (1-sided)
alpha <-.025
# Type II error (1-power)
beta<-.1
```

\subsection{Deriving a design with no interim analyses}

The above information is suffcient to design a trial with no interim analyses.
Note that when calling \texttt{nSurv}, we transform the median time-to-event ($m$) to an exponential event rate ($\lambda$) with the formula
$$\lambda=\log(2)/m.$$


```{r,tidy=FALSE}
x <- nSurv(R = R,
           gamma = gamma,
           eta = eta,
           minfup = minfup,
           T = T,
           lambdaC = log(2)/median,
           hr = hr,
           hr0 = hr0,
           beta = beta,
           alpha = alpha)
```

Next we store some text for use in output; only need to edit \texttt{timename}, \texttt{timename1} and \texttt{ep} as the following code is used to format enrollment rate input from above. See the following paragraph below to see where this text is used. Note that this code is not needed for Table \ref{tab2} produced using \texttt{xprint}.


```{r,echo=TRUE}
# time units
timename <- "months"
timename1 <- "month"
# endpoint name
ep <-"overall survival" 
# make a string with enrollment rates (assumes gamma is a single value or vector)
nR <- length(x$R)
if (nR==1){
  
  enrolrates <- paste("constant at a rate of ",round(gamma,1),"per",timename1,".")
  
} else{
  
enrolrates <- paste(c("piecewise constant at rates of ",
 paste(round(as.vector(x$gamma),1)," for ",timename," ",cumsum(c(0,x$R[1:(nR-1)])),
       "-",cumsum(x$R),collapse=", "),sep=""),collapse="")
}

```


The median time-to-event is assumed to be `r median` `r timename` in the control group.
`r if (hr0==1) paste("The trial is designed to demonstrate superiority of experimental treatment over control with an assumed hazard ratio of ",hr)` `r if (hr0 != 1) paste("The trial is designed to reject a null hypothesis hazard ratio of ",hr0," in favor of a hazard ratio of",hr)` with `r 100*(1-beta)`\% power and a one-sided Type I error rate of `r 100*alpha`.
The total sample size is `r 2*ceiling(x$n/2)` and a
total of `r ceiling(x$d)` endpoints is required for analysis.
Planned recruitment duration is `r sum(x$R)` `r timename` and the minimum follow-up planned is `r round(x$minfup,1)` `r timename`.
Thus, the total expected study duration is `r round(max(x$T),1)` `r timename`.
Enrollment is assumed to be `r enrolrates`.
The assumed dropout rate is `r 100*eta`\% per `r timename1`.

\section{Group sequential design}
Now we move on to a group sequential design.
We set up the number of analyses, timing and spending function parameters.

```{r}
# number of analyses (interim + final)
k <- 2
# timing of interim analyses (k-1 increasing numbers >0 and <1)
timing <- c(.4)
# efficacy bound spending function
sfu <- sfHSD 
# efficacy bound spending parameter specification
sfupar <- -10
# futility bound spending function
sfl <- sfHSD 
# futility bound spending parameter specification
sflpar <- 2
```


Type II error (1-power) may be set up differently than for a fixed design so that more meaningful futility analyses can be performed during the course of the trial.

```{r TypeII}
# Type II error=1-Power
beta <- .2
```


```{r,echo=TRUE,results="hide"}

# footnote text for table
footnote <- paste("P\\{Cross\\} is the probability of crossing the given bound (efficacy or futility) at or before the given analysis under the assumed hazard ratio (HR). Design assumes futility bound is discretionary (non-binding), but smaller upper boundary crossing probabilities shown here assume trial stops at first boundary crossing (binding bounds).")
# caption text for table
caption <- paste("Overall survival trial design with HR=",hr,", ",100*(1-beta),"\\% power and ",100*alpha,"\\% Type 1 error.",sep="")

```


Now we are prepared to generate the design.

```{r,tidy=FALSE}

# generate design
x <- gsSurv(k=k,timing=timing,R=R,gamma=gamma,eta=eta,
            minfup=minfup,T=T,lambdaC=log(2)/median,
            hr=hr,hr0=hr0,beta=beta,alpha=alpha,
            sfu=sfu,sfupar=sfupar,sfl=sfl,sflpar=sflpar)
# make a string with enrollment rates 
# (assumes gamma is a single value or vector)
nR <- length(x$R)
if (nR==1){enrolrates <- paste("constant at a rate of ",
                          round(gamma,1),"per",timename1,".")
} else{
enrolrates <- paste(c("piecewise constant at rates of ",
 paste(round(as.vector(x$gamma),1)," for ",timename," ",
       cumsum(c(0,x$R[1:(nR-1)])),
       "-",cumsum(x$R),collapse=", "),sep=""),collapse="")
}

```

For a comparative trial we consider a 2-arm group sequential design with `r ep` as the primary endpoint as shown in Table \ref{tab1}; an alternate of the table is demonstrated in \ref{tab2}.
Timing, number of events, sample size, boundaries (Z-values, nominal p-values, approximate hazard ratios) are shown as well as the probability of crossing study boundaries under the null and alternate hypotheses.
Bounds are determined by Hwang-Shih-DeCani spending functions with $\gamma=`r sfupar`$ ($\alpha$-spending) and $\gamma=`r sflpar`$ ($\beta$-spending).
The median time-to-event is assumed to be `r median` `r timename` in the control group.
`r if (hr0==1) paste("The trial is designed to demonstrate superiority of experimental treatment over control with an assumed hazard ratio of ",hr,".",sep="")`
The total sample size is `r ceiling(sum(x$eNE[k,]+x$eNC[k,]))` and a
total of `r ceiling(sum(x$eDE[k,]+x$eDC[k,]))` endpoints is required for the final analysis.
Planned recruitment duration is `r sum(x$R)` `r timename` and the minimum follow-up planned is `r round(x$minfup,1)` `r timename`.
Thus, the total expected study duration is `r round(max(x$T),1)` `r timename`.
Enrollment is assumed to be `r enrolrates`.
The assumed dropout rate is `r 100*eta`\% per `r timename1`.
`r if (x$k==2) paste("There is a single interim analysis planned after",ceiling(x$eDE[1,]+x$eDC[1,]),"events have accrued which is expected after approximately",round(x$T[1],1),timename,sep=" ")`.

```{r tab1, results="asis",echo=TRUE,message=FALSE,tidy=FALSE}

print(xtable(x,footnote=footnote,caption=caption,
             label="tab1"), 
      include.rownames=F, 
      sanitize.text.function=function(x) x)

```


```{r tab2, results="asis",tidy=FALSE,message=FALSE,tidy=FALSE}

xprint(xtable(gsBoundSummary(x, logdelta=TRUE), 
              digits=4, label="tab2", 
              caption=summary(x)))

```


Following are plots of the Z-values (Figure \ref{fig:p1}) and approximate hazard ratios (Figure \ref{fig:p2}) at the design bounds.

```{r p1, fig.height=3.5,tidy=FALSE,fig.cap="Z-value bound plot"}
plot(x,cex=.8,xlab="Number of events")
```

```{r p2, fig.height=3.5,fig.cap="Hazard ratio bound plot"}
plot(x,plottype="hr",cex=.8,xlab="Number of events")
```
