---
title: "Template instructions for time-to-event group sequential design"
author: "Keaven M. Anderson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
references:
  - id: KimTsiatis
    title: Study duration for clinical trials with survival response and early stopping rule.
    author: 
      - family: Kyungmann
        given: Kim
      - family: Tsiatis
        given: Anastasios
    container-title: Kim and Tsiatis (1990)
    volume: 46
    URL: 'https://www.ncbi.nlm.nih.gov/pubmed/2350572'
    publisher: Biometrics
    page: 81-92
    type: article-journal
    issued:
      year: 1990
      month: 3
  - id: LachinFoulkes
    title: Evaluation of sample size and power for analyses of survival with allowance for nonuniform patient entry, losses to follow-up noncompliance, and stratification.
    author: 
      - family: Lachin
        given: John
      - family: Foulkes
        given: Mary
    container-title: Lachin and Foulkes(1986)
    volume: 42
    URL: 'https://www.jstor.org/stable/2531201'
    DOI: 10.2307/2531201
    publisher: Biometrics
    page: 507-519
    type: article-journal
    issued:
      year: 1986
      month: 9
vignette: >
  %\VignetteIndexEntry{Template instructions for time-to-event group sequential design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

options( width = 58 )

```

\section{Introduction}

This vignette provides a instructions for a template for time-to-event sample size calculations for fixed designs using \texttt{nSurv} and group sequential designs using \texttt{gsSurv}.
It may be useful to edit the [template](gsSurvTemplate.html), so that you can reuse it on a regular basis for sample size calculations for time-to-event study planning.
The template only uses the simplest options with a single stratum and exponential failure and dropout rates.
The template can be modified to accomodate multiple strata and/or piecewise exponential failure and dropout rates; this was not chosen here since the simplest options are a) often used and b) simplest to learn and apply for beginners. 
Note that we produce tabular, textual and graphical output; examining the source file to see how this is done will enable you to easily customize to fit your purposes.
You will need the knitr R package, which I find simpler to use than Sweave (although you could make minor edits and use Sweave).
I have found using knitr and the RStudio development environment to be a good combination.
Within this instruction vignette for the template, we generally show all of the code used to generate results. You will want to suppress this in your own template as demonstrated in the [template](gsSurvTemplate.html).

We apply the @LachinFoulkes sample size method and extend it to group sequential design.
This method fixes the duration of a study and varies enrollment rates to power a trial.
We also use the @LachinFoulkes basic power equation to compute sample size along the lines of @KimTsiatis where enrollment rates are fixed and enrollment duration is allowed to vary to enroll a sufficient sample size to power a study.
While \texttt{nSurv} and \texttt{gsSurv} allow stratified populations and piecewise exponential failure rates, we restrict ourselves to a single stratum with an exponential failure distribution here; see the help file for \texttt{gsSurv} for examples with a stratified population or piecewise exponential failure.

Some detail in specification comes with the flexibility allowed by the @LachinFoulkes method.
Thus a template is helpful to simplify use.

We present a relatively simple example of time-to-event sample size calculation for a group sequential design.

\subsection{Basic assumptions and fixed design sample size}
Following is code using the gsDesign R package. 


The first chunk of code sets up enrollment and dropout information.

```{r enrollment,message=FALSE}
library(gsDesign)

# enrollment period durations; the last of these will be extended if T=NULL below
R <- c(1,2,3,4)
# relative enrollment rates during above periods
gamma<-c(1,1.5,2.5,4)
# study duration
# T can be set to NULL if you want to 
# fix enrollment and vary study duration
T <- 36 
# follow-up duration of last patient enrolled
minfup <- 12
# randomization ratio, experimental/control
ratio <- 1
```

Next we provide information about the median time to event in the control group, dropout rate, hazard ratios under the null and alternate hypotheses for experimental therapy compared to control, and the desired Type I and II error rates.

```{r ep}

# median control time-to-event
median <- 12
# exponential dropout rate per unit of time
eta <- .001
# hypothesized experimental/control hazard ratio
hr <- .75
# null hazard ratio (1 for superiority, >1 for non-inferiority)
hr0 <- 1
# Type I error (1-sided)
alpha <-.025
# Type II error (1-power)
beta<-.1

```

Finally, we design a trial with no interim analyses under these assumptions.
Note that when calling \texttt{nSurv}, we transform the median time-to-event ($m$) to an exponential event rate ($\lambda$) with the formula
$$\lambda=\log(2)/m.$$


```{r nsurv,tidy=FALSE}

x <- nSurv(R = R,
           gamma = gamma,
           eta = eta,
           minfup = minfup,
           T = T,
           lambdaC = log(2)/median,
           hr = hr,
           hr0 = hr0,
           beta = beta,
           alpha = alpha)

```

Next we store some text for use in output; only need to edit \texttt{timename}, \texttt{timename1} and \texttt{ep} as the following code is used to format enrollment rate input from above. See the following paragraph below to see where this text is used. Note that this code is not needed for Table \ref{tab2} produced using \texttt{xprint}.


```{r textstuff,echo=TRUE,results="hide"}
# time units
timename <- "months"
timename1 <- "month"
# endpoint name
ep <-"overall survival" 
# make a string with enrollment rates (assumes gamma is a single value or vector)
nR <- length(x$R)
if (nR==1){
  
  enrolrates <- paste("constant at a rate of ",round(gamma,1),"per",timename1,".")
  
} else{
  
enrolrates <- paste(c("piecewise constant at rates of ",
 paste(round(as.vector(x$gamma),1)," for ",timename," ",cumsum(c(0,x$R[1:(nR-1)])),
       "-",cumsum(x$R),collapse=", "),sep=""),collapse="")
}

```


The median time-to-event is assumed to be `r median` `r timename` in the control group.
`r if (hr0==1) paste("The trial is designed to demonstrate superiority of experimental treatment over control with an assumed hazard ratio of ",hr)` `r if (hr0 != 1) paste("The trial is designed to reject a null hypothesis hazard ratio of ",hr0," in favor of a hazard ratio of",hr)` with `r 100*(1-beta)`\% power and a one-sided Type I error rate of `r 100*alpha`.
The total sample size is `r 2*ceiling(x$n/2)` and a
total of `r ceiling(x$d)` endpoints is required for analysis.
Planned recruitment duration is `r sum(x$R)` `r timename` and the minimum follow-up planned is `r round(x$minfup,1)` `r timename`.
Thus, the total expected study duration is `r round(max(x$T),1)` `r timename`.
Enrollment is assumed to be `r enrolrates`.
The assumed dropout rate is `r 100*eta`\% per `r timename1`.

\section{Group sequential design}
Now we move on to a group sequential design.
We set up the number of analyses, timing and spending function parameters.

```{r sf}

# number of analyses (interim + final)
k <- 2
# timing of interim analyses (k-1 increasing numbers >0 and <1)
timing <- c(.4)
# efficacy bound spending function
sfu <- sfHSD 
# efficacy bound spending parameter specification
sfupar <- -10
# futility bound spending function
sfl <- sfHSD 
# futility bound spending parameter specification
sflpar <- 2

```


Type II error (1-power) may be set up differently than for a fixed design so that more meaningful futility analyses can be performed during the course of the trial.

```{r TypeII}

# Type II error=1-Power
beta <- .2

```


```{r captext,echo=FALSE,results="hide"}

# footnote text for table
footnote <- paste("P\\{Cross\\} is the probability of crossing the given bound (efficacy or futility) at or before the given analysis under the assumed hazard ratio (HR). Design assumes futility bound is discretionary (non-binding), but smaller upper boundary crossing probabilities shown here assume trial stops at first boundary crossing (binding bounds).")
# caption text for table
caption <- paste("Overall survival trial design with HR=",hr,", ",100*(1-beta),"\\% power and ",100*alpha,"\\% Type 1 error.",sep="")

```


Now we are prepared to generate the design.

```{r gsdesign,tidy=FALSE}

# generate design
x <- gsSurv(k=k,timing=timing,R=R,gamma=gamma,eta=eta,
            minfup=minfup,T=T,lambdaC=log(2)/median,
            hr=hr,hr0=hr0,beta=beta,alpha=alpha,
            sfu=sfu,sfupar=sfupar,sfl=sfl,sflpar=sflpar)
# make a string with enrollment rates 
# (assumes gamma is a single value or vector)
nR <- length(x$R)
if (nR==1){enrolrates <- paste("constant at a rate of ",
                          round(gamma,1),"per",timename1,".")
} else{
enrolrates <- paste(c("piecewise constant at rates of ",
 paste(round(as.vector(x$gamma),1)," for ",timename," ",
       cumsum(c(0,x$R[1:(nR-1)])),
       "-",cumsum(x$R),collapse=", "),sep=""),collapse="")
}

```

For a comparative trial we consider a 2-arm group sequential design with `r ep` as the primary endpoint as shown in Table \ref{tab1}; an alternate of the table is demonstrated in \ref{tab2}.
Timing, number of events, sample size, boundaries (Z-values, nominal p-values, approximate hazard ratios) are shown as well as the probability of crossing study boundaries under the null and alternate hypotheses.
Bounds are determined by Hwang-Shih-DeCani spending functions with $\gamma=`r sfupar`$ ($\alpha$-spending) and $\gamma=`r sflpar`$ ($\beta$-spending).
The median time-to-event is assumed to be `r median` `r timename` in the control group.
`r if (hr0==1) paste("The trial is designed to demonstrate superiority of experimental treatment over control with an assumed hazard ratio of ",hr,".",sep="")`
The total sample size is `r ceiling(sum(x$eNE[k,]+x$eNC[k,]))` and a
total of `r ceiling(sum(x$eDE[k,]+x$eDC[k,]))` endpoints is required for the final analysis.
Planned recruitment duration is `r sum(x$R)` `r timename` and the minimum follow-up planned is `r round(x$minfup,1)` `r timename`.
Thus, the total expected study duration is `r round(max(x$T),1)` `r timename`.
Enrollment is assumed to be `r enrolrates`.
The assumed dropout rate is `r 100*eta`\% per `r timename1`.
`r if (x$k==2) paste("There is a single interim analysis planned after",ceiling(x$eDE[1,]+x$eDC[1,]),"events have accrued which is expected after approximately",round(x$T[1],1),timename,sep=" ")`.

```{r tab1, results="asis",echo=TRUE,message=FALSE,tidy=FALSE}

print(xtable(x,footnote=footnote,caption=caption,
             label="tab1"), 
      include.rownames=F, 
      sanitize.text.function=function(x) x)

```


```{r tab2, results="asis",tidy=FALSE,message=FALSE,tidy=FALSE}

xprint(xtable(gsBoundSummary(x, logdelta=TRUE), 
              digits=4, label="tab2", 
              caption=summary(x)))

```


Following are plots of the Z-values (Figure \ref{fig:p1}) and approximate hazard ratios (Figure \ref{fig:p2}) at the design bounds.

```{r p1, fig.height=3.5,tidy=FALSE,fig.cap="Z-value bound plot"}

plot(x,cex=.8,xlab="Number of events")

```

```{r p2, fig.height=3.5,fig.cap="Hazard ratio bound plot"}

plot(x,plottype="hr",cex=.8,xlab="Number of events")

```
