---
title: "Template for time-to-event group sequential design"
author: "Keaven M. Anderson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
references:
  - id: KimTsiatis
    title: Study duration for clinical trials with survival response and early stopping rule.
    author: 
      - family: Kyungmann
        given: Kim
      - family: Tsiatis
        given: Anastasios
    container-title: Kim and Tsiatis (1990)
    volume: 46
    URL: 'https://www.ncbi.nlm.nih.gov/pubmed/2350572'
    publisher: Biometrics
    page: 81-92
    type: article-journal
    issued:
      year: 1990
      month: 3
  - id: LachinFoulkes
    title: Evaluation of sample size and power for analyses of survival with allowance for nonuniform patient entry, losses to follow-up noncompliance, and stratification.
    author: 
      - family: Lachin
        given: John
      - family: Foulkes
        given: Mary
    container-title: Lachin and Foulkes(1986)
    volume: 42
    URL: 'https://www.jstor.org/stable/2531201'
    DOI: 10.2307/2531201
    publisher: Biometrics
    page: 507-519
    type: article-journal
    issued:
      year: 1986
      month: 9

vignette: >
  %\VignetteIndexEntry{Template for time-to-event group sequential design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This vignette provides a template for time-to-event sample size calculations for fixed designs using \texttt{nSurv} and group sequential designs using \texttt{gsSurv}.
Edit this template (from gsDesign R package source /inst/doc/gsSurvTemplate.rnw) so that you can reuse it on a regular basis for sample size calculations for time-to-event study planning.
The template only uses the simplest options with a single stratum and exponential failure and dropout rates.
The template can be modified to accomodate multiple strata and/or piecewise exponential failure and dropout rates; this was not chosen here since the simplest options are a) often used and b) simplest to learn and apply for beginners. 
Note that we produce tabular, textual and graphical output; examining the source file to see how this is done will enable you to easily customize to fit your purposes.
You will need the knitr R package, which I find simpler to use than Sweave (although you could make minor edits and use Sweave).
I have found using knitr and the RStudio development environment to be a good combination.
Within this template, we suppress printing of all of the code used to generate results. 
The file /inst/doc/gsSurvTemplateInstructions.pdf is an alternate version of this document that shows code associated with this template.

We apply the @LachinFoulkes sample size method and extend it to group sequential design.
This method fixes the duration of a study and varies enrollment rates to power a trial.
{Alternate text when enrollment rate is fixed and enrollment duration is allowed to vary: \em We use the @LachinFoulkes basic power equation to compute sample size along the lines of @KimTsiatis where enrollment rates are fixed and enrollment duration is allowed to vary to enroll a sufficient sample size to power a study.}
The sample size method assumes proportional hazards between treatment groups.
Accrual of events of events over time is approximated assuming failure rates are exponential.

```{r enrollment, message=FALSE, echo=FALSE, results="hide"}
# code here applies to both fixed and group sequential design sections below.
library(gsDesign)
library(xtable)
# enrollment period durations; the last of these will be extended if T=NULL below
R <- c(1,2,3,4)
# relative enrollment rates during above periods
gamma<-c(1,1.5,2.5,4)
# study duration
# T can be set to NULL if you want to 
# fix enrollment and vary study duration
T <- 36 
# follow-up duration of last patient enrolled
minfup <- 12
# randomization ratio, experimental/control
ratio <- 1
# median control time-to-event
median <- 12
# exponential dropout rate per unit of time
eta <- .001
# hypothesized experimental/control hazard ratio
hr <- .75
# null hazard ratio (1 for superiority, >1 for non-inferiority)
hr0 <- 1
# Type I error (1-sided)
alpha <-.025
# Type II error (1-power)
beta<-.2
x <- nSurv(R=R,gamma=gamma,eta=eta,minfup=minfup,
           T=T,lambdaC=log(2)/median,
           hr=hr,hr0=hr0,beta=beta,alpha=alpha)
# time units
timename <- "months"
timename1 <- "month"
# endpoint name
ep <-"overall survival" 
# make a string with enrollment rates (assumes gamma is a single value or vector)
nR <- length(x$R)
if (nR==1){enrolrates <- paste("constant at a rate of ",round(gamma,1),"per",timename1,".")
} else{
enrolrates <- paste(c("piecewise constant at rates of ",
 paste(round(as.vector(x$gamma),1)," for ",timename," ",cumsum(c(0,x$R[1:(nR-1)])),
       "-",cumsum(x$R),collapse=", "),sep=""),collapse="")
}
```

## Fixed design sample size

__This section can be deleted if you are only interested in group sequential design.__

The median time-to-event is assumed to be \Sexpr{median} \Sexpr{timename} in the control group.

\Sexpr{if (hr0==1) paste("The trial is designed to demonstrate superiority of experimental treatment over control with an assumed hazard ratio of ",hr)} \Sexpr{if (hr0 != 1) paste("The trial is designed to reject a null hypothesis hazard ratio of ",hr0," in favor of a hazard ratio of",hr)} with \Sexpr{100*(1-beta)}\% power and a one-sided Type I error rate of \Sexpr{100*alpha}.
The total sample size is \Sexpr{2*ceiling(x$n/2)} and a
total of \Sexpr{ceiling(x$d)} endpoints is required for analysis.
Planned recruitment duration is \Sexpr{sum(x$R)} \Sexpr{timename} and the minimum follow-up planned is \Sexpr{round(x$minfup,1)} \Sexpr{timename}.
Thus, the total expected study duration is \Sexpr{round(max(x$T),1)} \Sexpr{timename}.
Enrollment is assumed to be \Sexpr{enrolrates}.
The assumed dropout rate is \Sexpr{100*eta}\% per \Sexpr{timename1}.

# Group sequential design

```{r sf,echo=FALSE,results="hide"}
# number of analyses (interim + final)
k <- 2
# timing of interim analyses (k-1 increasing numbers >0 and <1)
timing <- c(.4)
# efficacy bound spending function
sfu <- sfHSD 
# efficacy bound spending parameter specification
sfupar <- -10
# futility bound spending function
sfl <- sfHSD 
# futility bound spending parameter specification
sflpar <- 2
# footnote text for table
footnote <- paste("P\\{Cross\\} is the probability of crossing the given bound (efficacy or futility) at or before the given analysis under the assumed hazard ratio (HR). Design assumes futility bound is discretionary (non-binding), but smaller upper boundary crossing probabilities shown here assume trial stops at first boundary crossing (binding bounds).")
# caption text for table
caption <- paste("Overall survival trial design with HR=",hr,", ",100*(1-beta),"\\% power and ",100*alpha,"\\% Type 1 error.",sep="")
# generate design
x <- gsSurv(k=k,timing=timing,R=R,gamma=gamma,eta=eta,
            minfup=minfup,T=T,lambdaC=log(2)/median,
            hr=hr,hr0=hr0,beta=beta,alpha=alpha,
            sfu=sfu,sfupar=sfupar,sfl=sfl,sflpar=sflpar)
# make a string with enrollment rates (assumes gamma is a single value or vector)
nR <- length(x$R)
if (nR==1){enrolrates <- paste("constant at a rate of ",round(gamma,1),"per",timename1,".")
} else{
enrolrates <- paste(c("piecewise constant at rates of ",
 paste(round(as.vector(x$gamma),1)," for ",timename," ",cumsum(c(0,x$R[1:(nR-1)])),
       "-",cumsum(x$R),collapse=", "),sep=""),collapse="")
}
```

For a comparative trial we consider a 2-arm group sequential design with \Sexpr{ep} as the primary endpoint as shown in Table \ref{tab1} (a second version of the table is in \ref{tab2}.
Timing, number of events, sample size, boundaries (Z-values, nominal p-values, approximate hazard ratios) are shown as well as the probability of crossing study boundaries under the null and alternate hypotheses.
Bounds are determined by Hwang-Shih-DeCani spending functions with $\gamma=\Sexpr{sfupar}$ ($\alpha$-spending) and $\gamma=\Sexpr{sflpar}$ ($\beta$-spending).
The median time-to-event is assumed to be \Sexpr{median} \Sexpr{timename} in the control group.
\Sexpr{if (hr0==1) paste("The trial is designed to demonstrate superiority of experimental treatment over control with an assumed hazard ratio of ",hr,".",sep="")}
The total sample size is \Sexpr{ceiling(sum(x$eNE[k,]+x$eNC[k,]))} and a
total of \Sexpr{ceiling(sum(x$eDE[k,]+x$eDC[k,]))} endpoints is required for the final analysis.
Planned recruitment duration is \Sexpr{sum(x$R)} \Sexpr{timename} and the minimum follow-up planned is \Sexpr{round(x$minfup,1)} \Sexpr{timename}.
Thus, the total expected study duration is \Sexpr{round(max(x$T),1)} \Sexpr{timename}.
Enrollment is assumed to be \Sexpr{enrolrates}.
The assumed dropout rate is \Sexpr{100*eta}\% per \Sexpr{timename1}.
\Sexpr{if (x$k==2) paste("There is a single interim analysis planned after",ceiling(x$eDE[1,]+x$eDC[1,]),"events have accrued which is expected after approximately",round(x$T[1],1),timename,sep=" ")}.

```{r tab1, results="asis",echo=FALSE,message=FALSE,tidy=FALSE}
print(xtable(x,footnote=footnote,caption=caption,
             label="tab1"), 
      include.rownames=F, 
      sanitize.text.function=function(x) x)
```

```{r tab2, results="asis",echo=FALSE,message=FALSE,tidy=FALSE}
xprint(xtable(gsBoundSummary(x, logdelta=TRUE), digits=4, label="tab2", caption=summary(x)))
```

Following are plots of the Z-values (Figure \ref{fig:p1}) and approximate hazard ratios (Figure \ref{fig:p2}) at the design bounds.

```{r p1, fig.height=3.5,tidy=FALSE,fig.cap="Z-value bound plot",echo=FALSE}
plot(x,cex=.8,xlab="Number of events")
```

```{r p2, fig.height=3.5,fig.cap="Hazard ratio bound plot",echo=FALSE}
plot(x,plottype="hr",cex=.8,xlab="Number of events")
```

# References

