---
title: "Power Computation for Group Sequential Survival Designs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Power Computation for Group Sequential Survival Designs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
library(gsDesign)
```

## Motivation

`gsSurv()` and `gsSurvCalendar()` derive sample sizes and enrollment rates
to achieve a target power for a given hazard ratio. In practice, we often
want to answer the reverse question: *given a specified hazard ratio, fixed enrollment, dropout, and
analysis timing assumptions, what power does the design achieve?*

Common scenarios include:

- **Sensitivity analysis**: What happens to power if the true hazard ratio
  is 0.75 instead of the design assumption of 0.65?
- **Changing alpha**: What if the multiplicity scheme allows one-sided $\alpha = 0.025$
  instead of $\alpha = 0.0125$?
- **Modified enrollment**: What if enrollment is slower than planned?
- **Different analysis times**: What if interim analyses occur at calendar
  times that differ from the original design?

`gsSurvPower()` addresses these questions by computing power for a group
sequential survival design under user-specified assumptions.

## Design

`gsSurvPower()` accepts an optional `gsSurv` object `x` that provides defaults
for all parameters. Any parameter the user explicitly specifies overrides the
corresponding value from `x`. When `x` is not provided, all design parameters
must be specified directly.

### Hazard ratio roles

Two distinct hazard ratios play different roles:

- **`hr`**: The assumed hazard ratio under which power is computed. This is the
  "what-if" treatment effect.
- **`hr1`**: The design hazard ratio used to calibrate futility bounds
  (`test.type` 3, 4, 7, 8 only; not used for `test.type` 5, 6 or harm bounds).
  When `x` is provided, `hr1` defaults to `x$hr` (the effect size the trial was
  originally designed for). Futility bounds remain calibrated to the design
  assumption even when power is evaluated under a different `hr`.

### Analysis timing: calendar time vs. event-driven

Analysis times can be specified by calendar time, by target event counts, or
by a combination of criteria. The choice has an important consequence for
sensitivity analyses:

- **`plannedCalendarTime`** fixes the calendar time of each analysis.
  Expected events are then recomputed under the assumed HR.
  A worse HR (closer to 1) produces *more* expected events at the same
  calendar time because the experimental arm fails faster.
  This gives an "unconditional" power that reflects how the assumed
  treatment effect influences event accrual.

- **`targetEvents`** fixes the event count at each analysis. The calendar
  time is solved for under the assumed HR. Since the event counts are held
  constant, the information fractions do not change with HR, and the
  resulting power matches the `gsDesign` power plot (`plot(x, plottype = 2)`)
  to numerical precision.

Both modes are useful. Calendar-time analyses are natural when the protocol
specifies analysis dates; event-driven analyses are natural when the protocol
specifies event targets.

Additional criteria can be combined per-analysis, each specified as a scalar
(recycled to all `k` analyses) or a vector of length `k` with `NA` for
"not applicable":

- `maxExtension`: Maximum time beyond the floor to wait for target events.
- `minTimeFromPreviousAnalysis`: Minimum elapsed time since the previous analysis.
- `minN`: Minimum sample size enrolled before analysis.
- `minFollowUp`: Minimum follow-up after `minN` is reached.

When multiple criteria apply to a single analysis, the analysis time is the
maximum of all floor criteria, with `targetEvents` potentially extending
beyond the floor. `maxExtension` acts as a hard cap: the analysis time never
exceeds `plannedCalendarTime + maxExtension` (or `T[i-1] + maxExtension`
when no calendar time is specified), even if other criteria such as
`minTimeFromPreviousAnalysis` or `minN + minFollowUp` would push it later.

### Spending and method

- **`spending`**: One of `"information"` (default) or `"calendar"`.
  Information-based spending tracks the fraction of statistical information
  accumulated; calendar-based spending sets `usTime = lsTime = T / max(T)`.
  Custom spending times can also be passed via `usTime` and `lsTime`.

- **`method`**: One of `"LachinFoulkes"` (default), `"Schoenfeld"`,
  `"Freedman"`, or `"BernsteinLagakos"`. Controls how fixed-design events
  (`n.fix`) are computed when `x` is not provided. When `x` is provided,
  `x$n.fix` is used directly for exact consistency with the design.

### Stratified targetEvents

`targetEvents` accepts a scalar (recycled), a vector of length `k`
(one overall target per analysis), or a matrix with `k` rows and
`nstrata` columns (per-stratum targets). A vector of length `k` is
always interpreted as overall targets; to specify per-stratum targets
for a single analysis, use a 1-row matrix.

## Example: design and power computation

We first create a group sequential design, then evaluate power under the
design assumptions and under an alternative hazard ratio.

```{r gssurvpower_design}
design <- gsSurv(
  k = 3, test.type = 4, alpha = 0.025, sided = 1, beta = 0.1,
  sfu = sfHSD, sfupar = -4, sfl = sfHSD, sflpar = -2,
  lambdaC = log(2) / 12, hr = 0.7, hr0 = 1,
  eta = 0.01, gamma = 10, R = 16, minfup = 12, T = 28
)
design
```

### Power at design assumptions

Passing the design's analysis times with the same hazard ratio exactly
reproduces the design power (90%). Internally, `gsSurvPower()` uses `x$n.fix`
(when `x` is provided) so that the drift parameter $\theta$ and bounds are
normalized identically to `gsSurv()`. The assumed HR's drift is obtained
by method-specific scaling:

- For Schoenfeld, LachinFoulkes, and BernsteinLagakos:
  $\theta_{\text{assumed}} = \theta_{\text{design}} \times |\log(\text{hr}/\text{hr}_0)| / |\log(\text{hr}_1/\text{hr}_0)|$
- For Freedman:
  $\theta_{\text{assumed}} = \theta_{\text{design}} \times |\delta(\text{hr})| / |\delta(\text{hr}_1)|$
  where $\delta(\text{hr}) = (\text{hr} - 1) / (\text{hr} + 1/\text{ratio})$

This reproduces the design power exactly at `hr = hr1` and scales
correctly for other hazard ratios, matching `rpact::getPowerSurvival()`
to within 0.5% across methods.

```{r gssurvpower_at_design}
pwr_design <- gsSurvPower(x = design, plannedCalendarTime = design$T)
cat("Design power:", round((1 - design$beta) * 100, 1), "%\n")
cat("gsSurvPower:  ", round(pwr_design$power * 100, 1), "%\n")
```

### Power under a different hazard ratio

Suppose the true treatment effect is HR = 0.8 instead of the design
assumption of 0.7:

```{r gssurvpower_worse_hr}
pwr_worse <- gsSurvPower(x = design, hr = 0.8, plannedCalendarTime = design$T)
cat("Power at HR = 0.8:", round(pwr_worse$power * 100, 1), "%\n")
```

The futility bounds remain calibrated to the design HR (0.7), but power
is evaluated under the assumed HR (0.8).

### Power over a range of hazard ratios

```{r gssurvpower_sensitivity}
hr_grid <- seq(0.55, 0.90, by = 0.05)
power_vals <- sapply(hr_grid, function(h) {
  p <- gsSurvPower(x = design, hr = h, plannedCalendarTime = design$T)
  p$power
})
results <- data.frame(HR = hr_grid, Power = round(power_vals * 100, 1))
results
```

## Comparison with gsDesign power plots

The `gsDesign` package provides power plots via `plot(design, plottype = 2)`.
These hold event counts fixed at the design values and vary only the drift
parameter $\theta$. The table below compares three approaches across a range
of hazard ratios:

- **gsDesign**: `gsProbability()` with design bounds, design events,
  scaled $\theta$. This is what `plot(design, plottype = 2)` computes.
- **gsSurvPower (fixed events)**: `gsSurvPower()` with
  `targetEvents = design_events`. Events are held constant; calendar
  times adjust.
- **gsSurvPower (fixed calendar)**: `gsSurvPower()` with
  `plannedCalendarTime = design$T`. Calendar times are held constant;
  events change with the assumed HR.

```{r gssurvpower_vs_gsdesign}
design_events <- design$n.I
hr_grid <- seq(0.55, 0.95, by = 0.05)

comparison <- data.frame(
  HR = hr_grid,
  gsDesign_plot = sapply(hr_grid, function(h) {
    delta_ratio <- abs(log(h)) / abs(log(design$hr))
    theta_h <- design$delta * delta_ratio
    gsp <- gsDesign::gsProbability(
      k = design$k, theta = theta_h,
      n.I = design$n.I,
      a = design$lower$bound, b = design$upper$bound, r = 18)
    sum(gsp$upper$prob)
  }),
  fixed_events = sapply(hr_grid, function(h) {
    gsSurvPower(x = design, hr = h, targetEvents = design_events)$power
  }),
  fixed_calendar = sapply(hr_grid, function(h) {
    gsSurvPower(x = design, hr = h, plannedCalendarTime = design$T)$power
  })
)
comparison[, -1] <- round(comparison[, -1] * 100, 2)
comparison
```

**Key observations:**

- The `gsDesign_plot` and `fixed_events` columns match to numerical
  precision because both condition on the same event counts at each
  analysis. When using `targetEvents`, `gsSurvPower()` reproduces the
  `gsDesign` power plot exactly.

- The `fixed_calendar` column differs modestly because fixing calendar
  times allows the expected event count to change with the assumed HR.
  A worse HR (closer to 1) produces more expected events at the same
  calendar time, since the experimental arm has a higher failure rate.
  This slightly changes the statistical information at each analysis,
  producing an "unconditional" power that accounts for the interplay
  between treatment effect and event accrual.

### Bounds stability

When using `targetEvents`, the efficacy and futility bounds do not change
with the assumed HR. The bounds are determined entirely by the design
parameters (alpha/beta spending, information fractions, `n.fix`) and are
reused directly from the input design `x` when the timing matches:

```{r gssurvpower_bounds}
design_events <- design$n.I
cat("Design bounds (Z-scale):\n")
cat("  Efficacy:", round(design$upper$bound, 4), "\n")
cat("  Futility:", round(design$lower$bound, 4), "\n\n")

for (h in c(0.5, 0.7, 0.8, 1.0)) {
  pwr <- gsSurvPower(x = design, hr = h, targetEvents = design_events)
  cat(sprintf("HR=%.1f  Efficacy: %s  Futility: %s  (identical: %s)\n",
    h,
    paste(round(pwr$upper$bound, 4), collapse = ", "),
    paste(round(pwr$lower$bound, 4), collapse = ", "),
    identical(pwr$upper$bound, design$upper$bound) &&
      identical(pwr$lower$bound, design$lower$bound)))
}
```

With `plannedCalendarTime`, different assumed HRs produce different expected
event counts and therefore different information fractions, so the bounds
are appropriately recomputed via `gsDesign::gsDesign()`.

## Changing alpha

A common use case is evaluating power at a different one-sided alpha level —
for example, when a graphical multiplicity procedure allocates
$\alpha = 0.05$ to one hypothesis instead of the design's $\alpha = 0.025$.

When `x` is provided and the information fractions (timing) match the
original design, `gsSurvPower()` recalculates **efficacy bounds** at the
new alpha using `gsDesign(test.type = 1)` (efficacy-only) while
**preserving the original futility bounds** from `x`. This follows the
same convention as `gsBoundSummary()`. Any futility bound that would
exceed the new (less stringent) efficacy bound is clipped.

When timing changes (e.g., different `targetEvents`), both bounds are
recomputed from scratch using the full `test.type` and spending functions.

```{r alpha_change}
# Original design: one-sided alpha = 0.025
cat("=== Original design (alpha = 0.025) ===\n")
cat("Efficacy bounds:", round(design$upper$bound, 4), "\n")
cat("Futility bounds:", round(design$lower$bound, 4), "\n\n")

# Power at alpha = 0.05 with same event counts (timing preserved)
design_events <- design$n.I
pwr_a05 <- gsSurvPower(x = design, alpha = 0.05, targetEvents = design_events)

cat("=== gsSurvPower at alpha = 0.05 ===\n")
cat("Efficacy bounds:", round(pwr_a05$upper$bound, 4), "\n")
cat("Futility bounds:", round(pwr_a05$lower$bound, 4), "\n")
cat("Power:          ", round(pwr_a05$power * 100, 1), "%\n\n")

# Cross-check: gsBoundSummary at the same alternate alpha
# (only test.type 1, 4, 6, 7, 8 are supported)
cat("=== gsBoundSummary (alpha = 0.05) ===\n")
print(gsBoundSummary(design, alpha = 0.05))
```

Note that `gsBoundSummary()` adds an $\alpha = 0.05$ column for
`test.type` 1, 4, 6, 7, and 8. For binding types (3, 5),
`gsBoundSummary()` does not support alternate alpha, but
`gsSurvPower()` handles them using the same approach: recompute efficacy
with `test.type = 1` at the new alpha and keep original futility bounds.

### Binding type example (test.type = 3)

```{r alpha_binding}
design3 <- gsSurv(
  k = 3, test.type = 3, alpha = 0.025, sided = 1, beta = 0.1,
  sfu = sfHSD, sfupar = -4, sfl = sfHSD, sflpar = -2,
  lambdaC = log(2) / 12, hr = 0.7, eta = 0.01,
  gamma = 10, R = 16, minfup = 12, T = 28
)
events3 <- design3$n.I

pwr3_a05 <- gsSurvPower(x = design3, alpha = 0.05, targetEvents = events3)

cat("=== Binding futility (test.type=3) at alpha = 0.05 ===\n")
cat("Original efficacy:", round(design3$upper$bound, 4), "\n")
cat("New efficacy:     ", round(pwr3_a05$upper$bound, 4), "\n")
cat("Original futility:", round(design3$lower$bound, 4), "\n")
cat("New futility:     ", round(pwr3_a05$lower$bound, 4), "\n")
cat("Power:            ", round(pwr3_a05$power * 100, 1), "%\n")
```

The futility bounds are preserved from the original design. At the final
analysis where the original futility bound equals the original efficacy
bound, the new efficacy bound (which is lower) becomes the clip point.

The output retains the original `test.type` and records the new `alpha`:

```{r alpha_output_fields}
cat("test.type:", pwr3_a05$test.type, "(same as input design)\n")
cat("alpha:    ", pwr3_a05$alpha, "(updated to new value)\n")
```

## Example: event-based timing

Instead of calendar times, analyses can be triggered by target event counts:

```{r gssurvpower_events}
pwr_events <- gsSurvPower(
  x = design,
  targetEvents = c(75, 150, 225)
)
cat("Analysis times:", round(pwr_events$T, 1), "\n")
cat("Events at each analysis:",
    round(pwr_events$n.I, 1), "\n")
cat("Power:", round(pwr_events$power * 100, 1), "%\n")
```

## Example: biomarker subgroup to stratified design

A common scenario is designing a trial for a biomarker-defined subgroup,
then assessing what power the same enrollment provides for the overall
(stratified) population under a more conservative treatment effect.
Note that the `gsDesign2` package could be used to design with different
hazard ratios in the biomarker-positive and biomarker-negative populations
simultaneously; here we illustrate the simpler approach using `gsSurvPower()`.

### Step 1: Design for the biomarker-positive subgroup

Suppose 60% of the population is biomarker-positive (prevalence = 0.6),
the control median survival in this subgroup is 12 months,
the hazard ratio is 0.65, and we target 90% power at one-sided
$\alpha = 0.0125$ (e.g., from a graphical multiplicity allocation):

```{r biomarker_design}
prevalence <- 0.6
median_bm_pos <- 12    # control median in biomarker+ (months)
median_bm_neg <- 10    # control median in biomarker- (shorter prognosis)

bm_design <- gsSurvCalendar(
  test.type = 4, alpha = 0.0125, beta = 0.1,
  sfu = sfHSD, sfupar = -4, sfl = sfHSD, sflpar = -2,
  calendarTime = c(12, 24, 36),
  lambdaC = log(2) / median_bm_pos, hr = 0.65, eta = 0.01,
  gamma = 10, R = 18, minfup = 18, ratio = 1
)

summary(bm_design)
gsBoundSummary(bm_design)
```

### Step 2: Power for the overall (stratified) population

Now consider enrolling the entire population using the same enrollment
duration and analysis calendar times. The biomarker-positive enrollment
rate comes from the subgroup design. The biomarker-negative enrollment
rate is proportionate based on prevalence: if biomarker-positive patients
enroll at rate $\gamma_{+}$, biomarker-negative patients enroll at rate
$\gamma_{-} = \gamma_{+} \times (1 - p) / p$.

We use a **stratified** approach: `lambdaC` and `gamma` are specified as
matrices with two columns (one per stratum), allowing different control
hazard rates and enrollment rates in each biomarker subgroup.
The overall hazard ratio assumed is 0.75 (attenuated because the
biomarker-negative subgroup has a weaker treatment effect).

```{r stratified_power}
# Control hazard rates by stratum
lambdaC_pos <- log(2) / median_bm_pos
lambdaC_neg <- log(2) / median_bm_neg

# Enrollment rates by stratum (proportionate to prevalence)
gamma_pos <- bm_design$gamma
gamma_neg <- bm_design$gamma * (1 - prevalence) / prevalence

# Stratified inputs: matrix with columns = strata
lambdaC_strat <- matrix(c(lambdaC_pos, lambdaC_neg), ncol = 2)
gamma_strat <- matrix(c(gamma_pos, gamma_neg), ncol = 2)

pwr_overall <- gsSurvPower(
  k = 3, test.type = 4, alpha = 0.0125, sided = 1,
  sfu = sfHSD, sfupar = -4, sfl = sfHSD, sflpar = -2,
  lambdaC = lambdaC_strat, hr = 0.75, eta = 0.01,
  gamma = gamma_strat, R = 18, ratio = 1,
  plannedCalendarTime = c(12, 24, 36)
)

summary(pwr_overall)
gsBoundSummary(pwr_overall)
```

The overall design enrolls more patients (the full population rather than
just the 60% biomarker-positive subgroup) and has a higher event rate
in the biomarker-negative stratum (shorter control median), but assumes a
weaker overall treatment effect (HR = 0.75 vs. 0.65).
