---
title: "Futility and harm bounds for overall survival monitoring"
author: "Keaven Anderson"
output:
  rmarkdown::html_vignette:
    css: vignette.css
bibliography: gsDesign.bib
vignette: >
  %\VignetteIndexEntry{Futility and harm bounds for overall survival monitoring}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  dev.args = list(type = "cairo", antialias = "gray"),
  dpi = 80,
  fig.retina = 2,
  fig.width = 7.2916667,
  fig.asp = 0.618,
  fig.align = "center",
  out.width = "80%"
)

options(width = 68)
```

## Introduction

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(gsDesign)
library(knitr)
library(lt)
```

When clinical trials include overall survival (OS) as a secondary or exploratory endpoint, 
regulators may recommend not only monitoring for early evidence of efficacy and futility, 
but also for potential *harm* --- that is, evidence that the experimental treatment may 
be *worsening* survival relative to control.
This article demonstrates how the **gsDesign** package supports group sequential designs with 
three boundaries: an **efficacy** (upper) bound, a **futility** (lower) bound, and a **harm** bound, 
using `test.type = 7` (binding) and `test.type = 8` (non-binding).
For sparse mortality data without a separate futility rule, the section
[Sparse-event mortality monitoring](#sparse-event-mortality-monitoring)
uses `test.type = 6` and exact binomial boundary calculations.


### Regulatory context: FDA guidance on OS monitoring in oncology

The August 2025 FDA draft guidance *Approaches to Assessment of Overall
Survival in Oncology Clinical Trials* [@FDAOS2025] discusses OS as a safety
endpoint even when another endpoint establishes efficacy. It remains draft,
nonbinding guidance, not a requirement to use the method illustrated here.
Sections II, III.A, and III.B.2 discuss:

- Indolent diseases and long survival, where demonstrating OS superiority
  may be impractical.
- Prespecified harm assessments, justified thresholds, sufficient events, and
  precision to assess clinically relevant mortality detriment.
- Event-driven interim analyses when appropriate, with independent data
  monitoring. Some small trials or trials with low event rates may not warrant
  interim OS analyses.
- The distinction between identifying a harm signal and ruling out an
  unacceptable degree of harm.

The guidance does not prescribe 0.1/Pocock spending or a separate futility
boundary in every trial. Types 7 and 8 are useful when both futility and harm
are wanted; type 6 can instead supply a non-binding harm bound alone on the
lower side. The design and its assumptions require context-specific justification.

The harm bound implemented in gsDesign is a new method that is easy to use --- a principled, straightforward 
extension of the widely used group sequential spending function framework.
While we believe this approach is understandable, useful, and flexible, other methods for monitoring potential 
harm may also be considered.
However, there are limitations with this approach.
The first example has higher mortality risk than many cases. The later
sparse-event example examines what the approach can and cannot establish
when deaths are uncommon.

### Design framework overview

In a standard two-sided asymmetric group sequential design (`test.type = 3` or `4`), there are two boundaries:

- **Efficacy (upper) bound**: Reject $H_0$ if the test statistic exceeds this boundary (evidence of treatment benefit).
- **Futility (lower) bound**: Stop for futility if the test statistic falls below this boundary (insufficient evidence of treatment benefit).

The harm bound extension (`test.type = 7` or `8`) adds a third boundary:

- **Harm bound**: Signal that the experimental treatment may be harming patients (evidence of a *detrimental* effect).

The harm bound lies below the futility bound.
At each analysis, there are four possible outcomes:

1. **Cross the efficacy bound** (above): Stop for efficacy.
2. **Between the efficacy and futility bounds**: Continue the trial.
3. **Cross the futility bound but not the harm bound** (between futility and harm): Stop for futility.
4. **Cross the harm bound** (below): Stop for harm.

The harm bound is intended to flag sufficiently small observed p-values
*favoring control*. The nominal threshold depends on the analysis schedule
and spending specification; it is not a fixed p-value cutoff at every look.
That is, the harm bound flags evidence that the experimental treatment may be *worsening* survival --- a negative treatment effect on the log hazard ratio scale.

## Design with non-binding bounds (`test.type = 8`)

We demonstrate a survival design using `gsSurvCalendar()` with `test.type = 8` (non-binding futility and harm bounds).
The scenario is based on a 1:1 randomized trial monitoring overall survival with:

- **Median control survival**: 3 years (36 months), i.e., $\lambda_C = \log(2)/36$.
- **Target hazard ratio**: HR = 0.75 (25% reduction in hazard).
- **Power**: 90% ($\beta = 0.1$).
- **One-sided $\alpha$**: 0.0125 (e.g., the OS component of a trial with multiplicity adjustment).
- **Enrollment**: Uniform enrollment over 18 months.
- **Study duration**: 5 years (60 months) with planned analyses at years 1, 2, 3, 4, and 5 from start of enrollment.

The `astar` parameter controls the total spending for the harm bound under $H_0$.
The defaults are `astar = 0.1` (also selected by input `astar = 0`) and
`sfharm = sfLDPocock`. We specify these explicitly below for clarity.
The 10% target describes harm crossing under $H_0$ **if futility stopping is
ignored**, while still accounting for earlier efficacy and harm stops.
Actual harm stopping can be less frequent when futility is followed.
The existing cap that keeps harm at or below an active futility boundary can
also reduce the attainable spending.

### Spending function specification

We specify:

- **Efficacy bound**: Lan-DeMets O'Brien-Fleming (`sfLDOF`) spending function (conservative, spending little $\alpha$ at early analyses).
- **Futility bound**: Hwang-Shih-DeCani (HSD) spending function with $\gamma = -2$ (moderate $\beta$-spending under $H_1$).
- **Harm bound**: Lan-DeMets Pocock (`sfLDPocock`) spending function (spending under $H_0$ for detecting harm).

```{r}
x8 <- gsSurvCalendar(
  test.type = 8,
  alpha = 0.0125,
  beta = 0.1,
  astar = 0.1,
  calendarTime = c(12, 24, 36, 48, 60),
  sfu = sfLDOF,
  sfl = sfHSD, sflpar = -2,
  sfharm = sfLDPocock,
  lambdaC = log(2) / 36,
  hr = 0.75,
  R = 18,
  minfup = 42
)
```

### Summary

The `summary()` method provides a concise description of the design:

```{r}
cat(strwrap(summary(x8), width = 65), sep = "\n")
```

### Detailed boundary table

The `gsBoundSummary()` function produces a tabular summary with columns for each boundary.
By default, `B-value`, `Spending`, `CP`, `CP H1`, and `PP` are excluded.
We note that for the first interim analysis, the efficacy bound is so extreme it is effectively impossible to cross.
However, the harm and futility bounds are more moderate, allowing for early stopping if there is evidence of harm or futility.
The futility bound is an indicator of why bounds are often non-binding --- the futility bound is not intended to be a strict stopping rule, 
but rather a signal that the trial may be unlikely to succeed if it continues.
Crossing the harm bound is a stronger indication that the treatment may be harmful, and the trial should be at least paused with a recommendation to review the safety and other endpoint data.

```{r}
gsBoundSummary(x8) |> lt()
```

Conditional power (CP, CP H1) and predictive power (PP) can also be included in the summary.
Below we show the full table with all statistics, including conditional and predictive power at each boundary:

```{r}
gsBoundSummary(x8, exclude = c()) |> lt()
```

### Interpreting the boundaries

The design has five analyses at calendar times of 12, 24, 36, 48, and 60 months.
At each analysis, the test statistic (Z-value) is compared against three boundaries:

```{r}
bounds <- data.frame(
  Analysis = 1:x8$k,
  Month = x8$T,
  Events = ceiling(x8$n.I),
  Harm = round(x8$harm$bound, 2),
  Futility = round(x8$lower$bound, 2),
  Efficacy = round(x8$upper$bound, 2)
)
bounds |>
  lt() |>
  lt_header("Z-value boundaries at each analysis")
```

**Decision rules at an analysis where all three bounds are active:**

- If $Z >$ efficacy bound: Stop for efficacy (reject $H_0$).
- If futility bound $< Z \leq$ efficacy bound: Continue the trial.
- If harm bound $< Z \leq$ futility bound: Stop for futility.
- If $Z \leq$ harm bound: Stop for harm.

When both lower bounds are active, the harm bound is always at or below the futility bound.
If futility is skipped but harm is tested, the harm bound is the sole active lower stopping boundary.
The harm and futility bounds may coincide when the uncapped harm boundary
would exceed the futility boundary; the cap can prevent full harm spending.

### Boundary crossing probabilities

We examine the operating characteristics under two scenarios: no treatment effect (HR = 1, i.e., under $H_0$) and the design alternative (HR = 0.75).
When harm and futility are both active, `x8$lower$prob` and `x8$harm$prob` are
reported as mutually exclusive stopping outcomes. Thus, the probability of
crossing the futility threshold is the sum of the two lower tail components.

```{r}
probs <- data.frame(
  Scenario = c(rep("Under H0 (HR=1)", x8$k), rep("Under H1 (HR=0.75)", x8$k)),
  Analysis = rep(1:x8$k, 2),
  Month = rep(x8$T, 2),
  `P(Efficacy)` = c(cumsum(x8$upper$prob[, 1]), cumsum(x8$upper$prob[, 2])),
  `P(Futility only)` = c(cumsum(x8$lower$prob[, 1]), cumsum(x8$lower$prob[, 2])),
  `P(Harm)` = c(cumsum(x8$harm$prob[, 1]), cumsum(x8$harm$prob[, 2])),
  `P(Futility or Harm)` = c(
    cumsum(x8$lower$prob[, 1] + x8$harm$prob[, 1]),
    cumsum(x8$lower$prob[, 2] + x8$harm$prob[, 2])
  ),
  check.names = FALSE
)
probs |>
  lt() |>
  lt_format(columns = 2:7, decimals = 4) |>
  lt_header("Cumulative boundary crossing probabilities")
```

Under $H_0$, the actual cumulative probability of stopping for harm is
approximately `r round(sum(x8$harm$prob[, 1]), 4)` when all active stopping
rules are followed. This need not equal `astar`: trials that stop for futility
cannot subsequently stop for harm.
The cumulative probability of crossing the futility threshold, inclusive of harm, is approximately `r round(sum(x8$lower$prob[, 1] + x8$harm$prob[, 1]), 4)`.
Under $H_1$ (HR = 0.75), crossing the harm bound is very unlikely (`r round(sum(x8$harm$prob[, 2]), 4)`), since the treatment is beneficial.

### Harm calibration versus actual stopping

For both binding and non-binding designs, harm calibration ignores futility
stopping. This avoids making a harm threshold near zero merely to spend a
specified amount among the selected trials that survived earlier futility
monitoring. Efficacy and earlier harm stopping are still included.

The following example tests harm at all three looks, skips efficacy at IA1,
and tests futility only at IA1. It uses the new harm defaults. We compare the
target spending, harm crossing without futility, and actual harm stopping:

```{r}
xh <- gsDesign(
  test.type = 8, timing = c(.5, .75),
  testUpper = c(FALSE, TRUE, TRUE),
  testLower = c(TRUE, FALSE, FALSE),
  testHarm = TRUE
)
harm_without_futility <- gsProbability(
  k = xh$k, theta = 0, n.I = xh$n.I,
  a = xh$harm$bound, b = xh$upper$bound
)
data.frame(
  Analysis = c("IA1", "IA2", "Final"),
  `Harm Z` = xh$harm$bound,
  `Nominal harm p` = pnorm(xh$harm$bound),
  `Cumulative target` = cumsum(xh$harm$spend),
  `Harm without futility` = cumsum(harm_without_futility$lower$prob[, 1]),
  `Harm with futility` = cumsum(xh$harm$prob[, 1]),
  check.names = FALSE
) |> lt() |> lt_format(columns = 2:6, decimals = 4)
```

At IA2 the probability of an actual harm stop includes only paths that
continued past IA1. Thus the cumulative harm probability is
$P_0(Z_1 \le h_1) + P_0(Z_1 > f_1, Z_2 \le h_2)$, where $h_i$ denotes a
harm boundary and $f_1$ the IA1 futility boundary. Calibration without
futility instead uses $h_1$ in place of $f_1$ in the second term.

`gsBoundSummary()` reports **actual cumulative stopping probabilities**, not
the hypothetical probabilities used to calibrate harm spending. Its optional
`Spending` rows report the spending targets, so these two rows need not agree.
Also, its `p (1-sided)` rows use the efficacy direction; the nominal p-value
favoring harm is `pnorm(xh$harm$bound)`, the complement of that displayed
p-value. `print(xh)` labels the harm direction values as `Nominal p`.

Changing the example to `test.type = 7` retains the same harm calibration
convention. Binding status changes efficacy calibration: type 7 accounts for
both lower stopping rules in protecting alpha, whereas type 8 ignores them
for that purpose. Power and actual stopping probabilities include both rules
in either design. Ignoring a binding futility rule in practice can inflate
efficacy Type I error, even though it was ignored for harm calibration.

### Visualization

All standard `plot()` types are supported for `test.type = 7` and `8` designs, with a third line (or set of lines) shown for the harm bound.

#### Z-value boundaries

The default plot shows Z-value boundaries at each analysis.
Three boundaries are displayed: efficacy (upper), futility (lower), and harm (below futility).

```{r, fig.cap = "Z-value boundaries for non-binding harm bound design"}
plot(x8)
```

#### Boundary crossing probabilities

The power plot (`plottype = 2`) shows cumulative boundary crossing probabilities as a function of the treatment effect.
Three sets of lines appear: upper bound (cumulative efficacy crossing probability), `1-(Futility or harm)`, and `1-Harm`.
Because the harm boundary is nested below the futility boundary when both are active, crossing the futility threshold includes both futility-only and harm stops.
The `1-(Futility or harm)` curve therefore subtracts `x8$lower$prob + x8$harm$prob`, while the `1-Harm` curve subtracts harm crossings only.
This aggregation is performed only for plotting: the probability arrays stored in `x8` remain mutually exclusive so that efficacy, futility-only, and harm outcomes add without double counting.
When the underlying treatment effect favors control, the high probability of crossing the harm bound indicates that the harm bound is sensitive and serves its intended purpose.

```{r, fig.cap = "Boundary crossing probabilities for non-binding harm bound design"}
plot(x8, plottype = 2)
```

#### Approximate treatment effect at boundaries

The effect size plot (`plottype = 3`) shows the approximate treatment effect at each boundary.
For survival designs, this is expressed as the approximate hazard ratio at the boundary.

```{r, fig.cap = "Approximate treatment effect at boundaries"}
plot(x8, plottype = 3)
```

#### Conditional power at boundaries

Conditional power (`plottype = 4`) at each interim analysis is shown for all three boundaries.
This is generally not a very useful plot.

```{r, fig.cap = "Conditional power at boundaries"}
plot(x8, plottype = 4)
```

#### Spending function plot

The spending function plot (`plottype = 5`) shows the three spending functions: $\alpha$ (efficacy), $\beta$ (futility), and harm.

```{r, fig.cap = "Spending functions for non-binding harm bound design"}
plot(x8, plottype = 5)
```

#### B-values at boundaries

B-values (`plottype = 7`) are Z-values scaled by $\sqrt{t}$ where $t$ is the information fraction.
As discussed by @PLWBook, the expected value of B-values increases linearly with the information fraction under the assumption of a constant treatment effect (proportional hazards).
This linear relationship makes B-values useful for visual assessment of treatment effect trends across interim analyses: departures from linearity may suggest non-proportional hazards or other changes in treatment effect over time.
Three boundary lines are shown: efficacy, futility, and harm.

```{r, fig.cap = "B-values at boundaries"}
plot(x8, plottype = 7)
```

## Design with binding bounds (`test.type = 7`)

For `test.type = 7`, both the futility and harm bounds are **binding** --- meaning the computation of the efficacy bound assumes the trial *will* stop if either bound is crossed.
This yields a slightly less conservative efficacy bound (easier to cross), but at the cost of inflated Type I error if the stopping rule is not strictly followed.

We first create a binding design with $\alpha = 0.0125$ to compare with the non-binding design above:

```{r}
x7 <- gsSurvCalendar(
  test.type = 7,
  alpha = 0.0125,
  beta = 0.1,
  astar = 0.1,
  calendarTime = c(12, 24, 36, 48, 60),
  sfu = sfLDOF,
  sfl = sfHSD, sflpar = -2,
  sfharm = sfLDPocock,
  lambdaC = log(2) / 36,
  hr = 0.75,
  R = 18,
  minfup = 42
)
```

### Comparing binding and non-binding

```{r}
comparison <- data.frame(
  Bound = c("Efficacy", "Futility", "Harm"),
  `Binding (type 7)` = c(
    paste(round(x7$upper$bound, 3), collapse = ", "),
    paste(round(x7$lower$bound, 3), collapse = ", "),
    paste(round(x7$harm$bound, 3), collapse = ", ")
  ),
  `Non-binding (type 8)` = c(
    paste(round(x8$upper$bound, 3), collapse = ", "),
    paste(round(x8$lower$bound, 3), collapse = ", "),
    paste(round(x8$harm$bound, 3), collapse = ", ")
  ),
  check.names = FALSE
)
comparison |>
  lt() |>
  lt_header("Comparison of binding vs. non-binding Z-value boundaries")
```

Note that the efficacy bounds for `test.type = 7` (binding) are slightly lower (easier to cross) than for `test.type = 8` (non-binding).
The maximum number of events for `test.type = 7` (`r ceiling(max(x7$n.I))`) is also slightly smaller than for `test.type = 8` (`r ceiling(max(x8$n.I))`), reflecting the assumption that the trial will stop at the lower bounds.

```{r}
gsBoundSummary(x7) |> lt()
```

### Efficacy bounds at alternate $\alpha$ levels

The `gsBoundSummary()` function accepts an `alpha` argument to display efficacy bounds at one or more alternate $\alpha$ levels alongside the original design.
Each alternate alpha column retains the `testUpper` schedule from the design, so an efficacy analysis that was skipped remains inactive and every efficacy characteristic at that analysis, including cumulative crossing probability, is `NA`.
Here we show the non-binding design (`x8`) with efficacy bounds for both $\alpha = 0.0125$ (the design level) and $\alpha = 0.025$:

```{r}
gsBoundSummary(x8, alpha = 0.025) |> lt()
```

Alternate alpha summaries are supported for `test.type = 8`, where both
futility and harm bounds are non-binding. They are not supported for
`test.type = 7`: a binding harm or futility bound is outside the non-binding
sequential p-value framework used for Maurer--Bretz graphical multiple testing.

## Sparse-event mortality monitoring

### A harm bound without a separate futility bound

Consider a randomized trial with a primary efficacy endpoint other than OS,
such as progression-free survival, and relatively few expected deaths. A
prespecified mortality harm signal can be useful without requiring a separate
rule for stopping because OS benefit appears unlikely. This is a hypothetical
illustration, not an FDA-specified disease scenario or recommended protocol.

Use `test.type = 6`: its lower Z-bound spends under the null hypothesis and
is **non-binding for efficacy Type I error**. Here we interpret that lower
bound as harm; there is no additional beta spending futility bound.
For type 6 the inputs are `astar = 0.1` and `sfl = sfLDPocock`, **not**
`sfharm`. They must be supplied explicitly: type 6 retains its general-purpose
defaults of `astar = 1 - alpha` (selected by input 0) and HSD lower spending.
The lower boundary is stored in `lower`, even if a generic summary labels it
as futility. Its clinical interpretation in this example is harm.

We plan analyses at 6, 23, and 40 total deaths, with 1:1 randomization,
enrollment of 100 participants per month for 16 months, no dropout, and a
constant control mortality hazard of 0.001 per month. An HR of 0.8 is used
only as a planning scenario for the expected timing and power, not as evidence
of mortality benefit. `gsSurvPower()` evaluates this fixed event schedule;
it does not increase the event target to obtain 90% OS power.

```{r sparse-design}
sparse <- gsSurvPower(
  k = 3, test.type = 6,
  alpha = .025, sided = 1,
  astar = .1, sfl = sfLDPocock,
  targetEvents = c(6, 23, 40),
  lambdaC = .001, hr = .8, hr0 = 1,
  gamma = 100, R = 16, ratio = 1, eta = 0
)
# Use the planned final event count as the explicit spending denominator.
event_fraction <- sparse$n.I / tail(sparse$n.I, 1)
sparse_exact <- toBinomialExact(
  sparse, usTime = event_fraction, lsTime = event_fraction
)
```

The example retains an OS efficacy boundary to illustrate non-binding harm
monitoring alongside an alpha-controlled efficacy test. It does not allocate
alpha to another primary endpoint or supply a multiple endpoint testing
strategy. If OS is solely a safety endpoint, the efficacy stopping rules and
operating characteristics should be specified accordingly rather than
claiming an OS efficacy objective that the trial does not have.

### Integer death count boundaries

In the exact binomial representation the direction reverses: a high count
of experimental arm deaths signals harm. `upper$bound` is therefore the
harm cutoff, while `lower$bound` is the efficacy cutoff. The probabilities
account for stopping at any earlier active boundary.

```{r sparse-boundaries}
data.frame(
  `Total deaths` = sparse_exact$n.I,
  `Experimental deaths for harm (at least)` = sparse_exact$upper$bound,
  `Cumulative harm probability under HR=1` =
    cumsum(sparse_exact$upper$prob[, 1]),
  check.names = FALSE
) |> lt() |> lt_format(columns = 3, decimals = 4)

# Efficacy Type I error is evaluated ignoring the non-binding harm boundary.
efficacy_without_harm <- gsBinomialExact(
  k = sparse_exact$k, theta = .5, n.I = sparse_exact$n.I,
  a = sparse_exact$lower$bound, b = sparse_exact$n.I + 1L
)
stopifnot(sum(efficacy_without_harm$lower$prob) <= sparse$alpha + 1e-10)
```

At the first look, a signal requires all six deaths to be in the experimental
arm. At later looks the cumulative cutoff is applied only if no earlier
stopping boundary has been crossed. The total null harm crossing probability
is `r round(100 * sum(sparse_exact$upper$prob[, 1]), 1)`%, below the 10% target
because only integer death counts are available. The efficacy Type I error
without harm stopping is
`r round(100 * sum(efficacy_without_harm$lower$prob), 2)`%, below 2.5%.

### Sensitivity to excess mortality

For balanced exposure and uncommon events, approximate the probability that
a death is in the experimental arm by $p = HR/(1 + HR)$, where
$HR = \lambda_E/\lambda_C$. More generally the corresponding conditional
Poisson model uses $p = HR\,r/(1 + HR\,r)$ with exposure ratio $r$.
The recursion below is exact for independent binomial event increments with
constant $p$, **not automatically exact for arbitrary censored survival data**.
Risk set imbalance, differential censoring, or changing hazards can invalidate
the simple mapping from randomization ratio and HR. Such settings need
appropriate survival model calculations or simulation.

```{r sparse-sensitivity}
harm_hr <- c(1, 1.25, 1.5, 2)
harm_oc <- gsBinomialExact(
  k = sparse_exact$k, theta = harm_hr / (1 + harm_hr),
  n.I = sparse_exact$n.I,
  a = sparse_exact$lower$bound, b = sparse_exact$upper$bound
)
data.frame(
  `True mortality HR` = harm_hr,
  `Probability of a harm signal` = colSums(harm_oc$upper$prob),
  check.names = FALSE
) |> lt() |> lt_format(columns = 2, decimals = 3)
```

Even at HR = 1.5, the chance of a signal is only
`r round(100 * sum(harm_oc$upper$prob[, 3]), 1)`% in this example. Failure to
cross the boundary therefore cannot establish that a clinically important
mortality increase has been excluded. Increasing harm spending can improve
sensitivity, but also increases false signals; it cannot replace information
from additional deaths and follow-up. A non-binding designation concerns
efficacy alpha calculations, not the clinical importance of reviewing a signal.

### Detecting harm versus ruling out harm

A harm monitoring test asks whether data provide evidence of excess mortality
relative to HR = 1. A reassurance objective instead asks whether the data can
exclude a prespecified unacceptable HR, such as 1.5. The latter value is an
illustration, not a universal safety margin. One approach is to require an
appropriately constructed upper confidence bound to fall below that margin.

To illustrate the lack of precision, suppose a **single fixed analysis** has
20 deaths in each arm. Under the same binomial model, transform a one-sided
95% upper bound for the experimental share into an upper HR bound:

```{r sparse-precision}
p_upper <- binom.test(20, 40, alternative = "less", conf.level = .95)$conf.int[2]
hr_upper <- p_upper / (1 - p_upper)
data.frame(
  `Experimental deaths` = 20, `Control deaths` = 20,
  `One-sided 95% upper HR bound` = hr_upper,
  check.names = FALSE
) |> lt() |> lt_format(columns = 3, decimals = 2)
```

Despite a balanced observed count, this upper bound exceeds 1.5. This is a
fixed analysis precision illustration, **not a sequentially adjusted interval**
for the monitored trial. Repeated or stopping-selected reassurance assessments
require an interval or test valid for that monitoring plan. A vignette example
must not turn "no harm signal" into a claim of safety: the harm margin,
precision, event target, follow-up, and operating characteristics for ruling
out harm all need their own justification.

## Practical considerations

### Choice of spending functions

The choice of spending functions for the three boundaries should reflect regulatory and scientific considerations:

- **Efficacy**: A conservative spending function such as Lan-DeMets O'Brien-Fleming (`sfLDOF`) is typical, spending very little $\alpha$ at early interim analyses when limited information is available.
- **Futility**: Moderate spending (e.g., HSD with $\gamma = -2$) allows early stopping for futility when the treatment effect is clearly absent.
- **Harm**: The Lan-DeMets Pocock (`sfLDPocock`) spending function provides more aggressive spending at early analyses, which is appropriate for harm monitoring since detecting a detrimental effect early is critical for patient safety.

### Total harm spending and the number of analyses

Lan--DeMets Pocock spending with total 0.1 allocates cumulative probability
$A(t) = 0.1\log\{1 + (e-1)t\}$ by spending time $t$. It allocates relatively
more spending early than a conservative O'Brien--Fleming choice, helping to
flag moderately small nominal p-values favoring harm throughout monitoring.
The total 0.1 is a repeated monitoring probability under the null, **not** a
nominal 0.1 cutoff at every analysis. Neither the total nor the spending
function guarantees that any particular nominal p-value will trigger harm
stopping at every look.

With more analyses, a user may wish to choose a **larger total harm spending**
(`astar`) to retain similarly permissive nominal thresholds. This increases
the allowed probability of a false harm signal under the null. It does not
mean increasing futility spending (`beta`), which has a different purpose and
is calibrated under the design alternative. Inspect the actual thresholds
for the intended timing and number of analyses. For example:

```{r}
harm_nominal <- function(k, astar) {
  d <- gsDesign(k = k, test.type = 8, astar = astar)
  data.frame(
    Analyses = k, Total = astar, Look = seq_len(k),
    `Nominal harm p` = pnorm(d$harm$bound), check.names = FALSE
  )
}
rbind(harm_nominal(3, .1), harm_nominal(6, .1), harm_nominal(6, .2)) |>
  lt() |> lt_format(columns = 4, decimals = 4)
```

The larger total in this illustration is a sensitivity analysis, not a
universal recommendation. The choice requires considering both sensitivity
to potential harm and false harm signals.

### Interpreting the harm bound

The harm bound is crossed when the observed p-value *favoring control* is at
or below the analysis-specific nominal harm threshold.
In terms of the test statistic, a negative Z-value indicates that the hazard rate is higher in the experimental arm than the control arm --- i.e., the experimental treatment appears to be worsening survival.
When the Z-value falls below the harm bound, this constitutes a statistical signal that the treatment may be harmful, and the trial should be stopped with a recommendation to review the safety data.

The harm spending is computed under $H_0$ (no treatment effect), ignoring
futility stopping but retaining efficacy stopping. It budgets the probability
of a false harm signal under those assumptions; following futility can reduce
the actual probability further.

### Harm bound capping

In the implementation, the harm bound is automatically **capped** so it never exceeds the futility bound when both are active.
This ensures the ordering harm bound $\leq$ futility bound $\leq$ efficacy bound at analyses where all three are tested, while allowing harm to be the active lower boundary when futility is skipped.

### When to use `test.type = 7` vs. `test.type = 8`

- **`test.type = 8` (non-binding)** is most often preferred in practice. Regulators will generally expect non-binding bounds, which preserve Type I error control regardless of whether the stopping rules are strictly followed. Since Data Monitoring Committees (DMCs) typically retain discretion to continue or stop a trial based on the totality of the evidence, the non-binding approach ensures that the statistical validity of the efficacy analysis is maintained even if a futility or harm boundary is crossed but the trial continues.
- **`test.type = 7` (binding)** is appropriate when there is a firm commitment to stop the trial upon crossing any boundary. This provides a small efficiency gain (slightly easier efficacy bounds and fewer required events) but requires strict protocol adherence. If the trial does not stop after crossing a binding boundary, Type I error may be inflated.

In most regulatory settings, `test.type = 8` is the safer and more common choice.

### Binding and non-binding harm monitoring

When futility and harm are both tested at an analysis, the harm boundary lies below the futility boundary and therefore does not add another stopping region. Crossing probabilities are nevertheless partitioned into mutually exclusive harm, futility, and efficacy outcomes.

When harm is tested at an analysis where futility is skipped, the harm boundary is the active lower stopping boundary. This stopping probability is included in sample size derivation and, for `test.type = 7`, in the binding efficacy bound calculation. For `test.type = 8`, efficacy bounds continue to use the non-binding Type I error convention, while power and expected information account for actual harm and futility stopping.

### Adjusting the boundaries

The boundaries are adjustable through several design parameters:

- **Alternate `astar`**: Controls the Type I error allocated to excess OS harm detection.
- **Alternate spending functions**: Different spending functions for efficacy, futility, and harm boundaries change the aggressiveness of each boundary across analyses.
- **Alternate timing of analyses**: Changing the calendar times of interim analyses shifts the information available at each look.

Regardless of the statistical design, bounds must be clinically, ethically, and statistically sound.
As previously noted, this approach is one option to address the regulatory expectation for OS harm monitoring, 
but other approaches may also be considered.

## References
