---
title: "Multi-season studies for rare events"
output: rmarkdown::html_vignette
bibliography: gsDesign.bib
vignette: >
  %\VignetteIndexEntry{Multi-season studies for rare events}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "svg",
  fig.ext = "svg",
  fig.width = 7.2916667,
  fig.asp = 0.618,
  fig.align = "center",
  out.width = "80%"
)

options(width = 58)
```

```{r, message=FALSE, warning=FALSE}
library(gsDesign)
library(gt)
library(tibble)
```

## Introduction

This vignette demonstrates a practical design workflow for rare seasonal events.
The motivating use case is a vaccine efficacy trial with annual high-risk seasons.
The same ideas can be adapted to infectious disease studies where simple superiority
or non-inferiority may be considered.

We focus on exact binomial monitoring with spending functions:

- analyses after each season with spending times `1/3, 2/3, 1`,
- efficacy spending based on `sfHSD` with `gamma = 1` (Pocock-like),
- exact binomial p-values from `repeatedPValueBinomialExact()` and
  `sequentialPValueBinomialExact()`.

We include both a fixed group sequential design and a blinded information-adaptive
option that can increase future enrollment if events accrue slowly.
Conceptually, enrollment is concentrated in early fall (for example, September
and October), and each analysis uses a data cut at the end of the corresponding
high-risk season (for example, May 1).

## Assumptions and parameterization

We use a super-superiority example:

- null vaccine efficacy (VE) = 30% (`HR0 = 0.7`),
- alternative VE = 80% (`HR = 0.2`),
- one-sided `alpha = 0.025`, `beta = 0.1`,
- randomization ratio = 3:1 (experimental:control),
- control seasonal event rate = 0.3% over 6 months,
- dropout modeled as exponential with 10% dropout over each 6-month season.

```{r}
alpha <- 0.025
beta <- 0.1
ratio <- 3

ve0 <- 0.30
ve1 <- 0.80
hr0 <- 1 - ve0
hr1 <- 1 - ve1

seasonal_event_rate_control <- 0.003
season_length_years <- 0.5
dropout_6mo <- 0.10

timing <- c(1 / 3, 2 / 3, 1)
```

For the exact binomial approximation, the probability that an event is in the
experimental group is

```{r}
p_event_experimental <- function(ve, ratio = 3) {
  ratio / (ratio + 1 / (1 - ve))
}

p0 <- p_event_experimental(ve0, ratio)
p1 <- p_event_experimental(ve1, ratio)
c(p0 = p0, p1 = p1)
```

## Initial group sequential setup

To define spending-function monitoring, we first construct a time-to-event
`gsSurv` object and then convert to integer event looks and exact binomial
bounds. Here we use three looks with timing fractions `1/3, 2/3, 1`.
Enrollment is modeled as 2 months on and 10 months off each year for 3 years.

```{r}
lambdaC <- -log(1 - seasonal_event_rate_control) / season_length_years
eta <- -log(1 - dropout_6mo) / season_length_years
enroll_pattern <- c(1, 0, 1, 0, 1, 0)
enroll_periods <- c(2, 10, 2, 10, 2, 10)

design_tte <- gsSurv(
  k = 3,
  test.type = 4,
  alpha = alpha,
  beta = beta,
  timing = timing[1:2],
  sfu = sfHSD,
  sfupar = 1,
  sfl = sfHSD,
  sflpar = -2,
  lambdaC = lambdaC,
  hr = hr1,
  hr0 = hr0,
  eta = eta,
  gamma = enroll_pattern,
  R = enroll_periods,
  T = 42,
  minfup = 6,
  ratio = ratio
) |>
  toInteger()

planned_final_events <- design_tte$n.I[design_tte$k]
planned_counts <- as.integer(round(planned_final_events * timing))
planned_counts[3] <- planned_final_events
planned_counts[2] <- max(planned_counts[2], planned_counts[1] + 1L)
planned_counts[3] <- max(planned_counts[3], planned_counts[2] + 1L)

design_exact <- toBinomialExact(design_tte, observedEvents = planned_counts)
```

The table below summarizes planned event looks and exact efficacy bounds.

```{r}
target_alpha_spend <- design_tte$upper$sf(
  alpha = alpha,
  t = timing,
  param = design_tte$upper$param
)$spend

achieved_alpha_spend <- cumsum(
  gsBinomialExact(
    k = design_exact$k,
    theta = design_exact$theta[1],
    n.I = design_exact$n.I,
    a = design_exact$lower$bound,
    b = design_exact$n.I + 1
  )$lower$prob[, 1]
)

tibble(
  Season = 1:3,
  `Spending time` = timing,
  `Planned total events` = design_exact$n.I,
  `Exact efficacy bound (x <= a)` = design_exact$lower$bound,
  `Target cumulative alpha spend` = target_alpha_spend,
  `Achieved cumulative alpha spend` = achieved_alpha_spend
) |>
  gt() |>
  fmt_number(columns = 2, decimals = 3) |>
  fmt_number(columns = 5:6, decimals = 4) |>
  tab_header(
    title = "Planned exact binomial seasonal monitoring",
    subtitle = "Super-superiority example with Pocock-like efficacy spending"
  )
```

## Example repeated and sequential p-values

Suppose observed event totals at the three seasonal analyses are equal to the
planned totals and observed experimental events are as below.

```{r}
example_x <- pmax(0, design_exact$lower$bound - c(0L, 1L, 0L))
example_p <- repeatedPValueBinomialExact(
  gsD = design_tte,
  n.I = design_exact$n.I,
  x = example_x
)
example_p
```

The sequential p-value is the minimum repeated p-value:

```{r}
sequentialPValueBinomialExact(
  gsD = design_tte,
  n.I = design_exact$n.I,
  x = example_x
)
```

## Seasonal event/dropout model for simulation

For each arm and season, we use a competing-risk approximation with constant
event and dropout hazards over 6 months:

- event before dropout,
- dropout before event,
- neither event nor dropout (carries into next season).

In the simulation, we add a new enrollment cohort each season to represent
annual enrollment windows.

```{r}
seasonal_probs <- function(ve, lambda_c, eta, season_years = 0.5) {
  lambda_e <- lambda_c * (1 - ve)
  arm_probs <- function(lambda_event) {
    total <- lambda_event + eta
    p_event <- lambda_event / total * (1 - exp(-total * season_years))
    p_drop <- eta / total * (1 - exp(-total * season_years))
    p_stay <- exp(-total * season_years)
    c(event = p_event, dropout = p_drop, stay = p_stay)
  }
  list(control = arm_probs(lambda_c), experimental = arm_probs(lambda_e))
}

probs_alt <- seasonal_probs(ve = ve1, lambda_c = lambdaC, eta = eta)
probs_null <- seasonal_probs(ve = ve0, lambda_c = lambdaC, eta = eta)
```

To keep the simulation aligned with the planned final information, we select a
per-season enrollment target so expected total events under the alternative are
close to the planned final event count.

```{r}
expected_events_3_seasons <- function(p_event, p_stay) {
  p_event * (3 + 2 * p_stay + p_stay^2)
}

f_control <- expected_events_3_seasons(
  p_event = probs_alt$control["event"],
  p_stay = probs_alt$control["stay"]
)
f_experimental <- expected_events_3_seasons(
  p_event = probs_alt$experimental["event"],
  p_stay = probs_alt$experimental["stay"]
)

enroll_control_per_season <- ceiling(planned_final_events / (f_control + ratio * f_experimental))
enroll_experimental_per_season <- ratio * enroll_control_per_season

tibble(
  Arm = c("Control", "Experimental"),
  `Enrolled each season` = c(enroll_control_per_season, enroll_experimental_per_season)
) |>
  gt() |>
  tab_header(title = "Simulation enrollment assumptions")
```

## Fixed and adaptive simulation functions

The adaptive rule is intentionally simple and blinded:

- after season 1, increase season-2 enrollment if total events are below
  planned look-1 events,
- after season 2, optionally increase season-3 enrollment similarly,
- spending fractions remain fixed at `1/3, 2/3, 1`.

```{r}
simulate_one_trial <- function(
    ve,
    adaptive = FALSE,
    max_multiplier = 2,
    enroll_control = enroll_control_per_season,
    enroll_experimental = enroll_experimental_per_season) {
  probs <- seasonal_probs(ve = ve, lambda_c = lambdaC, eta = eta)

  enroll_c <- rep(enroll_control, 3L)
  enroll_e <- rep(enroll_experimental, 3L)
  at_risk_c <- 0L
  at_risk_e <- 0L

  cum_total <- integer(3L)
  cum_exp <- integer(3L)

  for (season in 1:3) {
    if (adaptive && season == 2) {
      mult <- min(max_multiplier, max(1, planned_counts[1] / max(cum_total[1], 1)))
      enroll_c[2] <- as.integer(ceiling(enroll_control * mult))
      enroll_e[2] <- as.integer(ceiling(enroll_experimental * mult))
    }
    if (adaptive && season == 3) {
      mult <- min(max_multiplier, max(1, planned_counts[2] / max(cum_total[2], 1)))
      enroll_c[3] <- as.integer(ceiling(enroll_control * mult))
      enroll_e[3] <- as.integer(ceiling(enroll_experimental * mult))
    }

    at_risk_c <- at_risk_c + enroll_c[season]
    at_risk_e <- at_risk_e + enroll_e[season]

    out_c <- as.integer(rmultinom(1, at_risk_c, probs$control))
    out_e <- as.integer(rmultinom(1, at_risk_e, probs$experimental))

    inc_total <- out_c[1] + out_e[1]
    inc_exp <- out_e[1]

    cum_total[season] <- if (season == 1) inc_total else cum_total[season - 1] + inc_total
    cum_exp[season] <- if (season == 1) inc_exp else cum_exp[season - 1] + inc_exp

    at_risk_c <- out_c[3]
    at_risk_e <- out_e[3]
  }

  # Use analyses through the first look where cumulative events reach/exceed
  # planned final events, so only one look has spending time >= 1.
  reached_final <- which(cum_total >= planned_final_events)
  looks <- if (length(reached_final) == 0) 1:3 else seq_len(reached_final[1])

  if (cum_total[looks[1]] < 1) {
    return(list(reject = FALSE, looks = length(looks), total_events = cum_total[3]))
  }

  # Internal helper is used only for vignette simulation speed.
  bounds_at_alpha <- gsDesign:::binomialExactLowerBound(
    gsD = design_tte,
    n.I = cum_total[looks],
    alpha = alpha
  )
  reject <- any(cum_exp[looks] <= bounds_at_alpha)

  list(
    reject = reject,
    looks = length(looks),
    total_events = cum_total[3]
  )
}

simulate_scenario <- function(ve, adaptive = FALSE, nsim = 600, seed = 1) {
  set.seed(seed)
  out <- replicate(nsim, simulate_one_trial(ve = ve, adaptive = adaptive), simplify = FALSE)
  tibble(
    reject = as.logical(vapply(out, `[[`, logical(1), "reject")),
    looks = as.integer(vapply(out, `[[`, integer(1), "looks")),
    total_events = as.integer(vapply(out, `[[`, integer(1), "total_events"))
  )
}
```

## Lightweight runnable simulation

```{r}
sim_fixed_null <- simulate_scenario(ve = ve0, adaptive = FALSE, nsim = 150, seed = 101)
sim_fixed_alt <- simulate_scenario(ve = ve1, adaptive = FALSE, nsim = 150, seed = 202)
sim_adapt_null <- simulate_scenario(ve = ve0, adaptive = TRUE, nsim = 150, seed = 303)
sim_adapt_alt <- simulate_scenario(ve = ve1, adaptive = TRUE, nsim = 150, seed = 404)

summarize_oc <- function(x) {
  p <- mean(x)
  se <- sqrt(p * (1 - p) / length(x))
  c(estimate = p, mc_se = se)
}

oc <- rbind(
  `Fixed: Type I error (VE = 30%)` = summarize_oc(sim_fixed_null$reject),
  `Fixed: Power (VE = 80%)` = summarize_oc(sim_fixed_alt$reject),
  `Adaptive: Type I error (VE = 30%)` = summarize_oc(sim_adapt_null$reject),
  `Adaptive: Power (VE = 80%)` = summarize_oc(sim_adapt_alt$reject)
)

tibble(
  Scenario = rownames(oc),
  Estimate = oc[, "estimate"],
  `Monte Carlo SE` = oc[, "mc_se"]
) |>
  gt() |>
  fmt_number(columns = 2:3, decimals = 4) |>
  tab_header(
    title = "Lightweight simulation results",
    subtitle = "Exact-binomial monitoring with seasonal analyses"
  )
```

```{r}
tibble(
  Scenario = c("Fixed", "Adaptive"),
  `Mean final events under VE=30%` = c(mean(sim_fixed_null$total_events), mean(sim_adapt_null$total_events)),
  `Mean final events under VE=80%` = c(mean(sim_fixed_alt$total_events), mean(sim_adapt_alt$total_events)),
  `Mean looks used under VE=30%` = c(mean(sim_fixed_null$looks), mean(sim_adapt_null$looks)),
  `Mean looks used under VE=80%` = c(mean(sim_fixed_alt$looks), mean(sim_adapt_alt$looks))
) |>
  gt() |>
  fmt_number(columns = 2:5, decimals = 2) |>
  tab_header(title = "Monitoring behavior in simulation")
```

## Larger offline runs (template)

The chunk below is intentionally not executed in package builds.

```{r, eval=FALSE}
# Suggested offline settings
type1_nsim <- 20000
power_nsim <- 3500

sim_fixed_null_big <- simulate_scenario(ve = ve0, adaptive = FALSE, nsim = type1_nsim, seed = 5001)
sim_adapt_null_big <- simulate_scenario(ve = ve0, adaptive = TRUE, nsim = type1_nsim, seed = 5002)

sim_fixed_alt_big <- simulate_scenario(ve = ve1, adaptive = FALSE, nsim = power_nsim, seed = 6001)
sim_adapt_alt_big <- simulate_scenario(ve = ve1, adaptive = TRUE, nsim = power_nsim, seed = 6002)
```

## Notes and extensions

- The adaptive pathway shown here is blinded because it uses total event counts
  only; treatment-group differences are not used to update enrollment.
- Spending fractions are kept fixed at `1/3, 2/3, 1`.
- For modified intention-to-treat analyses, additional exclusion/dropout
  mechanisms can be layered into the simulation by reducing at-risk counts
  before each seasonal event/dropout draw.
- Simple superiority and non-inferiority variants can be studied by changing
  `hr0`/`hr1` and rebuilding the same workflow.

## References
