---
title: "Multi-season studies for rare events"
output: rmarkdown::html_vignette
bibliography: gsDesign.bib
vignette: >
  %\VignetteIndexEntry{Multi-season studies for rare events}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "svg",
  fig.ext = "svg",
  fig.width = 7.2916667,
  fig.asp = 0.618,
  fig.align = "center",
  out.width = "80%"
)

options(width = 58)
```

```{r, message=FALSE, warning=FALSE}
library(gsDesign)
library(gt)
library(tibble)
```

## Introduction

This vignette demonstrates a practical design workflow for rare seasonal events.
The motivating use case is a vaccine efficacy trial with annual high-risk seasons.
The same ideas can be adapted to infectious disease studies where simple superiority
or non-inferiority may be considered.

We focus on exact binomial monitoring with spending functions:

- analyses after each season with spending times `1/3, 2/3, 1`,
- efficacy spending based on `sfHSD` with `gamma = 1` (Pocock-like),
- exact binomial p-values from `repeatedPValueBinomialExact()` and
  `sequentialPValueBinomialExact()`.

We include both a fixed group sequential design and a blinded information-adaptive
option that can increase future enrollment if events accrue slowly.
Conceptually, enrollment is concentrated in early fall (for example, September
and October), and each analysis uses a data cut at the end of the corresponding
high-risk season (for example, May 1).

## Assumptions and parameterization

We use a super-superiority example:

- null vaccine efficacy (VE) = 30% (`HR0 = 0.7`),
- alternative VE = 80% (`HR = 0.2`),
- one-sided `alpha = 0.025`, `beta = 0.1`,
- randomization ratio = 3:1 (experimental:control),
- control seasonal event rate = 0.3% over 6 months,
- dropout modeled as exponential with 10% dropout over each 6-month season.

```{r}
alpha <- 0.025
beta <- 0.1
ratio <- 3

ve0 <- 0.30
ve1 <- 0.80
hr0 <- 1 - ve0
hr1 <- 1 - ve1

seasonal_event_rate_control <- 0.003
season_length_years <- 0.5
dropout_6mo <- 0.10

timing <- c(1 / 3, 2 / 3, 1)
```

For the exact binomial approximation, the probability that an event is in the
experimental group is

```{r}
p_event_experimental <- function(ve, ratio = 3) {
  ratio / (ratio + 1 / (1 - ve))
}

p0 <- p_event_experimental(ve0, ratio)
p1 <- p_event_experimental(ve1, ratio)
c(p0 = p0, p1 = p1)
```

## Initial group sequential setup

To define spending-function monitoring, we first construct a time-to-event
`gsSurv` object and then convert to integer event looks and exact binomial
bounds. Here we use three looks with timing fractions `1/3, 2/3, 1`.
Enrollment is modeled as 2 months on and 10 months off each year for 3 years.

```{r}
lambdaC <- -log(1 - seasonal_event_rate_control) / season_length_years
eta <- -log(1 - dropout_6mo) / season_length_years
enroll_pattern <- c(1, 0, 1, 0, 1, 0)
enroll_periods <- c(2, 10, 2, 10, 2, 10)

design_tte <- gsSurv(
  k = 3,                  # Number of analyses (3 seasons)
  test.type = 4,          # Non-binding lower bound framework
  alpha = alpha,          # One-sided Type I error
  beta = beta,            # Type II error
  timing = timing[1:2],   # Interim information fractions (season 1 and 2)
  sfu = sfHSD,            # Efficacy spending function
  sfupar = 1,             # Pocock-like efficacy spending
  sfl = sfHSD,            # Futility spending function
  sflpar = -2,            # Futility spending parameter
  lambdaC = lambdaC,      # Control hazard rate during high-risk season
  hr = hr1,               # Alternative hypothesis HR
  hr0 = hr0,              # Null hypothesis HR
  eta = eta,              # Dropout hazard rate
  gamma = enroll_pattern, # Relative enrollment rates by period
  R = enroll_periods,     # Period durations (2 months on, 10 months off)
  T = 42,                 # Trial duration in months
  minfup = 6,             # Minimum follow-up for final enrollees
  ratio = ratio,          # Experimental:control randomization ratio
  testLower = c(TRUE, FALSE, FALSE) # Futility only at IA1
) |>
  toInteger()

gsBoundSummary(design_tte)

planned_final_events <- design_tte$n.I[design_tte$k]
planned_counts <- as.integer(round(planned_final_events * timing))
planned_counts[3] <- planned_final_events
planned_counts[2] <- max(planned_counts[2], planned_counts[1] + 1L)
planned_counts[3] <- max(planned_counts[3], planned_counts[2] + 1L)

design_exact <- toBinomialExact(design_tte, observedEvents = planned_counts)
planned_cum_enrollment <- as.integer(round(rowSums(design_tte$eNC + design_tte$eNE)))

planned_enrollment_period <- as.numeric(rowSums(as.matrix(design_tte$gamma))) * enroll_periods
season_id <- rep(1:3, each = 2)
planned_enrollment_by_season <- as.integer(round(tapply(planned_enrollment_period, season_id, sum)))
planned_enrollment_control <- as.integer(round(planned_enrollment_by_season / (1 + ratio)))
planned_enrollment_experimental <- planned_enrollment_by_season - planned_enrollment_control
```

The table above is the initial survival-design approximation and includes
the approximate futility `Z`-boundary. The table below summarizes planned
event looks and exact binomial efficacy/futility bounds.
In this table, `x` is the cumulative number of observed events in the
experimental arm at a given analysis.

```{r}
target_alpha_spend <- design_tte$upper$sf(
  alpha = alpha,
  t = timing,
  param = design_tte$upper$param
)$spend

achieved_alpha_spend <- cumsum(
  gsBinomialExact(
    k = design_exact$k,            # Number of analyses
    theta = design_exact$theta[1], # Null event probability in experimental arm
    n.I = design_exact$n.I,        # Cumulative events by analysis
    a = design_exact$lower$bound,  # Efficacy bounds (x <= a crosses)
    b = design_exact$n.I + 1       # Non-binding upper bound for alpha-spend check
  )$lower$prob[, 1]
)

achieved_power_h1 <- cumsum(
  gsBinomialExact(
    k = design_exact$k,            # Number of analyses
    theta = design_exact$theta[2], # Alternative event probability in experimental arm
    n.I = design_exact$n.I,        # Cumulative events by analysis
    a = design_exact$lower$bound,  # Efficacy bounds
    b = design_exact$n.I + 1       # Non-binding upper bound for cumulative efficacy probability
  )$lower$prob[, 1]
)

ve_from_bound <- function(x, n, ratio) {
  out <- rep(NA_real_, length(x))
  ok <- x >= 0 & x < n
  if (any(ok)) {
    p <- x[ok] / n[ok]
    hr <- p / (ratio * (1 - p))
    out[ok] <- 1 - hr
  }
  out
}

futility_active <- if (!is.null(design_tte$testLower)) {
  tl <- design_tte$testLower
  if (length(tl) == 1) tl <- rep(tl, design_exact$k)
  as.logical(tl)
} else {
  design_exact$upper$bound <= design_exact$n.I
}
nominal_p_futility <- rep(NA_real_, design_exact$k)
nominal_p_futility[futility_active] <- stats::pbinom(
  q = design_exact$upper$bound[futility_active],
  size = design_exact$n.I[futility_active],
  prob = p0
)

tibble(
  Season = 1:3,
  `Spending time` = timing,
  `Planned total events` = design_exact$n.I,
  `Approx cumulative enrollment` = planned_cum_enrollment,
  `Exact efficacy bound (x <= a)` = design_exact$lower$bound,
  `VE at bound (efficacy)` = ve_from_bound(design_exact$lower$bound, design_exact$n.I, ratio),
  `Nominal 1-sided p at bound (efficacy)` = stats::pbinom(design_exact$lower$bound, design_exact$n.I, p0),
  `Exact futility bound (x >= b)` = ifelse(futility_active, design_exact$upper$bound, NA_integer_),
  `VE at bound (futility)` = ve_from_bound(
    ifelse(futility_active, design_exact$upper$bound, -1L),
    design_exact$n.I,
    ratio
  ),
  `Nominal 1-sided p at bound (futility)` = nominal_p_futility,
  `Target cumulative alpha spend` = target_alpha_spend,
  `Achieved cumulative alpha spend` = achieved_alpha_spend,
  `Cumulative power under H1` = achieved_power_h1
) |>
  gt() |>
  fmt_number(columns = 2, decimals = 3) |>
  fmt_percent(columns = c(`VE at bound (efficacy)`, `VE at bound (futility)`), decimals = 1) |>
  fmt_number(
    columns = c(
      `Nominal 1-sided p at bound (efficacy)`,
      `Nominal 1-sided p at bound (futility)`,
      `Target cumulative alpha spend`,
      `Achieved cumulative alpha spend`,
      `Cumulative power under H1`
    ),
    decimals = 4
  ) |>
  tab_spanner(
    label = "Efficacy",
    columns = c(
      `Exact efficacy bound (x <= a)`,
      `VE at bound (efficacy)`,
      `Nominal 1-sided p at bound (efficacy)`
    )
  ) |>
  tab_spanner(
    label = "Futility",
    columns = c(
      `Exact futility bound (x >= b)`,
      `VE at bound (futility)`,
      `Nominal 1-sided p at bound (futility)`
    )
  ) |>
  tab_header(
    title = "Planned exact binomial seasonal monitoring",
    subtitle = "Super-superiority example with Pocock-like efficacy spending"
  ) |>
  tab_footnote(
    footnote = "x denotes cumulative observed events in the experimental arm; efficacy is established when x is at or below the listed efficacy bound.",
    locations = cells_column_labels(columns = `Exact efficacy bound (x <= a)`)
  ) |>
  tab_footnote(
    footnote = "Futility is specified only at IA1; blank futility entries indicate no futility stopping boundary at that analysis.",
    locations = cells_column_labels(columns = `Exact futility bound (x >= b)`)
  )
```

The next table gives planned enrollment/sample size by season and overall.

```{r}
enrollment_table <- tibble(
  Season = as.character(1:3),
  `Control planned enrollment` = planned_enrollment_control,
  `Experimental planned enrollment` = planned_enrollment_experimental
) |>
  dplyr::mutate(
    `Total planned enrollment` = `Control planned enrollment` + `Experimental planned enrollment`,
    `Cumulative planned enrollment` = cumsum(`Total planned enrollment`)
  )

dplyr::bind_rows(
  enrollment_table,
  tibble(
    Season = "Overall",
    `Control planned enrollment` = sum(enrollment_table$`Control planned enrollment`),
    `Experimental planned enrollment` = sum(enrollment_table$`Experimental planned enrollment`),
    `Total planned enrollment` = sum(enrollment_table$`Total planned enrollment`),
    `Cumulative planned enrollment` = sum(enrollment_table$`Total planned enrollment`)
  )
) |>
  gt() |>
  tab_header(title = "Planned enrollment by season and overall")
```

## Example repeated and sequential p-values

Suppose observed event totals at the three seasonal analyses are equal to the
planned totals and observed experimental events are as below.
Here, `x` is the observed number of events in the experimental arm.
The offset sets season 2 to be one event above the efficacy bound to
demonstrate that the repeated p-value is greater than 0.025.

```{r}
x_offset_from_efficacy <- c(0L, 1L, 0L)
example_x <- pmax(0L, design_exact$lower$bound + x_offset_from_efficacy)
example_p <- repeatedPValueBinomialExact(
  gsD = design_tte,
  n.I = design_exact$n.I,
  x = example_x
)
example_p
```

The sequential p-value is the minimum repeated p-value:

```{r}
sequentialPValueBinomialExact(
  gsD = design_tte,
  n.I = design_exact$n.I,
  x = example_x
)
```

## Update exact bounds at analysis time

The official package path for updating exact-binomial bounds at analysis time is
to call `toBinomialExact()` with `observedEvents`.

The spending denominator remains the original planned final event count from the
design object, so earlier analyses are not changed when observed totals differ
from plan. At look `j`, spending time is based on
`observedEvents[j] / planned_final_events` (capped at 1). If a look reaches the
planned final event count, all remaining spending is used at that look.

If the final look is under-run but you still want to use full alpha at the last
analysis, set `maxSpend = TRUE`. This only changes the final spending time to 1
and leaves earlier looks unchanged.

For explicit control, you can pass `usTime` (and for `test.type = 4`, `lsTime`)
directly to `toBinomialExact()`, following the same spending-time conventions as
`gsDesign()` and `gsSurv()`.
As above, setting `x` one event above an updated efficacy bound at a look gives
a repeated p-value above 0.025 for that analysis.

```{r}
observed_counts_update <- c(
  planned_counts[1],
  planned_counts[2],
  max(planned_counts[2] + 1L, planned_counts[3] - 5L)
)
update_exact <- toBinomialExact(design_tte, observedEvents = observed_counts_update)
update_exact_full <- toBinomialExact(
  design_tte,
  observedEvents = observed_counts_update,
  maxSpend = TRUE
)

tibble(
  Analysis = seq_along(update_exact$n.I),
  `Observed total events` = update_exact$n.I,
  `Updated efficacy bound (x <= a), default spending` = update_exact$lower$bound,
  `Updated efficacy bound (x <= a), maxSpend=TRUE` = update_exact_full$lower$bound,
  `Updated futility bound, default spending` = update_exact$upper$bound,
  `Updated futility bound, maxSpend=TRUE` = update_exact_full$upper$bound
) |>
  gt() |>
  tab_header(title = "Updated exact bounds using observedEvents")
```

## Lightweight runnable simulation

We use the package helper `simBinomialSeasonalExact()` to run both fixed and
blinded-adaptive seasonal scenarios. In this vignette, we set
`final_full_spending = TRUE` to force full alpha spending at the final look
even when the final observed total event count is below plan.

```{r}
sim_light <- simBinomialSeasonalExact(
  gsD = design_tte,
  ve = c(`H0 (VE=30%)` = ve0, `H1 (VE=80%)` = ve1),
  nsim = c(150, 150),
  control_event_rate = c(0.003, 0.003),
  season_length = season_length_years,
  dropout_rate = dropout_6mo,
  planned_counts = planned_counts,
  adaptive = c(FALSE, TRUE),
  max_multiplier = 2,
  final_full_spending = TRUE,
  seed = 101
)
```

```{r}
oc <- sim_light$summary |>
  dplyr::mutate(
    Scenario = ifelse(adaptive, paste0("Adaptive: ", scenario), paste0("Fixed: ", scenario))
  ) |>
  dplyr::select(
    Scenario,
    `Efficacy crossing probability` = rejection_rate,
    `Futility stopping probability` = futility_stop_rate,
    `MC SE (efficacy)` = mc_se,
    `MC SE (futility)` = futility_mc_se,
    `Mean total events` = mean_total_events,
    `Mean total enrolled` = mean_total_enrolled,
    `Mean looks used` = mean_looks
  )

oc |>
  gt() |>
  fmt_number(columns = 2:5, decimals = 4) |>
  fmt_number(columns = 6:8, decimals = 2) |>
  tab_header(
    title = "Lightweight simulation results",
    subtitle = "Exact-binomial monitoring with seasonal analyses"
  ) |>
  tab_footnote(
    footnote = "For VE=30% scenarios, efficacy crossing probability is Type I error under the non-binding futility convention (futility crossings do not block later efficacy crossings).",
    locations = cells_column_labels(columns = `Efficacy crossing probability`)
  )
```

## Example with lower-than-planned event rates

To illustrate adaptation when events are lower than planned, we halve the
seasonal control event rate from 0.3% to 0.15% and compare fixed versus
adaptive monitoring for both `VE = 30%` and `VE = 80%`.

```{r}
sim_low <- simBinomialSeasonalExact(
  gsD = design_tte,
  ve = c(`H0 (VE=30%)` = ve0, `H1 (VE=80%)` = ve1),
  nsim = c(300, 300),
  control_event_rate = c(0.0015, 0.0015),
  season_length = season_length_years,
  dropout_rate = dropout_6mo,
  planned_counts = planned_counts,
  adaptive = c(FALSE, TRUE),
  max_multiplier = 2,
  final_full_spending = TRUE,
  seed = 505
)

low <- sim_low$summary
tibble(
  Scenario = c(
    "Without adaptation: Type I error (VE=30%)",
    "With adaptation: Type I error (VE=30%)",
    "Without adaptation: Power (VE=80%)",
    "With adaptation: Power (VE=80%)"
  ),
  `Efficacy crossing probability` = c(
    low$rejection_rate[!low$adaptive & low$scenario == "H0 (VE=30%)"],
    low$rejection_rate[low$adaptive & low$scenario == "H0 (VE=30%)"],
    low$rejection_rate[!low$adaptive & low$scenario == "H1 (VE=80%)"],
    low$rejection_rate[low$adaptive & low$scenario == "H1 (VE=80%)"]
  ),
  `Futility stopping probability` = c(
    low$futility_stop_rate[!low$adaptive & low$scenario == "H0 (VE=30%)"],
    low$futility_stop_rate[low$adaptive & low$scenario == "H0 (VE=30%)"],
    low$futility_stop_rate[!low$adaptive & low$scenario == "H1 (VE=80%)"],
    low$futility_stop_rate[low$adaptive & low$scenario == "H1 (VE=80%)"]
  ),
  `Mean total events` = c(
    low$mean_total_events[!low$adaptive & low$scenario == "H0 (VE=30%)"],
    low$mean_total_events[low$adaptive & low$scenario == "H0 (VE=30%)"],
    low$mean_total_events[!low$adaptive & low$scenario == "H1 (VE=80%)"],
    low$mean_total_events[low$adaptive & low$scenario == "H1 (VE=80%)"]
  ),
  `Mean total enrolled` = c(
    low$mean_total_enrolled[!low$adaptive & low$scenario == "H0 (VE=30%)"],
    low$mean_total_enrolled[low$adaptive & low$scenario == "H0 (VE=30%)"],
    low$mean_total_enrolled[!low$adaptive & low$scenario == "H1 (VE=80%)"],
    low$mean_total_enrolled[low$adaptive & low$scenario == "H1 (VE=80%)"]
  ),
  `Mean looks used` = c(
    low$mean_looks[!low$adaptive & low$scenario == "H0 (VE=30%)"],
    low$mean_looks[low$adaptive & low$scenario == "H0 (VE=30%)"],
    low$mean_looks[!low$adaptive & low$scenario == "H1 (VE=80%)"],
    low$mean_looks[low$adaptive & low$scenario == "H1 (VE=80%)"]
  )
) |>
  gt() |>
  fmt_number(columns = 2:3, decimals = 4) |>
  fmt_number(columns = 4:6, decimals = 2) |>
  tab_header(
    title = "Lower-than-planned event rate illustration",
    subtitle = "Adaptive approach increases enrollment to recover information"
  ) |>
  tab_footnote(
    footnote = "Type I error rows use non-binding futility for efficacy crossing probability; futility stopping probability is shown separately.",
    locations = cells_column_labels(columns = `Efficacy crossing probability`)
  )
```

This table provides side-by-side comparisons of Type I error, power, and
futility stopping probability without and with adaptation under the
lower-than-planned event-rate scenario.

## Larger offline runs (template)

The chunk below is intentionally not executed in package builds.

```{r, eval=FALSE}
# Suggested offline settings
type1_nsim <- 20000
power_nsim <- 3500

sim_type1_big <- simBinomialSeasonalExact(
  gsD = design_tte,
  ve = c(`H0 (VE=30%)` = ve0),
  nsim = type1_nsim,
  control_event_rate = 0.003,
  season_length = season_length_years,
  dropout_rate = dropout_6mo,
  planned_counts = planned_counts,
  adaptive = c(FALSE, TRUE),
  final_full_spending = TRUE,
  seed = 5001
)

sim_power_big <- simBinomialSeasonalExact(
  gsD = design_tte,
  ve = c(`H1 (VE=80%)` = ve1),
  nsim = power_nsim,
  control_event_rate = 0.003,
  season_length = season_length_years,
  dropout_rate = dropout_6mo,
  planned_counts = planned_counts,
  adaptive = c(FALSE, TRUE),
  final_full_spending = TRUE,
  seed = 6001
)
```

## Notes and extensions

- The adaptive pathway shown here is blinded because it uses total event counts
  only; treatment-group differences are not used to update enrollment.
- Spending fractions are kept fixed at `1/3, 2/3, 1`.
- For modified intention-to-treat analyses, additional exclusion/dropout
  mechanisms can be layered into the simulation by reducing at-risk counts
  before each seasonal event/dropout draw.
- Simple superiority and non-inferiority variants can be studied by changing
  `hr0`/`hr1` and rebuilding the same workflow.

## References
