---
title: "One-Sample Binomial Problems"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: gsDesign.bib
vignette: >
  %\VignetteIndexEntry{One-Sample Binomial Problems}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

options( width = 58 )

```

# Introduction

This article/vignette provides an overview of functions considering designs for one-sample binomial outcomes.
All routines use exact calculations for sample size and power.
The underlying probability model is that we sample independent binary (Bernoulli) observations with probability of a successful outcome (hereafter referred to as response) equal to $p$.
The general design problem is rule out a null hypothesis $H_0: p=p_0$ versus an alternative $H_1: p=p_1$ where $0< p_0 < p_1< 1$.

- `nBinomial1Sample()`: compute power or sample size for fixed (non-group sequential) design
- `gsBinomialExact()`: compute boundary crossing probabilities for a group sequential design
- `binomialSPRT()`: derive a design using the sequential probability ratio test (SPRT)




# Fixed Sample Size

Suppose we have a random variable $X\sim\hbox{Binomial}(n,p)$ where $n$ is the number of independent observations of independent binary random variables with probability of a response equal to $p$.
Consider two response rates $p_0 < p_1$. Without loss of generality, we can assume the null hypothesis response rate is $p_0$ and the alternate hypothesis response rate is $p_1$.
We set a sample size $n$ and a critical value $b$ such that if $X\ge b$ then we reject the null hypothesis $H_0:p=p_0$. 
Type I error is by definition

$$\alpha=P(X\ge b|p=p_0)=\sum_{m=b}^n\binom{n}{m}p_0^m(1-p_0)^{n-m}.$$

Power is one minus the Type II error

$$1 - \beta= P(X\ge b|p=p_1)=\sum_{m=b}^n\binom{n}{m}p_1^m(1-p_1)^{n-m}.$$

For any given $n,b,p_0,p_1$ it is straightforward to compute these quantities. 
The critical value $b$ for a given sample size $n$ can be computed as the inverse of the binomial distribution.
For example, assuming $n=100$, $p_0=0.15$ and $\alpha=0.025$, we can compute $b$ as follows:

```{r}
b <- qbinom(p = .025, size = 100, prob = .15, lower.tail = FALSE)
b
```

Due to the discrete nature of the binomial distribtuion the actual Type I error is less than or equal to the specified 0.025. 


```{r}
alpha <- pbinom(q = b, size = 100, prob = .15, lower.tail = FALSE)
alpha
```

We note that with one fewer successes, the bound would have great than the desired 0.025 Type I error:

```{r}
pbinom(q = b-1, size = 100, prob = .15, lower.tail = FALSE)
```

Now we can compute power assuming a given $p_1>p_0$, say $p_1= 0.25$.

```{r}
pbinom(q = b, size = 100, prob = .25, lower.tail = FALSE)
```

A natural question now would be what $n$ is required for a test of $H_0: p=p_0=0.15$ against $H_1: p=p_1=0.25$ if we desire one-sided Type I error $\alpha=0.025$ and $90\%$ power. This can be computed with the function `nBinomial1Sample()` as follows:

```{r}
library(gsDesign)
nBinomial1Sample(p0 = 0.15, p1 = .25, alpha = .025, beta = .10, n=100:250, conservative = TRUE)
```
We note that the `beta` argument is 1 minus the targeted power, or 0.10 in this case.
We can check this as above, or more simply by computing power for different sample sizes:

```{r}
nBinomial1Sample(p0 = 0.15, p1 = .25, alpha = .025, n=173:182)
```

We note in this table that Type II error `beta` is below the targeted 0.10 if $n=173$, but not for $n=174,175,179$. 
If we had not used the `conservative=TRUE` argument above, the sample size selected would be:

```{r}
nBinomial1Sample(p0 = 0.15, p1 = .25, alpha = .025, beta = .10, n=100:250, conservative = FALSE)
```

Generally, given the possibility of overrunning on sample size, it may be good to use the `conservative=TRUE` option.

The three columns at the right side of the above table are:

`b`: The critical value for the number of responses out of `n` which will reject $H_0$.
`alphaR`: The realized (exact) Type I error realized; this will be less than or equal to the input `alpha` in all cases.
`Power`: The probability of crossing the upper bound $P(X\ge b|p=p_1).$

