---
title: "Reproducing PROC SEQDESIGN survival designs in gsDesign"
output: rmarkdown::html_vignette
bibliography: "gsDesign.bib"
vignette: >
  %\VignetteIndexEntry{Reproducing PROC SEQDESIGN survival designs in gsDesign}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(gsDesign)
```

## Overview

This vignette provides a thorough analysis of the differences between
the SAS SEQDESIGN procedure and the R package gsDesign for designing
group sequential clinical trials with time-to-event endpoints.

### Starting point: SAS PROC SEQDESIGN Example 109.14

Consider the example described in the SAS Documentation:
[Computing Sample Size for Survival Data with Uniform Accrual](https://documentation.sas.com/doc/en/statug/15.2/statug_seqdesign_examples14.htm).
The SAS code is as follows:

```sas
proc seqdesign;
    ErrorSpend: design nstages=4 method=errfuncobf;
    samplesize model=twosamplesurvival(
        nullhazard = 0.03466
        hazard     = 0.01733
        accrual    = uniform
        accrate    = 15
        acctime    = 18
        ceiling    = time);
run;
```

**SAS assumptions summary:**

- 4-stage group sequential design
- Lan-DeMets O'Brien-Fleming error spending function
- Two-sided symmetric test, total alpha = 0.05
- 90% power (beta = 0.10)
- Equally spaced information fractions: 0.25, 0.50, 0.75, 1.00
- Schoenfeld formula for required number of events
- Sample size formula: $N = \text{ACCRATE} \times \text{ACCTIME} = 15 \times 18 = 270$ (fixed)
- Study duration $T$ solved to achieve required events (CEILING=TIME)
- Hazard ratio $HR = 0.01733 / 0.03466 = 0.5$

Based on these assumptions, a user naively attempting to replicate this in R
with gsDesign may attempt code like the following:

```{r naive-r-attempt}
des <- gsSurv(
  k         = 4,
  test.type = 2,
  alpha     = 0.05,
  beta      = 0.10,
  sfu       = sfLDOF, # Lan-DeMets O'Brien--Fleming
  timing    = c(.25, .50, .75, 1), # Equal information fractions
  lambdaC   = 0.03466,
  hr        = 0.5,
  eta       = 0, # No dropout
  gamma     = 15, # Uniform accrual rate of 15/month
  R         = 18, # Accrual duration (months)
  T         = 25,
  ratio     = 1 # 1:1 randomization
)

des
```

Note that the output differs in multiple ways, including that the gsDesign
output indicates that fewer events will be required to achieve the desired
level of power compared to the SAS output. Such differences do not cast doubt
on the fidelity of either software, but rather highlight the underlying
differences in assumptions and defaults settings in the design of each
software tool.

### Side-by-side comparison table

```{r comparison-table}
events_sas <- c(22.27, 44.54, 66.81, 89.08)
events_a2 <- round(des$n.I, 2)

z_sas <- c(4.333, 2.963, 2.359, 2.014)
z_a2 <- round(des$upper$bound, 4)

N2a <- sum(des$eNC[4, ]) + sum(des$eNE[4, ])

knitr::kable(
  data.frame(
    Analysis = 1:4,
    Events_SAS = events_sas,
    Events_gsDesign = events_a2,
    Z_SAS = z_sas,
    Z_gsDesign = z_a2
  ),
  caption = "Naive side-by-side comparison of events and Z-boundaries at each look."
)
```

The rest of this vignette works to resolve the apparent discrepancies in
results, using the above example as a reference case. We identify the key
assumptions in each system, explain why naive parameter translation produces
different results, and show how to obtain matching output.
Those who would like to know more details about the design of the time-to-event
functionality in gsDesign should consult `vignette("SurvivalOverview")`.

## Key differences: SAS SEQDESIGN vs. R gsDesign

There are three fundamental differences that affect the output from the example above.

### 1. Events formula

- **SAS:** @Schoenfeld1981. Uses only the null-hypothesis variance.
- **gsDesign:** @LachinFoulkes by default.
  Uses both null and alternative hypothesis variances. L-F is slightly more
  conservative, giving \~1-2% more events than Schoenfeld for the same parameters.

### 2. Alpha handling in `gsSurv()`

O'Brien-Fleming spending makes things a bit ambiguous due to the one-sided
versus two-sided spending.

- **SAS:** Uses a symmetric two-sided test with alpha = 0.05.
  Per-tail, $z_{\alpha/2} = z_{0.025} = 1.96$.
  Boundaries and event calculations both use 0.025 per tail.
- **gsSurv():** The `test.type = 2` with `alpha = 0.05` creates the same
  symmetric boundaries, but computes a different group sequential design
  inflation factor because it counts power from both tails.
  SAS computes the inflation factor using one-sided power only.

A possible strategy for aligning `gsSurv()` and SAS calculations in this case
is to use `test.type = 1` with `alpha = 0.025` to match SAS per-tail logic,
as we will see momentarily.

### 3. What varies to achieve power

- **SAS:** Fixes the accrual rate and accrual time, solves for study duration.
- **gsSurv():** Fixes study duration (`T`) and varies the accrual rate (`gamma`)
  and accrual time (`R`).

## Aligning the two approaches

### `gsSurv()` with aligned parameters

Let's start our work with gsDesign by defining parameters to match SAS Example 109.14:

```{r commonparm, message=FALSE}
k <- 4
alpha <- 0.05 # Two-sided total alpha (SAS convention)
beta <- 0.10 # 1 - power = 0.10 -> 90% power
lambdaC <- 0.03466 # Control hazard rate
lambdaE <- 0.01733 # Experimental hazard rate
HR <- lambdaE / lambdaC # = 0.5
timing <- c(0.25, 0.50, 0.75, 1.00) # Equally spaced information fractions
accrate <- 15 # Uniform accrual rate (subjects per time unit)
R <- 18 # Accrual duration (time units)
N <- accrate * R
```

Instead of starting with a call to `gsDesign::gsDesign()`, we begin with
`gsDesign::gsSurv()`. The `gsSurv()` function combines `nSurv()` with
`gsDesign()` (group sequential boundaries) in one call.
It is the standard gsDesign function for designing time-to-event trials.

Two key differences from SAS remain even with the alpha parameter aligned with SAS:

- `gsSurv()` defaults to Lachin-Foulkes sample size; Lachin-Foulkes gives
  slightly more events than Schoenfeld.
- `gsSurv()` with $T$ specified **varies** the accrual rate to achieve power,
  rather than fixing $N$ and solving for $T$ as SAS does.
  To address this, we set the `T` parameter equal to the study duration time
  calculated by SAS.

```{r approach2}
des_2 <- gsSurv(
  k = 4,
  test.type = 1, # One-sided (matches SAS per-tail)
  alpha = 0.025, # alpha/2 per tail
  beta = 0.10,
  sfu = sfLDOF,
  timing = c(.25, .50, .75, 1),
  lambdaC = 0.03466,
  hr = 0.5,
  eta = 0, # Assume no dropout
  gamma = 15, # Input rate; gsSurv will **adjust** this
  R = 18,
  T = 25.13323, # This is the study duration computed by SAS
  ratio = 1,
  method = "Schoenfeld" # Use Schoenfeld bounds to match SAS
)
des_2
gsBoundSummary(des_2)
```

**Observations:**

Using `gsSurv()` with parameters that align with SAS settings, we notice that

- $Z$-boundaries: 4.333, 2.963, 2.359, 2.014 -> matches SAS
- Events: 23, 45, 67, 90 -> matches SAS; gsDesign is just taking the `ceiling()`
  of the SAS output.
- Total N: 280 (slightly more than SAS's fixed 270). This is a result of the
  fact that `gsSurv()` is varying `R` and `gamma` (corresponding conceptually
  to the parameters `acctime` and `accrate` in PROC SEQDESIGN) in order to
  maintain power.

### Bringing it full circle with SAS PROC SEQDESIGN

Observe the final values for `R` and `gamma` calculated by `gsSurv()` in the
"Key inputs" table output (see above) by `des_2`: `gamma = 14.568`, and
`R = 19.133` (slightly different than our input 15 and 18).
If we substitute in these values for the `accrate` and `acctime` parameters
in SAS, we will see that the SAS calculations produce a final sample size of
`N = 278.7`, aligning with the `gsSurv()` sample size up.

``` sas
proc seqdesign;
    ErrorSpend: design nstages=4 method=errfuncobf;
    samplesize model=twosamplesurvival(
        nullhazard = 0.03466
        hazard     = 0.01733
        accrual    = uniform
        accrate    = 14.568
        acctime    = 19.133
        ceiling    = time);
run;
```

## References
