---
title: "Reproducing PROC SEQDESIGN survival designs in gsDesign"
output: rmarkdown::html_vignette
bibliography: "gsDesign.bib"
vignette: >
  %\VignetteIndexEntry{Reproducing PROC SEQDESIGN survival designs in gsDesign}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(gsDesign)
```

## Overview

This vignette provides a thorough analysis of the differences between
the SAS SEQDESIGN procedure and the R package gsDesign for designing
group sequential clinical trials with time-to-event endpoints.

### Starting point: SAS PROC SEQDESIGN survival example

Consider the example described in the SAS Documentation:
[Computing Sample Size for Survival Data with Uniform Accrual](https://support.sas.com/documentation/cdl/en/statug/68162/HTML/default/statug_seqdesign_examples14.htm).
The first PROC SEQDESIGN call in that example does not specify `ACCTIME`; SAS
therefore derives a range of possible accrual times. The comparison below uses
the subsequent SAS call with `ACCTIME=18`, which fixes the maximum sample size
at `15 * 18 = 270` subjects and asks SAS to solve the follow-up time.

```sas
proc seqdesign;
    ErrorSpend: design nstages=4 method=errfuncobf
                ;
    samplesize model(ceiladjdesign=include)=twosamplesurvival(
        nullhazard = 0.03466
        hazard     = 0.01733
        accrual    = uniform
        accrate    = 15
        acctime    = 18);
run;
```

**SAS assumptions summary:**

- 4-stage group sequential design
- Lan-DeMets O'Brien-Fleming error spending function
- Two-sided symmetric test, total alpha = 0.05
- 90% power (beta = 0.10)
- Equally spaced information fractions: 0.25, 0.50, 0.75, 1.00
- Schoenfeld formula for required number of events
- Fixed accrual rate and duration, so the maximum sample size is
  $N = \text{ACCRATE} \times \text{ACCTIME} = 15 \times 18 = 270$
- Study duration $T$ solved to achieve the required events
- Hazard ratio $HR = 0.01733 / 0.03466 = 0.5$

SAS does produce a sample size in this example. In the "Sample Size Summary"
table for the `ACCTIME=18` case, it reports `Max Sample Size = 270`,
`Expected Sample Size (Null Ref) = 269.9206`,
`Expected Sample Size (Alt Ref) = 263.1141`,
`Follow-up Time = 7.133226`, and `Total Time = 25.13323`. The maximum sample
size is not calculated from the event formula; it is implied directly by the
fixed accrual rate and accrual duration:
`15 subjects/time unit * 18 time units = 270 subjects`.

SAS reports two sets of analysis quantities. We focus first on the
fractional-time design, where the analysis times are not rounded, and then
return to the `CEILING=TIME` adjusted design later.

```{r sas-reported-values}
sas_fractional <- data.frame(
  Analysis = 1:4,
  Events = c(22.26962, 44.53924, 66.80886, 89.07847),
  Calendar_time = c(11.2631, 16.2875, 20.4926, 25.13323),
  N = c(168.95, 244.31, 270.00, 270.00),
  Upper_Z = c(4.33263, 2.96333, 2.35902, 2.01409)
)

sas_ceiling_time <- data.frame(
  Analysis = 1:4,
  Events = c(25.11225, 48.22068, 69.38712, 92.92776),
  Calendar_time = c(12, 17, 21, 26),
  N = c(180, 255, 270, 270),
  Upper_Z = c(4.15591, 2.90189, 2.36973, 2.01362)
)
```

gsDesign uses one-sided alpha by default, so we need to convert the two-sided SAS alpha to a one-sided alpha.
The two-sided alpha is 0.05, so the one-sided alpha is 0.025.

```{r aligned-r-attempt}
des <- gsSurv(
  k         = 4,
  test.type = 2,
  alpha     = 0.025,
  beta      = 0.10,
  sfu       = sfLDOF, # Lan-DeMets O'Brien--Fleming
  timing    = c(.25, .50, .75, 1), # Equal information fractions
  lambdaC   = 0.03466,
  hr        = 0.5,
  eta       = 0, # No dropout
  gamma     = 15, # Uniform accrual rate of 15/month
  R         = 18, # Accrual duration (months)
  T         = NULL,
  minfup    = NULL,
  ratio     = 1, # 1:1 randomization
  method    = "Schoenfeld"
)

des |> gsBoundSummary(tdigits = 2, ddigits = 2) |> gt::gt()
```

This call now matches the SAS fractional-time design. The key inputs are
`alpha = 0.025`, which represents the one-sided alpha corresponding to SAS's
two-sided total alpha of 0.05; `method = "Schoenfeld"`, which matches the SAS
event formula; and `T = NULL, minfup = NULL`, which fixes the accrual duration
and solves the follow-up time.

### Side-by-side comparison table

```{r comparison-table}
events_a2 <- round(des$n.I, 2)

z_a2 <- round(des$upper$bound, 4)

N2a <- sum(des$eNC[4, ]) + sum(des$eNE[4, ])

knitr::kable(
  data.frame(
    Analysis = 1:4,
    Events_SAS = round(sas_fractional$Events, 2),
    Events_gsDesign = events_a2,
    Z_SAS = round(sas_fractional$Upper_Z, 4),
    Z_gsDesign = z_a2
  ),
  caption = "Side-by-side comparison of events and Z-boundaries at each look."
)
```

The rest of this vignette identifies the key assumptions in each system and
explains why alternative parameter translations can produce different results.
Those who would like to know more details about the design of the time-to-event
functionality in gsDesign should consult `vignette("SurvivalOverview")`.

## Key differences: SAS SEQDESIGN vs. R gsDesign

There are four practical translation points that affect the output from the
example above.

### 1. Event formula

- **SAS:** @Schoenfeld1981. Uses only the null-hypothesis variance.
- **gsDesign:** @LachinFoulkes by default.
  Uses both null and alternative hypothesis variances. L-F is slightly more
  conservative, giving \~1-2% more events than Schoenfeld for the same parameters.
  Use `method = "Schoenfeld"` to match the SAS event formula.

### 2. Alpha handling in `gsDesign()` and `gsSurv()`

`gsDesign()` stores and spends the **one-sided** Type I error. Thus, for a
symmetric two-sided design (`test.type = 2`), `alpha = 0.025` means 0.025 in
each tail, or 0.05 total two-sided Type I error.

This convention is deliberate: most group sequential computations in
`gsDesign()` are organized around the upper-bound crossing probability. A
symmetric two-sided design is represented by mirroring that upper-bound
spending in the lower tail. Thus the `alpha` argument is still the per-tail
crossing probability, even though the design has both an upper and lower
efficacy boundary.

`gsSurv()` follows the same convention internally. If a user prefers to enter
the total two-sided alpha, `alpha = 0.05, sided = 2` is equivalent to
`alpha = 0.025, sided = 1` for this example because `gsSurv()` passes
`alpha / sided` to `gsDesign()`.

Here we use `test.type = 2` and `alpha = 0.025` to make the symmetric
two-sided `gsDesign()` object match the SAS two-sided total alpha of 0.05.

### 3. Accrual duration and follow-up time

With `ACCTIME=18`, SAS fixes the accrual duration and total maximum sample size
and solves for the additional follow-up time. To match this in `gsSurv()`, set
both `T = NULL` and `minfup = NULL`. This tells `gsSurv()` to keep the input
accrual rate and accrual duration fixed, then solve the follow-up duration
needed for the final group sequential event requirement.

This is a common source of apparent disagreement. For a fixed-duration
survival design, `T` is the total study duration and `minfup` is the minimum
follow-up after enrollment closes. With scalar accrual, specifying both `T`
and `minfup` makes the implied accrual duration `T - minfup`; specifying only
one of them can therefore change which quantity `gsSurv()` solves for. The SAS
comparison fixes accrual at 18 time units and lets the total time be derived.

### 4. Fractional time vs. ceiling time

The original SAS fractional-time design has equal information fractions and
final events 89.07847. The SAS ceiling-time adjusted design rounds analysis
times to 12, 17, 21, and 26. These rounded calendar times imply unequal
information fractions and final events 92.92776. Both are valid SAS outputs;
they should not be mixed when comparing against gsDesign.

The distinction matters because SAS's fractional-time table answers "what
analysis times give the target information fractions?", while the ceiling-time
table answers "what happens after those analysis times are rounded up to whole
time units?" Once calendar times are rounded up, subjects are followed longer,
more events are expected, and the O'Brien-Fleming spending times are no longer
exactly 0.25, 0.50, 0.75, and 1.00.

The translation used below is summarized as follows:

```{r translation-table, echo=FALSE}
knitr::kable(
  data.frame(
    Quantity = c(
      "Two-sided Type I error",
      "Symmetric two-sided design",
      "Event formula",
      "Accrual duration",
      "Total study duration",
      "Follow-up after accrual"
    ),
    SAS = c(
      "alpha = 0.05 total",
      "Early stop to reject either side",
      "Schoenfeld log-rank information",
      "ACCTIME = 18",
      "Total Time = 25.13323",
      "Derived as 7.133226"
    ),
    gsDesign = c(
      "alpha = 0.025 per tail",
      "test.type = 2",
      "method = \"Schoenfeld\"",
      "R = 18",
      "T = NULL",
      "minfup = NULL"
    ),
    Reason = c(
      "gsDesign stores and spends one-sided alpha",
      "Mirrors the upper and lower efficacy boundaries",
      "Avoids Lachin-Foulkes default event calculation",
      "Keeps the same fixed accrual duration",
      "Lets gsSurv() solve total time from fixed accrual",
      "Lets gsSurv() solve the follow-up duration"
    ),
    check.names = FALSE
  ),
  caption = "Translation from the SAS PROC SEQDESIGN example to gsDesign inputs."
)
```

## Aligning the two approaches

### `gsSurv()` with aligned parameters

Let's start our work with gsDesign by defining parameters to match the SAS
fractional-time design:

```{r commonparm, message=FALSE}
k <- 4
alpha_sas <- 0.05 # Two-sided total alpha (SAS convention)
alpha_gsdesign <- alpha_sas / 2 # gsDesign uses one-sided alpha
beta <- 0.10 # 1 - power = 0.10 -> 90% power
lambdaC <- 0.03466 # Control hazard rate
lambdaE <- 0.01733 # Experimental hazard rate
HR <- lambdaE / lambdaC # = 0.5
timing <- c(0.25, 0.50, 0.75, 1.00) # Equally spaced information fractions
accrate <- 15 # Uniform accrual rate (subjects per time unit)
accrual_duration <- 18 # Accrual duration (time units)
sas_total_time <- 25.13323
sas_followup_time <- sas_total_time - accrual_duration
N <- accrate * accrual_duration
```

Instead of starting with a call to `gsDesign::gsDesign()`, we begin with
`gsDesign::gsSurv()`. The `gsSurv()` function combines `nSurv()` with
`gsDesign()` (group sequential boundaries) in one call.
It is the standard gsDesign function for designing time-to-event trials.

The call below uses the same two-sided symmetric structure as SAS:

- `test.type = 2` for symmetric two-sided boundaries;
- `alpha = 0.025`, the one-sided alpha corresponding to SAS's total
  two-sided alpha of 0.05;
- `method = "Schoenfeld"` for the SAS event formula;
- `T = NULL` and `minfup = NULL` so the accrual rate and accrual duration
  remain fixed while `gsSurv()` solves the follow-up duration.

```{r approach2}
des_2 <- gsSurv(
  k = 4,
  test.type = 2, # Symmetric two-sided design
  alpha = alpha_gsdesign, # One-sided alpha; SAS total alpha is 2 * this
  beta = 0.10,
  sfu = sfLDOF,
  timing = c(.25, .50, .75, 1),
  lambdaC = 0.03466,
  hr = 0.5,
  eta = 0, # Assume no dropout
  gamma = accrate,
  R = accrual_duration,
  T = NULL,
  minfup = NULL,
  ratio = 1,
  method = "Schoenfeld"
)
des_2
gsBoundSummary(des_2)
```

**Observations:**

Using `gsSurv()` with parameters that align with SAS settings, the
fractional-time output agrees with SAS. The final follow-up duration is
`r round(des_2$minfup, 5)`, giving total study duration
`r round(des_2$T[des_2$k], 5)`.

```{r fractional-time-match}
gs_fractional <- data.frame(
  Analysis = 1:k,
  Events_SAS = sas_fractional$Events,
  Events_gsDesign = des_2$n.I,
  Time_SAS = sas_fractional$Calendar_time,
  Time_gsDesign = des_2$T,
  N_SAS = sas_fractional$N,
  N_gsDesign = rowSums(des_2$eNC) + rowSums(des_2$eNE),
  Z_SAS = sas_fractional$Upper_Z,
  Z_gsDesign = des_2$upper$bound
)

knitr::kable(
  round(gs_fractional, 5),
  caption = "Fractional-time SAS output compared with gsSurv()."
)
```

The final event count is 89.07847 in both systems. The `print()` and
`gsBoundSummary()` methods display rounded integer sample sizes and events,
but the object retains the fractional values shown above.

If the SAS follow-up time is already known, the equivalent fixed-follow-up
translation is `T = NULL` with `minfup` set to the SAS follow-up time. In that
case, `gsSurv()` holds the follow-up duration fixed and solves the accrual
duration needed for the final group sequential event target:

```{r fixed-followup-check}
des_fixed_followup <- gsSurv(
  k = 4,
  test.type = 2,
  alpha = alpha_gsdesign,
  beta = 0.10,
  sfu = sfLDOF,
  timing = c(.25, .50, .75, 1),
  lambdaC = 0.03466,
  hr = 0.5,
  eta = 0,
  gamma = accrate,
  R = accrual_duration,
  T = NULL,
  minfup = sas_followup_time,
  ratio = 1,
  method = "Schoenfeld"
)

fixed_followup_check <- data.frame(
  Quantity = c(
    "Final study duration",
    "Accrual duration",
    "Final events",
    "Final N"
  ),
  Solved_followup = c(
    des_2$T[k],
    sum(des_2$R),
    des_2$n.I[k],
    sum(des_2$eNC[k, ] + des_2$eNE[k, ])
  ),
  Specified_followup = c(
    des_fixed_followup$T[k],
    sum(des_fixed_followup$R),
    des_fixed_followup$n.I[k],
    sum(des_fixed_followup$eNC[k, ] + des_fixed_followup$eNE[k, ])
  )
)
fixed_followup_check[-1] <- lapply(fixed_followup_check[-1], round, digits = 5)

knitr::kable(
  fixed_followup_check,
  caption = "Both fixed-accrual translations produce the same final design."
)
```

### Matching the SAS ceiling-time adjusted design

The SAS example also reports a `CEILING=TIME` adjusted design. This is not the
same design as the equal-information fractional-time output above. The analysis
times are rounded up to 12, 17, 21, and 26, which increases the expected events
and changes the information fractions.

We can reproduce the ceiling-time event counts directly from the fixed accrual
rate, fixed accrual duration, and exponential event rates. With equal
randomization, the control and experimental accrual rates are each 7.5 per time
unit.

```{r ceiling-time-match}
event_count_at_time <- function(time) {
  control_events <- eEvents(
    lambda = lambdaC,
    gamma = accrate / (1 + 1),
    R = accrual_duration,
    T = time
  )$d
  experimental_events <- eEvents(
    lambda = lambdaE,
    gamma = accrate / (1 + 1),
    R = accrual_duration,
    T = time
  )$d
  sum(control_events) + sum(experimental_events)
}

ceiling_times <- ceiling(sas_fractional$Calendar_time)
ceiling_events <- vapply(ceiling_times, event_count_at_time, numeric(1))

des_ceiling <- gsDesign(
  k = k,
  test.type = 2,
  alpha = alpha_gsdesign,
  beta = beta,
  sfu = sfLDOF,
  n.I = ceiling_events,
  timing = ceiling_events / max(ceiling_events)
)

gs_ceiling <- data.frame(
  Analysis = 1:k,
  Events_SAS = sas_ceiling_time$Events,
  Events_gsDesign = ceiling_events,
  Time_SAS = sas_ceiling_time$Calendar_time,
  Time_gsDesign = ceiling_times,
  N_SAS = sas_ceiling_time$N,
  N_gsDesign = pmin(accrate * ceiling_times, N),
  Z_SAS = sas_ceiling_time$Upper_Z,
  Z_gsDesign = des_ceiling$upper$bound
)

knitr::kable(
  round(gs_ceiling, 5),
  caption = "Ceiling-time SAS output compared with gsDesign()."
)
```

The remaining small differences in the printed Z-values are numerical rounding
differences. The important conceptual distinction is that the SAS
fractional-time and ceiling-time tables are different designs: one uses equal
information fractions, and the other uses rounded calendar analysis times.

## References
