---
title: "Spending Time"
author: "Keaven M. Anderson"
date: "`r Sys.Date()`"
output:
    rmarkdown::html_vignette:
      fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Spending time concept and example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(gsDesign)
```

# Introduction

This vignette provides examples demonstrating uses of spending time, a concept introduced in version 3.0-4. 
One simple example of this is by plugging a calendar time fraction \cite{LanDeMets1989} into an $\alpha$-spending function rather than an information fraction.
However, the concept is more general than this and may apply to other situations that will be demonstrated below.
The situations we will consider are:

* Fixed spending at each interim analysis as demonstrated with an approximation to a Haybittle-Peto bound.
* Spending time based on information fraction from a different population analyzed for the same endpoint where the final event count for the given hypothesis is unknown.
* These examples will also be used to demonstrate how to deal with interim analysis event counts during the course that differ substantially from planned. 


# Calendar Time

# Haybittle-Peto Bound Approximation

# Unknown Final Event Count

```{r}
# code here applies to both fixed and group sequential design sections below.
# name of primary endpoint and time units
ep <- "overall survival"
timename <- "months"
timename1 <- "month"
# enrollment period durations; the last of these will be extended if T=NULL below
R <- c(1,2,3,4)
# relative enrollment rates during above periods
gamma<-c(1,1.5,2.5,4)
# study duration
# T can be set to NULL if you want to 
# fix enrollment and vary study duration
T <- 36 
# follow-up duration of last patient enrolled
minfup <- 12
# randomization ratio, experimental/control
ratio <- 1
# median control time-to-event
median <- 12
# exponential dropout rate per unit of time
eta <- .001
# hypothesized experimental/control hazard ratio
hr <- .75
# null hazard ratio (1 for superiority, >1 for non-inferiority)
hr0 <- 1
# Type I error (1-sided)
alpha <-.025
# Type II error (1-power)
beta<-.2
# number of analyses (interim + final)
k <- 2
# timing of interim analyses (k-1 increasing numbers >0 and <1)
timing <- .7
# efficacy bound spending function
sfu <- sfLDOF 
# efficacy bound spending parameter specification
sfupar <- NULL
# futility bound spending function
sfl <- sfHSD 
# futility bound spending parameter specification
sflpar <- 2
# generate design
x <- gsSurv(k=k,timing=timing,R=R,gamma=gamma,eta=eta,
            minfup=minfup,T=T,lambdaC=log(2)/median,
            hr=hr,hr0=hr0,beta=beta,alpha=alpha,
            sfu=sfu,sfupar=sfupar,sfl=sfl,sflpar=sflpar)
```


For a comparative trial we consider a 2-arm group sequential design with `r ep` as the primary endpoint as shown in Table \ref{tab1} (a second version of the table is in \ref{tab2}.
Timing, number of events, sample size, boundaries (Z-values, nominal p-values, approximate hazard ratios) are shown as well as the probability of crossing study boundaries under the null and alternate hypotheses.
Bounds are determined by an O'Brien-Fleming spending function for superiority ($\alpha$-spending) and a Hwang-Shih-DeCani spending function for futility with $\gamma=$ `r sflpar` ($\beta$-spending).
The median time-to-event is assumed to be `r median` `r timename` in the control group.
`r if (hr0==1) paste("The trial is designed to demonstrate superiority of experimental treatment over control with an assumed hazard ratio of ",hr,".",sep="")`
The total sample size is `r ceiling(sum(x$eNE[k,]+x$eNC[k,]))` and a
total of `r ceiling(sum(x$eDE[k,]+x$eDC[k,]))` endpoints is required for the final analysis.
Planned recruitment duration is `r sum(x$R)` \Sexpr{timename} and the minimum follow-up planned is `r round(x$minfup,1)` `r timename`.
Thus, the total expected study duration is `r round(max(x$T),1)` `r timename`.
The assumed dropout rate is `r 100*eta`\% per `r timename1`.
`r if (x$k==2) paste("There is a single interim analysis planned after",ceiling(x$eDE[1,]+x$eDC[1,]),"events have accrued which is expected after approximately",round(x$T[1],1),timename,sep=" ")`.


```{r}
knitr::kable(gsBoundSummary(x),digits=4,row.names=FALSE,
             caption="Using kable to print a design summary.")
```



Following are plots of the Z-values and approximate hazard ratios at the design bounds.


```{r,fig.width=6,fig.height=4}
plot(x,cex=.8,xlab="Number of events")+
  theme(legend.position = "bottom")
```


```{r,fig.width=6,fig.height=4}
plot(x,plottype="hr",cex=.8,xlab="Number of events")+
  theme(legend.position = "bottom")
```