---
title: "Design summary functions for group sequential designs"
output:
  html_document:
    df_print: paged
---

There are various methods of summarizing group sequential designs created by the **gsDesign** package.
We summarize these with examples here.
We begin by loading the **gsDesign** package and deriving a group sequential design in a `gsDesign` object.


```{r, results='asis',warning=FALSE,message=FALSE}
library(gsDesign)
d <- gsDesign(n.fix=100)
class(d)
```

We can produce a simple text summary of this design as follows:

```{r}
summary(d)
```

The print function for a `gsDesign` object is not pretty, but does provide relatively detailed information on a design.

```{r}
print(d)
```



```{r}
x <- gsBoundSummary(d)
```


We begin by loading the **gsDesign** package and creating a `gsBoundSummary` object that is no different in structure from a `data.frame` object, but has had printing functionality slightly customized. 





With version 3.1.1 of gsDesign, we will consider deprecating or eliminating the dependency on the **xtable** package. 
This dependency was only there for convenience in using **xtable** without having to remember to reset defaults in the `xtable::print.xtable()` function.
We demonstrate how to do that here.
First of all, the code for `gsDesign::xprint()` in versions 3.1.0 and earlier was simply:

```{r,eval=FALSE}
xprint <- function(x, include.rownames = FALSE, hline.after = c(-1, which(x$Value == x[1, ]$Value) - 1, nrow(x)), ...) {
  xtable::print.xtable(x, hline.after = hline.after, include.rownames = include.rownames, ...)
}
```




How to update code that no longer functions exactly as previously is shown below.

```{r, results='asis',warning=FALSE,message=FALSE}
xprint(xtable(x),type="html")
```


```{r,results='asis'}
print(xtable(x), type = "html", include.rownames = FALSE, hline.after = c(-1, which(x$Value == x[1, ]$Value) - 1, nrow(x)))
```

```{r,results='asis'}
print(x, type = "html", include.rownames = FALSE, hline.after = c(-1, which(x$Value == x[1, ]$Value) - 1, nrow(x)))
```