---
title: "Integer sample size and event counts"
author: "Keaven Anderson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: gsDesign.bib

vignette: >
  %\VignetteIndexEntry{Integer sample size and event counts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction 

The **gsDesign** package was originally designed to have continuous sample size planned rather than integer-based sample size.
Designs with time-to-event outcomes also had non-integer event counts at times of analysis.
This vignette  documents the capability to convert to integer sample sizes and event counts.
This has a couple of implications on design characteristics:

- Information fraction on output will not be exactly as input due to rounding.
- Power on output will not be exactly as input.

This document goes through examples to demonstrate the calculations.
The new function as of July, 2023 is the `toInteger()` which operates on group sequential designs to convert to integer-based sample size and event counts at analyses.
We begin with an abbreviated example for a time-to-event endpoint design to demonstrate basic concepts.
We follow with a more extended example for a binary endpoint to explain more details.

# Time-to-event Endpoint Example

The initial design for a time-to-event endpoint in a 2-arm trial does not have integer sample size and event counts.
See comments in the code and output from the `summary()` function below to understand inputs.

```{r}
library(gsDesign)

x <- gsSurv(
  k = 3, # number of analyses
  test.type = 4, # asymmetric 2-sided design with non-binding futility bound
  alpha = 0.025, # 1-sided Type I error
  beta = 0.1, # Type II error (1 - power; 90% power)
  timing = c(.25, .7), # Fraction of final planned events at interim analyses
  sfu = sfLDOF, # O'Brien-Fleming-like spending for efficacy
  sfl = sfHSD,  # Hwang-Shih-DeCani spending for futility
  sflpar = -2.2,# Futility spending parameter to customize bound 
  lambdaC = log(2)/12, # 12 month median control survival
  hr = 0.75, # Alternate hypothesis hazard ratio
  eta = -log(.98) / 12, # 2% dropout rate per year
  # enrollment accelerates over 6 months to steady state 
  gamma = c(2.5, 5, 7.5, 10), # Relative enrollment rates
  # Duration of relative enrollment rate
  R = c(2, 2, 2, 100), 
  # Enrollment duration targeted to T - minfup = 12 months total
  T = 36, # Trial duration
  minfup = 24, # Minimum follow-up duration
  ratio = 1 # Randomization ratio is 1:1
)
```

We can summarize this textually as:

```{r, results='asis'}
cat(summary(x))
```

We now adapt this design to integer targeted events at each analysis as well as an sample size per arm at the end of the trial.
We provide a table summarizing bounds.
Due to rounding up of the final event count, we see slightly larger than the targeted 90% trial power in the last row of the efficacy column.

```{r}
# Adjust design to integer-based event counts at analyses
# and even integer-based final event count
xi <- toInteger(x)
gsBoundSummary(xi) # Summarize design bounds
```

We now summarize sample size and targeted events at analyses.

```{r}
# Integer event counts at analyses are integer
xi$n.I
# Control planned sample size at analyses
# Final analysis is integer; interim analyses before enrollment completion
# are continuous
xi$eNC
# Experimental analysis planned sample size at analyses
xi$eNE
```


# Binomial endpoint designs

## Fixed sample size

We present a simple example based on comparing binomial rates with interim analyses after 50% and 75% of events. We assume a 2:1 experimental:control randomization ratio. Note that the sample size is not an integer.

```{r}
n.fix <- nBinomial(p1 = .2, p2 = .1, alpha = .025, beta = .2, ratio = 2)
n.fix
```
If we replace the `beta` argument above with a integer sample size that is a multiple of 3 so that we get the desired 2:1 integer sample sizes per arm (432 = 144 control + 288 experimental targeted) we get slightly larger thant the targeted 80% power:

```{r}
nBinomial(p1 = .2, p2 = .1, alpha = .025, n = 432, ratio = 2)
```

## 1-sided design

Now we convert the fixed sample size `n.fix` from above to a 1-sided group sequential design  with interims after 50% and 75% of observations. Again, sample size at each analysis is not an integer. We use the Lan-DeMets spending function approximating an O'Brien-Fleming efficacy bound.

```{r}
x <- gsDesign(alpha = .025, beta = .2, n.fix = n.fix, test.type = 1, sfu = sfLDOF, timing = c(.5, .75))
x$n.I
```
Next we convert to integer sample sizes at each analysis. Interim sample sizes are rounded to the nearest integer.
The default `roundUpFinal = TRUE` rounds the final sample size to the nearest integer to 1 + the experimental:control randomization ratio. Thus, the final sample size of 441 below is a multiple of 3.

```{r}
x_integer <- toInteger(x, ratio = 2)
x_integer$n.I
```

Next we examine the efficacy bound of the 2 designs.

```{r}
x$upper$bound
x_integer$upper$bound
```

The differences are associated with slightly different timing of the analyses associated with the different sample sizes noted above:

```{r}
x$timing
x_integer$timing
```

These differences also make a difference in the cumulative Type I error associated with each analysis as shown below.


```{r}
# Continuous sample size design
cumsum(x$upper$prob[,1])
# Specified spending based on the spending function
x$upper$sf(alpha = x$alpha, t = x$timing, x$upper$param)$spend
```

```{r}
# Integer sample size design
cumsum(x_integer$upper$prob[,1])
# Specified spending based on the spending function
x$upper$sf(alpha = x_integer$alpha, t = x_integer$timing, x_integer$upper$param)$spend
```
Finally, we look at cumulative boundary crossing probabilities under the alternate hypothesis for each design.
Due to rounding up the final sample size, the integer-based design has slightly higher total power than the specified 80% (Type II error `beta = 0.2.`).
Interim power is slightly lower for the integer-based design since sample size is rounded to the nearest integer rather than rounded up as at the final analysis.

```{r}
cumsum(x$upper$prob[,2])
cumsum(x_integer$upper$prob[,2])
```

## Non-binding design

The default `test.type = 4` has a non-binding futility bound. We examine behavior of this design next.
The futility bound is moderately aggressive and, thus, there is a compensatory increase in sample size to retain power.
The parameter `delta1` is the natural parameter denoting the difference in response (or failure) rates of 0.2 vs. 0.1 that was specified in the call to `nBinomial()` above.


```{r}
xnb <- gsDesign(alpha = .025, beta = .2, n.fix = n.fix, test.type = 4, 
                sfu = sfLDOF, sfl = sfHSD, sflpar = -2, 
                timing = c(.5, .75), delta1 = .1)
# Continuous sample size for non-binding design
xnb$n.I
# Continuous sample size for one-sided design is smaller
x$n.I
```

As before, we convert to integer sample sizes at each analysis and see the slight deviations from the interim timing of 0.5 and 0.75.

```{r}
xnbi <- toInteger(xnb, ratio = 2)
xnbi$n.I
xnbi$timing
```

These differences also make a difference in the Type I error associated with each analysis

```{r}
cumsum(xnb$upper$prob[,1])
cumsum(xnbi$upper$prob[,1])
```

The Type I error ignoring the futility bounds just shown does not use the full targeted 0.025 as the calculations assume the trial stops for futility if an interim futility bound is crossed.
The non-binding Type I error assuming the trial does not stop for futility is:

```{r}
cumsum(xnbi$falseposnb)
```



Finally, we look at cumulative lower boundary crossing probabilities under the alternate hypothesis for the integer-based design and compare to the planned $\beta$-spending.
We note that the final Type II error spending is slightly lower than the targeted 0.2 due to rounding up the final sample size.  

```{r}
# Actual cumulative beta spent at each analysis
cumsum(xnbi$lower$prob[,2])
# Spending function target is the same at interims, but larger at final
xnbi$lower$sf(alpha=xnbi$beta, t = xnbi$n.I / max(xnbi$n.I), param = xnbi$lower$param)$spend
```

The $\beta$-spending lower than 0.2 in the first row above is due to the final sample size powering the trial to greater than 0.8 as seen below.

```{r}
sum(xnbi$upper$prob[,2])
```





# References
