---
title: "Integer sample size and event counts"
author: "Keaven Anderson"
output: rmarkdown::html_vignette
bibliography: gsDesign.bib
vignette: >
  %\VignetteIndexEntry{Integer sample size and event counts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

The **gsDesign** package was originally designed to have continuous sample size planned rather than integer-based sample size.
Designs with time-to-event outcomes also had non-integer event counts at times of analysis.
This vignette  documents the capability to convert to integer sample sizes and event counts.
This has a couple of implications on design characteristics:

- Information fraction on output will not be exactly as input due to rounding.
- Power on output will not be exactly as input.

This document goes through examples to demonstrate the calculations.
The new function as of July, 2023 is the `toInteger()` which operates on group sequential designs to convert to integer-based total sample size and event counts at analyses.
As of November, 2024 the rounding defaults are changing as documented below.
We begin with a summary of the method.
Then we provide an abbreviated example for a time-to-event endpoint design to demonstrate basic concepts.
We follow with a more extended example for a binary endpoint to explain more details.

## Summary of method

```{r}
library(gsDesign)
x <- gsSurv(ratio = 1, hr = .74)
y <- x |> toInteger()
# Continuous event counts
x$n.I
# Rounded event counts at interims, rounded up at final analysis
y$n.I
# Continuous sample size at interim and final analyses
as.numeric(x$eNE + x$eNC)
# Rounded up to even final sample size given that x$ratio = 1
# and rounding to multiple of x$ratio + 1
as.numeric(y$eNE + y$eNC)
# With roundUpFinal = FALSE, final sample size rounded to nearest integer
z <- x |> toInteger(roundUpFinal = FALSE)
as.numeric(z$eNE + z$eNC)
```


- When `ratio` is a positive integer, the final sample size is rounded to a multiple of `ratio + 1`.
  - For 1:1 randomization (experimental:control), set `ratio = 1` to round to an even sample size.
  - For 2:1 randomization, set `ratio = 2` to round to a multiple of 3.
  - For 3:2 randomization, set `ratio = 4` to round to a multiple of 5.
  - Note that for the final analysis the sample size is rounded up to the nearest multiple of `ratio + 1` when `roundUpFinal = TRUE` is specified. If `roundUpFinal = FALSE`, the final sample size is rounded to the nearest multiple of `ratio + 1`.
  

## Binomial endpoint designs

### Fixed sample size

We present a simple example based on comparing binomial rates with interim analyses after 50% and 75% of events. We assume a 2:1 experimental:control randomization ratio. Note that the sample size is not an integer. We target 80% power (`beta = .2`).

```{r}
n.fix <- nBinomial(p1 = .2, p2 = .1, alpha = .025, beta = .2, ratio = 2)
n.fix
```
If we replace the `beta` argument above with a integer sample size that is a multiple of 3 so that we get the desired 2:1 integer sample sizes per arm (432 = 144 control + 288 experimental targeted) we get slightly larger than the targeted 80% power:

```{r}
nBinomial(p1 = .2, p2 = .1, alpha = .025, n = 432, ratio = 2)
```

### 1-sided design

Now we convert the fixed sample size `n.fix` from above to a 1-sided group sequential design  with interims after 50% and 75% of observations. Again, sample size at each analysis is not an integer. We use the Lan-DeMets spending function approximating an O'Brien-Fleming efficacy bound.

```{r}
# 1-sided design (efficacy bound only; test.type = 1)
xb <- gsDesign(alpha = .025, beta = .2, n.fix = n.fix, test.type = 1, sfu = sfLDOF, timing = c(.5, .75))
# Continuous sample size (non-integer) at planned analyses
xb$n.I
```
Next we convert to integer sample sizes at each analysis. Interim sample sizes are rounded to the nearest integer.
The default `roundUpFinal = TRUE` rounds the final sample size to the nearest integer to 1 + the experimental:control randomization ratio. Thus, the final sample size of 441 below is a multiple of 3.

```{r}
# Convert to integer sample size with even multiple of ratio + 1
# i.e., multiple of 3 in this case at final analysis
x_integer <- toInteger(xb, ratio = 2)
x_integer$n.I
```

Next we examine the efficacy bound of the 2 designs as they are slightly different.

```{r}
# Bound for continuous sample size design
xb$upper$bound
# Bound for integer sample size design
x_integer$upper$bound
```

The differences are associated with slightly different timing of the analyses associated with the different sample sizes noted above:

```{r}
# Continuous design sample size fractions at analyses
xb$timing
# Integer design sample size fractions at analyses
x_integer$timing
```

These differences also make a difference in the cumulative Type I error associated with each analysis as shown below.


```{r}
# Continuous sample size design
cumsum(xb$upper$prob[, 1])
# Specified spending based on the spending function
xb$upper$sf(alpha = xb$alpha, t = xb$timing, xb$upper$param)$spend
```

```{r}
# Integer sample size design
cumsum(x_integer$upper$prob[, 1])
# Specified spending based on the spending function
# Slightly different from continuous design due to slightly different information fraction
x$upper$sf(alpha = x_integer$alpha, t = x_integer$timing, x_integer$upper$param)$spend
```
Finally, we look at cumulative boundary crossing probabilities under the alternate hypothesis for each design.
Due to rounding up the final sample size, the integer-based design has slightly higher total power than the specified 80% (Type II error `beta = 0.2.`).
Interim power is slightly lower for the integer-based design since sample size is rounded to the nearest integer rather than rounded up as at the final analysis.

```{r}
# Cumulative upper boundary crossing probability under alternate by analysis
# under alternate hypothesis for continuous sample size
cumsum(xb$upper$prob[, 2])
# Same for integer sample sizes at each analysis
cumsum(x_integer$upper$prob[, 2])
```

### Non-binding design

The default `test.type = 4` has a non-binding futility bound. We examine behavior of this design next.
The futility bound is moderately aggressive and, thus, there is a compensatory increase in sample size to retain power.
The parameter `delta1` is the natural parameter denoting the difference in response (or failure) rates of 0.2 vs. 0.1 that was specified in the call to `nBinomial()` above.

```{r}
# 2-sided asymmetric design with non-binding futility bound (test.type = 4)
xnb <- gsDesign(
  alpha = .025, beta = .2, n.fix = n.fix, test.type = 4,
  sfu = sfLDOF, sfl = sfHSD, sflpar = -2,
  timing = c(.5, .75), delta1 = .1
)
# Continuous sample size for non-binding design
xnb$n.I
```

As before, we convert to integer sample sizes at each analysis and see the slight deviations from the interim timing of 0.5 and 0.75.

```{r}
xnbi <- toInteger(xnb, ratio = 2)
# Integer design sample size at each analysis
xnbi$n.I
# Information fraction based on integer sample sizes
xnbi$timing
```

These differences also make a difference in the Type I error associated with each analysis

```{r}
# Type I error, continuous design
cumsum(xnb$upper$prob[, 1])
# Type I error, integer design
cumsum(xnbi$upper$prob[, 1])
```

The Type I error ignoring the futility bounds just shown does not use the full targeted 0.025 as the calculations assume the trial stops for futility if an interim futility bound is crossed.
The non-binding Type I error assuming the trial does not stop for futility is:

```{r}
# Type I error for integer design ignoring futility bound
cumsum(xnbi$falseposnb)
```

Finally, we look at cumulative lower boundary crossing probabilities under the alternate hypothesis for the integer-based design and compare to the planned $\beta$-spending.
We note that the final Type II error spending is slightly lower than the targeted 0.2 due to rounding up the final sample size.

```{r}
# Actual cumulative beta spent at each analysis
cumsum(xnbi$lower$prob[, 2])
# Spending function target is the same at interims, but larger at final
xnbi$lower$sf(alpha = xnbi$beta, t = xnbi$n.I / max(xnbi$n.I), param = xnbi$lower$param)$spend
```

The $\beta$-spending lower than 0.2 in the first row above is due to the final sample size powering the trial to greater than 0.8 as seen below.

```{r}
# beta-spending
sum(xnbi$upper$prob[, 2])
```

## References
