\subsection{A non-inferiority study for a new drug\label{sec:motivationNI}}
The \texttt{nBinomial()} function presented above was specifically designed to work for noninferiority trial design as well as superiority designs.
We consider a new treatment that is to be compared to a standard that has a successful treatment rate of 67.7\%.
An absolute margin of 7\% is considered an acceptable noninferiority margin.
The trial is to be powered at 90\% with 2.5\% Type I error (one-sided) using methods presented by Farrington and Manning \cite{FarringtonManning}.
The function call \texttt{nBinomial(p1=.677, p2=.677, delta0=.07)} shows that a fixed sample size of 1874 is adequate for this purpose.
There are some concerns about these assumptions, however. 
First, the control group event rate may be incorrect.
As the following code using event rates from .55 to .75 demonstrates, the required sample size may range from 1600 to over 2100.
\begin{verbatim}
> p <- c(.55, .6, .65, .7, .75)
> ceiling(nBinomial(p1=p,p2=p,delta0=.07))
[1] 2117 2054 1948 1800 1611
\end{verbatim}
More importantly, if the experimental group therapy does not work quite as well as control, there is a considerable dropoff in power to demonstrate non-inferiority.
Thus, there may be value in planning an interim futility analysis to stop the trial if the success rate with experimental therapy is trending substantially worse than with control.

\subsection{A diabetes outcomes trial example\label{sec:motivationDiab}}
Current regulatory standards for chronic therapies of diabetes require ensuring that a new drug in a treatment class does not have substantially inferior cardiovascular outcomes compared to an approved treatment or treatments \cite{DiabetesCV}.
While we do not claim the designs for this example presented here would be acceptable to regulators, the specifics of the guidance provide a nice background for the use of the \texttt{gsDesign} package to derive group sequential designs that fit a given problem.
The initial reason for presenting this example is that there is likely to be a genuine public health interest in showing any of the following for the two treatment arms compared:
\begin{itemize}
\item The two treatment arms are similar (equivalence).
\item One arm is similar to or better than the other (non-inferiority).
\item Either arm is superior to the other (2-sided testing of no difference).
\end{itemize}
The example is somewhat simplified here.
We assume patients with diabetes have a risk of a cardiovascular event of about 1.5\% per year and a 15\% dropout rate per year.
If each arm has the same cardiovascular risk as the other, we would like to have 90\% power to rule out a hazard ratio of 1.3 in either direction.
Type I error if one arm has an elevated hazard ratio of 1.3 compared to the other should be 2.5\%.
The trial is to enroll in 2 years and have a minimum follow-up of 4 years, leading to a total study time of 6 years.
The sample size routine \texttt{nSurvival()} is currently set up to have no treatment difference as the null hypothesis, whereas here we wish to use this as the alternative hypothesis.
Thus, in using \texttt{nSurvival()} we switch the role of Type I error \texttt{alpha} and Type II error \texttt{beta}
\begin{verbatim}
x <- nSurvival(lambda.0=-log(.985), lambda.1=-log(.985) * 1.3, eta=-log(.85),
        alpha=.2, beta=.025, Tr=2, Ts=6, type="rr", entry="unif")
x$Sample.size
x$Num.events
\end{verbatim}
Since the default in \texttt{nSurvival()} is 2-sided testing we have set \texttt{alpha=.2} to ensure there is a 10\% probability of rejecting no difference when there is a hazard ratio of 1.3 or 1/1.3 for control versus experimental treatment.
The above code suggests 10,800 patients should be enrolled and final analysis conducted when 617 cardiovascular events have been observed.
Generally, a confidence interval for the hazard ratio of experimental to control is used to express treatment differences at the end of this type of trial.
A confidence interval will rule out the specified treatment differences consistently with testing if, for example, the same proportional hazards regression model is used for both the a Wald test and the corresponding confidence interval.
The terminology of "control" and "experimental" is generally inappropriate when both therapies are approved.
However, for this example it is generally the case that a new therapy is being compared to an established one and there may be some asymmetry when considering the direction of inference.
Various questions arise concerning early stopping in a trial of this nature:
\begin{itemize}
\item While it would be desirable to stop early if the new therapy has a significantly lower cardiovascular event rate, a minimum amount of follow-up may be valuable to ensure longer-term safety and general acceptance of the results.
\item If a trend emerges in favor of the experimental treatment, it will likely be possible to demonstrate non-inferiority prior to being able to demonstrate superiority. If the trial remains blinded until superiority is demonstrated or until the final planned analysis, full acceptance of a useful new therapy may be delayed. 
As noted above, the value of long-term safety data may be more important than an early stop based on "short-term" endpoint.
\item From a sponsor's standpoint, it may be desirable to stop the trial if it becomes futile to demonstrate the experimental therapy is non-inferior to control; that is, there is an interim trend favoring control. However, if both treatment groups represent marketed products then from a public health standpoint it may be desirable to continue the trial to demonstrate a statistically significant advantage for the control treatment.
\end{itemize}
